### 2015-10-28 paper.tex
First

\documentclass{sig-alternate}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\newcommand{\superscript}[1]{\ensuremath{^{\textrm{#1}}}}
\def\sharedaffiliation{\end{tabular}\newline\begin{tabular}{c}}
\def\wu{\superscript{*}}
\def\wg{\superscript{\dag}}

\usepackage{listings}

% Typography
\usepackage{times}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage[normalem]{ulem}
\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=2,
            urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}
\def\UrlFont{\em}

% Graphics
\usepackage{tikz}
\usepackage{tkz-graph}
\usetikzlibrary{arrows,positioning,shapes.misc}
\usepackage{graphicx}
\definecolor{lightgrey}{RGB}{170, 170, 170}
\definecolor{darkblue}{RGB}{0, 0, 0}
\definecolor{darkred}{RGB}{170, 0, 0}
\definecolor{darkgreen}{RGB}{0, 110, 0}

% Acronyms
\usepackage{xspace}
\newcommand{\sparql}{{SPARQL}\xspace}
\newcommand{\sparqlo}{{SPARQL 1.1}\xspace}
\newcommand{\arq}{{ARQ}\xspace}
\newcommand{\wthreec}{{W\oldstylenums 3C}\xspace}
\newcommand{\sfive}{{S\oldstylenums 5}\xspace}
\newcommand{\select}{{SELECT}\xspace}
\newcommand{\construct}{{CONSTRUCT}\xspace}
\newcommand{\ask}{{ASK}\xspace}
\newcommand{\describe}{{DESCRIBE}\xspace}
\newcommand{\from}{{FROM}\xspace}
\newcommand{\odbc}{{odbc}\xspace}

% Tight lists
\usepackage{enumitem}
\setlist{nolistsep}

% Listings and Verbatim environment
\usepackage{fancyvrb}
\usepackage{relsize}
\usepackage{listings}
\usepackage{verbatim}
\newcommand{\smalllistingsize}{\fontsize{8pt}{9.5pt}}
\newcommand{\inlinelistingsize}{\fontsize{8.5pt}{11pt}}
\newcommand{\defaultlistingsize}{\inlinelistingsize}
\RecustomVerbatimCommand{\Verb}{Verb}{fontsize=\inlinelistingsize}
\RecustomVerbatimEnvironment{Verbatim}{Verbatim}{fontsize=\inlinelistingsize}
\lstset{frame=lines,captionpos=b,numberbychapter=false,escapechar=§,
        belowskip=1em,
        xleftmargin=2ex,
        framexleftmargin=2ex,
        basicstyle=\ttfamily\smalllistingsize\selectfont}
\crefname{lstlisting}{Listing}{Listings}
\definecolor{grey}{RGB}{130,130,130}

\usepackage{color}
\newcommand{\todo}[1]{\noindent\textcolor{red}{{\bf \{TODO}: #1{\bf \}}}}
\begin{document}

\title{What we can learn from one year of\\ public transit route planning API queries}
\numberofauthors{6}
\author{
\alignauthor
Pieter Colpaert\\
 \affaddr{\email{\texttt{pieter.colpaert@ugent.be}}}
\and
\alignauthor
Alvin Chua\\
 \affaddr{\email{\texttt{alvin.chua@asro.kuleuven.be}}}
\and
\alignauthor
Ruben Verborgh\\
 \affaddr{\email{\texttt{ruben.verborg@ugent.be}}}
\and
\alignauthor
Erik Mannens\\
 \affaddr{\email{\texttt{erik.mannens@ugent.be}}}
\and
\alignauthor
Rik Van de Walle\\
 \affaddr{\email{\texttt{rik.vandewalle@ugent.be}}}
\and
\alignauthor
Andrew Vande Moere\\
 \affaddr{\email{\texttt{andrew.vandemoere@asro.kuleuven.be}}}\\
}

\maketitle
\begin{abstract}
% Context
%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
The iRail project hosts a route planning API for the Belgian railway company.
This webservice provides a concise answer to a complex question with parameters such as ``departuretime'', ``from'' and ``to''.
% Need
In the field of urban planning, researchers need an indication of how people move between cities. 
Yet, getting this data from official sources has proven to be troublesome.
% Task
In this paper, we analyse the query logs from the iRail API, to tell us something about how people move between cities in Belgium.
% Object
We studied queries for the year 2013, which contains an average of 3000 queries per day.
% Findings
We found X, Y, Z interesting patterns (\todo{Alvin}), which correspond with reality.
% Conclusion
This proves that query logs from route planning systems are valuable as an indication of transit flows and may house interesting stories.
% Perspectives
However, many private owned route planners exists, which keep these query logs strictly confidential.
Bearing this in mind, we discuss the current state of the art in public transit schedule publishing, and formulate opportunities for gathering query logs when using a Linked Data Fragments approach.

\end{abstract}

\vspace{1em}

\section{Introduction}
\label{sec:introduction}

%What's iRail

\section{Gathering the logs}
\label{sec:logs}

The logs are available at \ldots

\section{Looking for patterns}
\label{sec:method}

\section{Results}
\label{sec:results}

\todo{add a couple of nice images with explanation here}

\subsection{Pattern 1: ...}
\label{sec:pattern1}

\subsection{Pattern 2: ...}
\label{sec:pattern2}

\subsection{Pattern 3: ...}
\label{sec:pattern3}

\section{Publishing transport data}

Thoughts on publishing transport data on the web and query logs in the case of GTFS, Linked Connections\cite{lc} (linked data fragments\cite{ldf}) and webservices.

\section{Conclusion}
\label{sec:conclusion}

Query logs are very interesting

% Fix spacing after References header (as line 1308 of sig-alternate.cls breaks it)
\let\oldsection\section
\renewcommand{\section}[2][1]{\oldsection{#1}\vspace{-3pt}}

\bibliographystyle{abbrv}
\bibliography{refs}
\end{document}

### 2015-12-16 paper.tex
Added introduction

@@
 
-%What's iRail
+Until today, public transport data remains absent from the Linked Open Data cloud\footnote{\url{http://lod-cloud.net/}}.
+Bizarre, one could notice, as many open transport datasets are already available, according to the Global Open Data Index of 2014\footnote{\url{http://index.okfn.org/dataset/timetables/2014/}}.
+There are currently two ways for public transit agencies to publish their data to the Web, as illustrated in \cref{fig:LDFAxis1}: 1) offering a route planning service, and/or 2) publishing data using the General Transit Feed Specification (GTFS).
+
+\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV-files, describing the rules for when a public transit vehicle will pass by on a certain location.
+The specification is a huge success: up to date, it is supported among all current open source route planning software systems, and it is used in products such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
+Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
+The goal of GTFS is to create a good exchange format which specific software packages can use to transform it into their format of choice.
+
+On the other hand, public transit agencies also publish their data by providing \emph{route planners}.
+These route planner offer a query service on top of the data and expose these over the Web.
+However, when machines should access these, rate limiters are common place: only a limited amount of queries should be done in order for the service to be able to stay online.
+
+There is a trade-off to be made between these two options, as illustrated in \cref{fig:LDFAxis1}.
+This trade-off axis was first introduced as a way to talk about server/client load with Linked Data Fragments~\cite{ldf}.
+In this paper we use this axis to talk about how well we can gather query logs when publishing transport data.
+
+% Linked Data Fragments axis
+\begin{figure}[t]
+  \begin{tikzpicture}[
+    x=.005\textwidth,
+    y=.1cm,
+    axis/.style={
+      line width=.8pt,
+      stealth-stealth,
+      color=black!65,
+    },
+    axislabel/.style={
+      inner sep=0,
+      font=\slshape\fontsize{8}{9}\selectfont,
+      color=black!70,
+    },
+    tick/.style={
+      line width=1pt,
+    },
+    ticklabel/.style={
+      anchor=south,
+      inner sep=0,
+      align=center,
+      font=\bfseries\fontsize{8}{8}\selectfont,
+      text depth=0pt,
+    },
+  ]
+    \newcommand\tick[2]{%
+      \draw[tick] (#1, 1.7) -- (#1, -.4pt);
+      \node[ticklabel] at (#1, 2.4) {#2};
+    }
+    \newcommand\vaguetick[1]{%
+      \draw[tick,densely dotted] (#1, 1.7) -- (#1, -.4pt);
+    }
+
+    \draw[axis] (-100, 0) -- (0, 0);
+    \node[axislabel,align=left,anchor=north west] at (-100, -2.3)
+    {
+%      generic requests\\
+      high client effort\\
+      high server availability\\
+      know how many times datadump was downloaded
+    };
+    \node[axislabel,align=right,anchor=north east] at (0, -2.3)
+    {
+%      specific requests\\
+      high server effort\\
+      low server availability\\
+      detailled query logs
+    };
+
+    \tick{-93}{GTFS data\\dump}
+    \tick{-10}{route planning\\service}
+%    \tick{-32}{Linked\\Connections}
+
+   % \vaguetick{-84}
+   % \vaguetick{-52}
+   % \vaguetick{  5}
+   % \vaguetick{-10}
+   % \vaguetick{ 28}
+   % \vaguetick{ 55}
+
+   %\node[align=center,font=\fontsize{8}{9}\selectfont\bfseries] at (0, -8.3)
+   %   {various types of\\Linked Data Fragments};
+  \end{tikzpicture}
+  \caption{
+    This axis illustrates two extremes.
+With GTFS data dumps, the data reusers do all the work, and no query logs can be used.
+With route planning services, the servers need to do all the work, yet detailled query logs can be studied.
+  }
+  \label{fig:LDFAxis1}
+\end{figure}
+
+% END AXIS
+
+The \emph{iRail} project\footnote{\url{http://hello.irail.be}} is a project started in 2008 to make the data of the Belgian railway accessible for developers.
+Ever since, the project offers developer both a GTFS data dump for third party apps and a route planning API.
+The API, during 2013/2014, got an average of 3000 route planning requests a day.
+In this paper, we study these logs and see whether we can find interesting pattern in it.
+Finally, we look at Linked Connections~\cite{lc} and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments approach.
 

### 2015-12-16 paper.tex
Relwork added

@@
 Andrew Vande Moere\\
-\affaddr{\email{\texttt{andrew.vandemoere@asro.kuleuven.be}}}\\
+\affaddr{\email{\texttt{andrew.vandemoere\\@asro.kuleuven.be}}}\\
 }
@@
 The API, during 2013/2014, got an average of 3000 route planning requests a day.
-In this paper, we study these logs and see whether we can find interesting pattern in it.
+
+Getting indications of people flows through public transit networks is a challenge.
+The data is tedious to get, mainly with public transit systems where passengers don't have to check-in and check-out.
+In this paper, we study these logs and see whether we can find interesting patterns.
 Finally, we look at Linked Connections~\cite{lc} and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments approach.
+
+\section{Related work}
+\label{sec:relwork}
+
+\emph{Flow analysis} is a topic of theoretical interest and practical importance in various disciplines. 
+Flow analysis is conventionally conducted to study spatial dynamics and identify routine patterns in the movement of people.
+For instance, interest in modelling traffic flows emerged from the strain placed on urban transportation systems during peak hours~\cite{roth,ferreira}.
+Likewise, insight into routine travel patterns is crucial for the conceptualisation of functional urban areas~\cite{servillo,sykora}, urban hierarchies~\cite{christaller} and other territorial structures.
+
+Over the past decade, large datasets have become increasingly commonplace due to the proliferation of sensor networks and portable devices like smartphones.
+Termed ``Big Data'' due to the large volume of data records that emerge from real-time sensing\cite{kitchin}, such datasets typically contain information of activities or processes linked to the space and time where they occur.
+In the domain of ``Smart City'' research, much has been accomplished with the use of ``Big Data'' to monitor human movement.
+Smart card data from public transport systems~\cite{roth,beecham}, taxi journeys~\cite{ferreira} as well as cellular call data~\cite{sevtsuk} have provided planners with new opportunities to develop greater understanding of mobility patterns in urban environments~\cite{batty}.
 
@@
 \label{sec:logs}
+
+In 2013, we have used a MySQL table to store each request in a table, including User-Agent string, the timestamp, whether or not the request succeeded, and so forth.
+
+We have cleaned this data by removing bots, such as the Google bot, but also bots that were using the API for analytical purposes instead.
 
@@
 
+There are obvious caveats associated with the use of such data as proxy for actual statistical counts
+
 \section{Results}
 \label{sec:results}
+\todo{
+We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
+Look at weekday-mornings: commuting towards Gent, Brussels, Antwerp.
+This is not as obvious in the Walloon region (except for Liege).
+The reverse occurs on weekday-evenings.
+Local patterns can also be observed and this corresponds to what we know: "cities are the central hubs".
+Walloon region: everything converges on Brussels
+I found that whereas there's almost no movement from Antwerp city in the week to West Flanders, there is movement from Antwerp in the weekend to the coast and back. 
+}
+
+\todo{This pair of chord diagrams capture the aggregated number of queries made between any two stations on weekdays and weekends between 6 to 9 in the morning.
+The chords that link two distinct stations, are coloured coded by region and proportionately scaled to the number of queries from a station of origin to a destination.
+For instance, approximately 3,530 queries were made from Leuven to Brussels than in the opposite direction (approximately 1,530) on weekdays.
+Reading both diagrams in this manner reveals the complexity of movement on the rail system and the significance of cities in daily travel.
+Brussels serves as the principal centre of rail activity in general, yet its centrality is more distinctive for the walloon region than Flanders.
+Queries from Walloonia with exception to Liege, are generally made from provincial stations towards Brussels instead of their respective provincial capitals.
+Queries in Flanders, on the hand, tend to be distributed among major cities otherwise known as the Flemish Diamond (Vlaamse Ruit), a network of four metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
+The difference between both patterns appear to correspond with existing measures of population density, providing valuable insight and alternative perspectives into the function of cities in rural and urban settings.}
 

### 2015-12-18 ldfaxis2.tex
Refactored intro a bit

--- 2015-10-28 paper.tex
@@
-\documentclass{sig-alternate}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-
-\newcommand{\superscript}[1]{\ensuremath{^{\textrm{#1}}}}
-\def\sharedaffiliation{\end{tabular}\newline\begin{tabular}{c}}
-\def\wu{\superscript{*}}
-\def\wg{\superscript{\dag}}
-
-\usepackage{listings}
-
-% Typography
-\usepackage{times}
-\usepackage{mathptmx}
-\usepackage{microtype}
-\usepackage[normalem]{ulem}
-\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=2,
-            urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
-\usepackage[capitalize,noabbrev]{cleveref}
-\def\UrlFont{\em}
-
-% Graphics
-\usepackage{tikz}
-\usepackage{tkz-graph}
-\usetikzlibrary{arrows,positioning,shapes.misc}
-\usepackage{graphicx}
-\definecolor{lightgrey}{RGB}{170, 170, 170}
-\definecolor{darkblue}{RGB}{0, 0, 0}
-\definecolor{darkred}{RGB}{170, 0, 0}
-\definecolor{darkgreen}{RGB}{0, 110, 0}
-
-% Acronyms
-\usepackage{xspace}
-\newcommand{\sparql}{{SPARQL}\xspace}
-\newcommand{\sparqlo}{{SPARQL 1.1}\xspace}
-\newcommand{\arq}{{ARQ}\xspace}
-\newcommand{\wthreec}{{W\oldstylenums 3C}\xspace}
-\newcommand{\sfive}{{S\oldstylenums 5}\xspace}
-\newcommand{\select}{{SELECT}\xspace}
-\newcommand{\construct}{{CONSTRUCT}\xspace}
-\newcommand{\ask}{{ASK}\xspace}
-\newcommand{\describe}{{DESCRIBE}\xspace}
-\newcommand{\from}{{FROM}\xspace}
-\newcommand{\odbc}{{odbc}\xspace}
-
-% Tight lists
-\usepackage{enumitem}
-\setlist{nolistsep}
-
-% Listings and Verbatim environment
-\usepackage{fancyvrb}
-\usepackage{relsize}
-\usepackage{listings}
-\usepackage{verbatim}
-\newcommand{\smalllistingsize}{\fontsize{8pt}{9.5pt}}
-\newcommand{\inlinelistingsize}{\fontsize{8.5pt}{11pt}}
-\newcommand{\defaultlistingsize}{\inlinelistingsize}
-\RecustomVerbatimCommand{\Verb}{Verb}{fontsize=\inlinelistingsize}
-\RecustomVerbatimEnvironment{Verbatim}{Verbatim}{fontsize=\inlinelistingsize}
-\lstset{frame=lines,captionpos=b,numberbychapter=false,escapechar=§,
-        belowskip=1em,
-        xleftmargin=2ex,
-        framexleftmargin=2ex,
-        basicstyle=\ttfamily\smalllistingsize\selectfont}
-\crefname{lstlisting}{Listing}{Listings}
-\definecolor{grey}{RGB}{130,130,130}
-
-\usepackage{color}
-\newcommand{\todo}[1]{\noindent\textcolor{red}{{\bf \{TODO}: #1{\bf \}}}}
-\begin{document}
-
-\title{What we can learn from one year of\\ public transit route planning API queries}
-\numberofauthors{6}
-\author{
-\alignauthor
-Pieter Colpaert\\
- \affaddr{\email{\texttt{pieter.colpaert@ugent.be}}}
-\and
-\alignauthor
-Alvin Chua\\
- \affaddr{\email{\texttt{alvin.chua@asro.kuleuven.be}}}
-\and
-\alignauthor
-Ruben Verborgh\\
- \affaddr{\email{\texttt{ruben.verborg@ugent.be}}}
-\and
-\alignauthor
-Erik Mannens\\
- \affaddr{\email{\texttt{erik.mannens@ugent.be}}}
-\and
-\alignauthor
-Rik Van de Walle\\
- \affaddr{\email{\texttt{rik.vandewalle@ugent.be}}}
-\and
-\alignauthor
-Andrew Vande Moere\\
- \affaddr{\email{\texttt{andrew.vandemoere@asro.kuleuven.be}}}\\
+% Linked Data Fragments axis
+\begin{figure}[t]
+  \begin{tikzpicture}[
+    x=.005\textwidth,
+    y=.1cm,
+    axis/.style={
+      line width=.8pt,
+      stealth-stealth,
+      color=black!65,
+    },
+    axislabel/.style={
+      inner sep=0,
+      font=\slshape\fontsize{8}{9}\selectfont,
+      color=black!70,
+    },
+    tick/.style={
+      line width=1pt,
+    },
+    ticklabel/.style={
+      anchor=south,
+      inner sep=0,
+      align=center,
+      font=\bfseries\fontsize{8}{8}\selectfont,
+      text depth=0pt,
+    },
+  ]
+    \newcommand\tick[2]{%
+      \draw[tick] (#1, 1.7) -- (#1, -.4pt);
+      \node[ticklabel] at (#1, 2.4) {#2};
+    }
+    \newcommand\vaguetick[1]{%
+      \draw[tick,densely dotted] (#1, 1.7) -- (#1, -.4pt);
     }
 
-\maketitle
-\begin{abstract}
-% Context
-%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
-The iRail project hosts a route planning API for the Belgian railway company.
-This webservice provides a concise answer to a complex question with parameters such as ``departuretime'', ``from'' and ``to''.
-% Need
-In the field of urban planning, researchers need an indication of how people move between cities. 
-Yet, getting this data from official sources has proven to be troublesome.
-% Task
-In this paper, we analyse the query logs from the iRail API, to tell us something about how people move between cities in Belgium.
-% Object
-We studied queries for the year 2013, which contains an average of 3000 queries per day.
-% Findings
-We found X, Y, Z interesting patterns (\todo{Alvin}), which correspond with reality.
-% Conclusion
-This proves that query logs from route planning systems are valuable as an indication of transit flows and may house interesting stories.
-% Perspectives
-However, many private owned route planners exists, which keep these query logs strictly confidential.
-Bearing this in mind, we discuss the current state of the art in public transit schedule publishing, and formulate opportunities for gathering query logs when using a Linked Data Fragments approach.
+    \draw[axis] (-100, 0) -- (0, 0);
+    \node[axislabel,align=left,anchor=north west] at (-100, -2.3)
+    {
+%      generic requests\\
+      high client effort\\
+      high server availability\\
+      know how many times datadump was downloaded
+    };
+    \node[axislabel,align=right,anchor=north east] at (0, -2.3)
+    {
+%      specific requests\\
+      high server effort\\
+      low server availability\\
+      detailled query logs
+    };
 
-\end{abstract}
+    \tick{-93}{GTFS data\\dump}
+    \tick{-10}{route planning\\service}
+    \tick{-50}{Linked\\Connections}
 
-\vspace{1em}
+    \vaguetick{-84}    \vaguetick{-70}
+    \vaguetick{-10}
+    \vaguetick{-28}
+    \vaguetick{-20}
 
-\section{Introduction}
-\label{sec:introduction}
-
-%What's iRail
-
-\section{Gathering the logs}
-\label{sec:logs}
-
-The logs are available at \ldots
-
-\section{Looking for patterns}
-\label{sec:method}
-
-\section{Results}
-\label{sec:results}
-
-\todo{add a couple of nice images with explanation here}
-
-\subsection{Pattern 1: ...}
-\label{sec:pattern1}
-
-\subsection{Pattern 2: ...}
-\label{sec:pattern2}
-
-\subsection{Pattern 3: ...}
-\label{sec:pattern3}
-
-\section{Publishing transport data}
-
-Thoughts on publishing transport data on the web and query logs in the case of GTFS, Linked Connections\cite{lc} (linked data fragments\cite{ldf}) and webservices.
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-Query logs are very interesting
-
-% Fix spacing after References header (as line 1308 of sig-alternate.cls breaks it)
-\let\oldsection\section
-\renewcommand{\section}[2][1]{\oldsection{#1}\vspace{-3pt}}
-
-\bibliographystyle{abbrv}
-\bibliography{refs}
-\end{document}
+   %\node[align=center,font=\fontsize{8}{9}\selectfont\bfseries] at (0, -8.3)
+   %   {various types of\\Linked Data Fragments};
+  \end{tikzpicture}
+  \caption{This axis extends \cref{fig:LDFAxis1} with Linked Connections: a different choice of trade-offs for route planning advice over the Web.
+  }
+  \label{fig:LDFAxis2}
+\end{figure}

### 2015-12-18 paper.tex
Refactored intro a bit

@@
 
-Until today, public transport data remains absent from the Linked Open Data cloud\footnote{\url{http://lod-cloud.net/}}.
+Today, public transport data still remains absent from the Linked Open Data cloud\footnote{\url{http://lod-cloud.net/}}.
 Bizarre, one could notice, as many open transport datasets are already available, according to the Global Open Data Index of 2014\footnote{\url{http://index.okfn.org/dataset/timetables/2014/}}.
-There are currently two ways for public transit agencies to publish their data to the Web, as illustrated in \cref{fig:LDFAxis1}: 1) offering a route planning service, and/or 2) publishing data using the General Transit Feed Specification (GTFS).
+We can imagine that, what contributed to this, are the de-facto standard ways to publish public transit data to the Web.
+There are currently two ways, as illustrated in \cref{fig:LDFAxis1}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
 
@@
 Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
-The goal of GTFS is to create a good exchange format which specific software packages can use to transform it into their format of choice.
+The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
 
@@
 
-There is a trade-off to be made between these two options, as illustrated in \cref{fig:LDFAxis1}.
-This trade-off axis was first introduced as a way to talk about server/client load with Linked Data Fragments~\cite{ldf}.
-In this paper we use this axis to talk about how well we can gather query logs when publishing transport data.
-
-% Linked Data Fragments axis
-\begin{figure}[t]
-  \begin{tikzpicture}[
-    x=.005\textwidth,
-    y=.1cm,
-    axis/.style={
-      line width=.8pt,
-      stealth-stealth,
-      color=black!65,
-    },
-    axislabel/.style={
-      inner sep=0,
-      font=\slshape\fontsize{8}{9}\selectfont,
-      color=black!70,
-    },
-    tick/.style={
-      line width=1pt,
-    },
-    ticklabel/.style={
-      anchor=south,
-      inner sep=0,
-      align=center,
-      font=\bfseries\fontsize{8}{8}\selectfont,
-      text depth=0pt,
-    },
-  ]
-    \newcommand\tick[2]{%
-      \draw[tick] (#1, 1.7) -- (#1, -.4pt);
-      \node[ticklabel] at (#1, 2.4) {#2};
-    }
-    \newcommand\vaguetick[1]{%
-      \draw[tick,densely dotted] (#1, 1.7) -- (#1, -.4pt);
-    }
-
-    \draw[axis] (-100, 0) -- (0, 0);
-    \node[axislabel,align=left,anchor=north west] at (-100, -2.3)
-    {
-%      generic requests\\
-      high client effort\\
-      high server availability\\
-      know how many times datadump was downloaded
-    };
-    \node[axislabel,align=right,anchor=north east] at (0, -2.3)
-    {
-%      specific requests\\
-      high server effort\\
-      low server availability\\
-      detailled query logs
-    };
-
-    \tick{-93}{GTFS data\\dump}
-    \tick{-10}{route planning\\service}
-%    \tick{-32}{Linked\\Connections}
-
-   % \vaguetick{-84}
-   % \vaguetick{-52}
-   % \vaguetick{  5}
-   % \vaguetick{-10}
-   % \vaguetick{ 28}
-   % \vaguetick{ 55}
-
-   %\node[align=center,font=\fontsize{8}{9}\selectfont\bfseries] at (0, -8.3)
-   %   {various types of\\Linked Data Fragments};
-  \end{tikzpicture}
-  \caption{
-    This axis illustrates two extremes.
-With GTFS data dumps, the data reusers do all the work, and no query logs can be used.
-With route planning services, the servers need to do all the work, yet detailled query logs can be studied.
-  }
-  \label{fig:LDFAxis1}
-\end{figure}
-
-% END AXIS
-
-The \emph{iRail} project\footnote{\url{http://hello.irail.be}} is a project started in 2008 to make the data of the Belgian railway accessible for developers.
-Ever since, the project offers developer both a GTFS data dump for third party apps and a route planning API.
-The API, during 2013/2014, got an average of 3000 route planning requests a day.
+\input{ldfaxis1}
 
@@
 The data is tedious to get, mainly with public transit systems where passengers don't have to check-in and check-out.
-In this paper, we study these logs and see whether we can find interesting patterns.
-Finally, we look at Linked Connections~\cite{lc} and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments approach.
+Nevertheless, people are calculating their intended routes by using the Web.
+The \emph{iRail} project\footnote{\url{http://hello.irail.be}} started in 2008 to make the data of the Belgian railway accessible for developers.
+Ever since, the project offers developer both a GTFS data dump for third party apps and a route planning API.
+
+In this paper, we give a small overview of related work trying to gather data for flow analysis.
+Next, we study the iRail query logs to find out whether we can find interesting patterns.
+Finally, we look at Linked Connections~\cite{lc}, a proposed new way of publishing queryable public transit data, and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments~\cite{ldf} approach.
 
@@
 
-In 2013, we have used a MySQL table to store each request in a table, including User-Agent string, the timestamp, whether or not the request succeeded, and so forth.
+In 2013, we have started to use a MySQL table to store each request in a table, including User-Agent string, the timestamp, whether or not the request succeeded, the departure station and the desired station of arrival.
+From an export of this table, we have filtered out only the route plannning queries that succeeded.
+The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/iraillogs.zip}}.
 
-We have cleaned this data by removing bots, such as the Google bot, but also bots that were using the API for analytical purposes instead.
+We have cleaned this data by removing the requests done by harvesters or search engine bots.
+Thanks to the user agents, this was a fairly straight-forward operation.
+The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
 
-The logs are available at \ldots
+Today, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+This amounts today to an average of 200k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\todo{\url{http://...}}} and \emph{Railer}\footnote{\todo{\url{http://...}}}.
 
@@
 
-There are obvious caveats associated with the use of such data as proxy for actual statistical counts
+There are obvious caveats associated with the use of such data as proxy for actual statistical counts.
 
@@
 
-\subsection{Pattern 3: ...}
-\label{sec:pattern3}
+\section{Publishing transport data and the impact on query logs}
+\label{sec:publishing}
 
-\section{Publishing transport data}
+Publishing public transport data today happens in two ways: by publishing a data dump of the timetables or by hosting a public route planning service.
+Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
+These trade-offs can be mapped on an axis, cfr. \cref{fig:LDFAxis2}, in which the 2 extremes are identified as: the server does all the processing versus the client does all the processing.
+
+Also for query logs we can make the same trade-off, in which the 2 extremes are the server knows only when someone downloads the entire datadump versus the server has a detailled query log.
+In this paper, we have studied the query logs of an API s
+
+%label: fig:LDFAxis2
+\input{ldfaxis2}
+
+\todo{Also in USEWOD2014 there was a paper on LDF and how we could know which SPARQL queries needed to be solved}
 
@@
 
-\section{Conclusion}
+\section{Conclusion and discussion}
 \label{sec:conclusion}

### 2015-12-21 ldfaxis2.tex
rewrote axises, rewrote abstract, added chord diagrams, etc.

@@
 
-    \draw[axis] (-100, 0) -- (0, 0);
+    \draw[axis] (-100, 0) -- (-5, 0);
     \node[axislabel,align=left,anchor=north west] at (-100, -2.3)
@@
     };
-    \node[axislabel,align=right,anchor=north east] at (0, -2.3)
+    \node[axislabel,align=right,anchor=north east] at (-5, -2.3)
     {
@@
     };
-
-    \tick{-93}{GTFS data\\dump}
-    \tick{-10}{route planning\\service}
-    \tick{-50}{Linked\\Connections}
-
-    \vaguetick{-84}    \vaguetick{-70}
-    \vaguetick{-10}
-    \vaguetick{-28}
-    \vaguetick{-20}
+    \node[text,ticklabel] at (-53, 2.4) {Linked\\Connections};
+    \tick{-90}{GTFS data\\dump}
+    \tick{-15}{route planning\\service}
+    \vaguetick{-23}
+    \vaguetick{-32}
+    \vaguetick{-41}
+    \vaguetick{-50}
+    \vaguetick{-60}
+    \vaguetick{-70}
+    \vaguetick{-80}
 
@@
   \end{tikzpicture}
-  \caption{This axis extends \cref{fig:LDFAxis1} with Linked Connections: a different choice of trade-offs for route planning advice over the Web.
-  }
+  \caption{This axis extends \cref{fig:LDFAxis1} with Linked Connections: the different ticks on the axis for client effort versus server expressivity.}
   \label{fig:LDFAxis2}

### 2015-12-21 paper.tex
rewrote axises, rewrote abstract, added chord diagrams, etc.

@@
 
-\title{What we can learn from one year of\\ public transit route planning API queries}
+\title{Studying public transit API query logs\\ to get an indication of travel flows}
 \numberofauthors{6}
@@
 The iRail project hosts a route planning API for the Belgian railway company.
-This webservice provides a concise answer to a complex question with parameters such as ``departuretime'', ``from'' and ``to''.
+This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
 % Need
 In the field of urban planning, researchers need an indication of how people move between cities. 
-Yet, getting this data from official sources has proven to be troublesome.
+%Yet, getting this data from official sources has proven to be troublesome.
 % Task
-In this paper, we analyse the query logs from the iRail API, to tell us something about how people move between cities in Belgium.
+In this paper, we analyse the query logs of the iRail API, to see whether it can tell us something about travel flows between cities in Belgium.
 % Object
-We studied queries for the year 2013, which contains an average of 3000 queries per day.
+We studied queries for the year 2013, which contains an average of $\sim3000$ interesting queries per day ($\sim0.4\%$ of the amount of passengers per weekday).
 % Findings
-We found X, Y, Z interesting patterns (\todo{Alvin}), which correspond with reality.
+We illustrate \todo{X} interesting patterns that correspond to reality and we show an outlier in the data.
 % Conclusion
-This proves that query logs from route planning systems are valuable as an indication of transit flows and may house interesting stories.
+Visualizing the data shows the query logs are a better representation than random data, yet more research is needed to understand how representative for travel flows they may be.
 % Perspectives
-However, many private owned route planners exists, which keep these query logs strictly confidential.
-Bearing this in mind, we discuss the current state of the art in public transit schedule publishing, and formulate opportunities for gathering query logs when using a Linked Data Fragments approach.
+In earlier work, we have suggested the Linked Data Fragments axis to discuss new approaches to publishing queryable public transit data for maximized use.
+We suggest a discussion for gathering query logs when using Linked Data Fragments approaches.
 
@@
 \label{sec:results}
+
 \todo{
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=8.1cm,align=center]{brussels}
+\caption{Double chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.
+Brussels acts as the central hub in the star-shaped railway network of Belgium.}
+\label{fig:brussels}
+\end{figure}
+
 \todo{This pair of chord diagrams capture the aggregated number of queries made between any two stations on weekdays and weekends between 6 to 9 in the morning.
@@
 
-\todo{add a couple of nice images with explanation here}
+\begin{figure}
+\centering
+\includegraphics[width=8.1cm,align=center]{antwerp}
+\caption{Double chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.
+The biggest transit flows to elsewhere in the province of Antwerp Central station is the hub for the province.}
+\label{fig:antwerp}
+\end{figure}
 
@@
 
-\section{Publishing transport data and the impact on query logs}
+\subsection{Pattern 3: cannot be true}
+\label{sec:pattern3}
+
+\section{Publishing transport data: the impact on query logs}
 \label{sec:publishing}
@@
 Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
-These trade-offs can be mapped on an axis, cfr. \cref{fig:LDFAxis2}, in which the 2 extremes are identified as: the server does all the processing versus the client does all the processing.
-
-Also for query logs we can make the same trade-off, in which the 2 extremes are the server knows only when someone downloads the entire datadump versus the server has a detailled query log.
-In this paper, we have studied the query logs of an API s
+These trade-offs can be mapped on an axis, cfr.
+\cref{fig:LDFAxis2}, in which the 2 extremes are identified as: the client does all the processing by downloading all the data at once, versus the server does all the processing by exposing \emph{expressive} APIs.
+Linked Connections~\cite{lc} proposes other ticks on this axis.
+By publishing fragments of the data needed for by the shortest path algorithm, the algorithm can be executed on the client.
+Furthermore, it lowers to load on the server, and it enables clients to federate queries over different servers, including different regions, transit modes and user's need.
 
@@
 \input{ldfaxis2}
+
+The expressivity of the server also affects the way logs can be gathered.
+When the server becomes less expressive, e.g., when it only allows to set a departure station and a departure time, then the server doesn't know the arrival station, which makes it harder to get an indication of the travel flow.
 

### 2015-12-21 paper.tex
small changes

@@
 % Perspectives
-In earlier work, we have suggested the Linked Data Fragments axis to discuss new approaches to publishing queryable public transit data for maximized use.
+In earlier work, we have suggested the Linked Data Fragments axis to discuss new approaches to publish queryable public transit data.
 We suggest a discussion for gathering query logs when using Linked Data Fragments approaches.
@@
 From an export of this table, we have filtered out only the route plannning queries that succeeded.
-The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/iraillogs.zip}}.
+The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 
@@
 Today, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-This amounts today to an average of 200k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\todo{\url{http://...}}} and \emph{Railer}\footnote{\todo{\url{http://...}}}.
+This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
+This amounts to $\sim11.4\%$ of all passengers\footnote{According to numbers of the railway maintainer Infrabel, 758,370 people take the train each day.}.
 

### 2015-12-22 paper.tex
Cleaned up a bit

@@
 Ever since, the project offers developer both a GTFS data dump for third party apps and a route planning API.
+The query logs of this API has been stored since 2013.
+Are these query logs interesting for urban planning purposes?
+Our hypothesis is that, while we should study the data with high level of sceptisicm (given enough data, you can find any pattern), our data will be better than random data.
 
-In this paper, we give a small overview of related work trying to gather data for flow analysis.
+In this paper, we give a small overview of related work to gather data for flow analysis.
 Next, we study the iRail query logs to find out whether we can find interesting patterns.
@@
 
-\section{Gathering the logs}
+\section{The logs}
 \label{sec:logs}
 
-In 2013, we have started to use a MySQL table to store each request in a table, including User-Agent string, the timestamp, whether or not the request succeeded, the departure station and the desired station of arrival.
-From an export of this table, we have filtered out only the route plannning queries that succeeded.
+In 2013, we have started to use a MySQL table to store each request to the iRail API in a table, including User-Agent string, the timestamp, whether or not the request succeeded, the departure station and the desired station of arrival.
+From an export of this table, we have filtered out only the route plannning queries that succeeded for the entire year.
 The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
@@
 The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
+Taking into account that 758,370 people take the train each day on a weekday (according to the railway network maintainer Infrabel) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.
 
-Today, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
 This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
-This amounts to $\sim11.4\%$ of all passengers\footnote{According to numbers of the railway maintainer Infrabel, 758,370 people take the train each day.}.
+On average, 14,357 of these are valid route planning queries (others include departures in a station, status of a vehicle, or a list of all stations).
+If each query would amount to indeed 1 passenger, this would today be equal to almost 2\% of all passengers.
+As we only gathered a small sample of this data by today, we didn't use the data of 2015, but used the data of 2013 instead, at the time when our servers didn't have caches in place.
 
-\section{Looking for patterns}
+There are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
+For instance, the data only captures an intention of a passenger: it's unsure whether the person actually took the train.
+Furthermore, different requests could be done by one person with an intention to travel.
+On top of that, our sample used for this paper can only represent a maximum of 0.4\% of the daily traffic flow.
+
+\section{Method}
 \label{sec:method}
 
-There are obvious caveats associated with the use of such data as proxy for actual statistical counts.
+In order to show that our data is better than random, we show that the data shows results that correspond with Belgian reality, whereas a random result would not.
+To do so, we have chosen to assemble the smaller stations under their lager administrative unit: provinces.
+The bigger cities, such as Ghent, Antwerp, Brussels, Mons and Liège, we kept.
+This way, we are able to create pair of chord diagrams.
+
+The random dataset was created as follows: \todo{...}
 
@@
 
-\todo{
-We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
-Look at weekday-mornings: commuting towards Gent, Brussels, Antwerp.
-This is not as obvious in the Walloon region (except for Liege).
-The reverse occurs on weekday-evenings.
-Local patterns can also be observed and this corresponds to what we know: "cities are the central hubs".
-Walloon region: everything converges on Brussels
-I found that whereas there's almost no movement from Antwerp city in the week to West Flanders, there is movement from Antwerp in the weekend to the coast and back. 
-}
-
 \begin{figure}
@@
 \includegraphics[width=8.1cm,align=center]{brussels}
-\caption{Double chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.
+\caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.
 Brussels acts as the central hub in the star-shaped railway network of Belgium.}
@@
 
-\todo{This pair of chord diagrams capture the aggregated number of queries made between any two stations on weekdays and weekends between 6 to 9 in the morning.
+The pair of chord diagrams in \cref{fig:brussels,fig:antwerp} capture the aggregated number of queries made between any two stations on weekdays and weekends between 6 to 9 in the morning.
 The chords that link two distinct stations, are coloured coded by region and proportionately scaled to the number of queries from a station of origin to a destination.
@@
 Queries from Walloonia with exception to Liege, are generally made from provincial stations towards Brussels instead of their respective provincial capitals.
-Queries in Flanders, on the hand, tend to be distributed among major cities otherwise known as the Flemish Diamond (Vlaamse Ruit), a network of four metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
-The difference between both patterns appear to correspond with existing measures of population density, providing valuable insight and alternative perspectives into the function of cities in rural and urban settings.}
+Queries in Flanders, on the hand, tend to be distributed among major cities otherwise known as the Flemish Diamond, a network of four metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
+The difference between both patterns appear to correspond with existing measures of population density, providing valuable insight and alternative perspectives into the function of cities in rural and urban settings.
 
+%We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
+Look at weekday-mornings: commuting towards Gent, Brussels, Antwerp.
+This is not as obvious in the Walloon region (except for Liege).
+The reverse occurs on weekday-evenings.
+Local patterns can also be observed and this corresponds to what we know: "cities are the central hubs".
+%Walloon region: everything converges on Brussels
+%I found that whereas there's almost no movement from Antwerp city in the week to West Flanders, there is movement from Antwerp in the weekend to the coast and back. 
 \begin{figure}
@@
 \includegraphics[width=8.1cm,align=center]{antwerp}
-\caption{Double chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.
+\caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.
 The biggest transit flows to elsewhere in the province of Antwerp Central station is the hub for the province.}
@@
 \end{figure}
-
-\subsection{Pattern 1: ...}
-\label{sec:pattern1}
-
-\subsection{Pattern 2: ...}
-\label{sec:pattern2}
-
-\subsection{Pattern 3: cannot be true}
-\label{sec:pattern3}
 

### 2015-12-22 paper.tex
Too tired - tomorrow another day

@@
 % Perspectives
-In earlier work, we have suggested the Linked Data Fragments axis to discuss new approaches to publish queryable public transit data.
-We suggest a discussion for gathering query logs when using Linked Data Fragments approaches.
+In earlier work, we suggested the Linked Data Fragments axis to discuss new approaches to publish queryable public transit data.
+Now, we suggest a discussion for gathering query logs when using Linked Data Fragments approaches.
 
@@
 Nevertheless, people are calculating their intended routes by using the Web.
+\emph{Can we get an indication of these transit flows by studying the query logs of Web-services?}
+
 The \emph{iRail} project\footnote{\url{http://hello.irail.be}} started in 2008 to make the data of the Belgian railway accessible for developers.
@@
 The query logs of this API has been stored since 2013.
-Are these query logs interesting for urban planning purposes?
 Our hypothesis is that, while we should study the data with high level of sceptisicm (given enough data, you can find any pattern), our data will be better than random data.
@@
 
-In 2013, we have started to use a MySQL table to store each request to the iRail API in a table, including User-Agent string, the timestamp, whether or not the request succeeded, the departure station and the desired station of arrival.
-From an export of this table, we have filtered out only the route plannning queries that succeeded for the entire year.
+In 2013, we started to use a MySQL table to store each request to the iRail API in a table, including User-Agent string, the timestamp, whether or not the request succeeded, the departure station and the desired station of arrival.
+From an export of this table, we filtered out only the route plannning queries that succeeded for the entire year.
 The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 
-We have cleaned this data by removing the requests done by harvesters or search engine bots.
+We cleaned this data by removing the requests done by harvesters or search engine bots.
 Thanks to the user agents, this was a fairly straight-forward operation.
@@
 In order to show that our data is better than random, we show that the data shows results that correspond with Belgian reality, whereas a random result would not.
-To do so, we have chosen to assemble the smaller stations under their lager administrative unit: provinces.
+To do so, we chose to assemble the smaller stations under their lager administrative unit: provinces.
 The bigger cities, such as Ghent, Antwerp, Brussels, Mons and Liège, we kept.

### 2015-12-22 paper.tex
Added histogram as an example
Update after confcall

@@
 % Conclusion
-Visualizing the data shows the query logs are a better representation than random data, yet more research is needed to understand how representative for travel flows they may be.
+We could not find a way \todo{to validate \ldots}, yet more research is needed to understand how representative for travel flows they may be.
 % Perspectives
@@
 \label{sec:introduction}
+
+\todo{Rework introduction at the end}
 
@@
 
+\todo{Write relwork at the end}
+
 \emph{Flow analysis} is a topic of theoretical interest and practical importance in various disciplines. 
@@
 
-There are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
-For instance, the data only captures an intention of a passenger: it's unsure whether the person actually took the train.
-Furthermore, different requests could be done by one person with an intention to travel.
-On top of that, our sample used for this paper can only represent a maximum of 0.4\% of the daily traffic flow.
+As we are dealing with query logs, privacy becomes an issue~\cite{silvestri}.
+However, we do not publish the IP addresses with each query, which makes it hard to track single users.
+To remove all doubt, we have asked and received the consent of the privacy commission of Belgium.
 
@@
 
-In order to show that our data is better than random, we show that the data shows results that correspond with Belgian reality, whereas a random result would not.
-To do so, we chose to assemble the smaller stations under their lager administrative unit: provinces.
-The bigger cities, such as Ghent, Antwerp, Brussels, Mons and Liège, we kept.
-This way, we are able to create pair of chord diagrams.
+\subsection{Processing}
+\todo{Each data record contains the names of both origin and destination (OD) stations, the user agent that made the query todo: Pieter this needs definition, as well as the time when the query was made.
 
-The random dataset was created as follows: \todo{...}
+As we were specifically interested in examining morning rush hour travel on weekdays, we filtered the dataset to exclude data records created outside the range of xxx hrs and xxx hrs on weekdays, and all data records created over the weekends.
+
+The iRail API serves a wide range of  applications that vary in terms of software architecture and functionality.
+End user applications such as BeTrains and Railer Pieter xxx and xxx query for time schedules when explicitly directed by the user.
+On the other hand, automated services like xxx make routine queries to gather data for further computation.
+Naturally, such automated queries are detected based on xxx.
+
+We exclude all automated queries to refine the dataset and remove potential bias.
+
+Queries made by automated user agents like xxx and xxx were also excluded from the dataset to remove biases that may over represent connections between certain stations.
+
+Additionally, data records with unparsable origin or destination station names due to spelling mistakes or invalid character encoding were also removed from the data set.}
+
+\subsection{Visualization}
+
+\todo{Alvin}
 
@@
 \centering
-\includegraphics[width=8.1cm,align=center]{brussels}
-\caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.
-Brussels acts as the central hub in the star-shaped railway network of Belgium.}
+\includegraphics[width=8.1cm,align=center]{histogram}
+\caption{Histogram of one weekday}
 \label{fig:brussels}
 \end{figure}
+
+In this chapter, we report on the results of studying visualizations with experts of Trein Tram Bus, a not for profit passenger federation.
+
+\subsection{Commuting pattern}
+
+We see people leaving in the morning and return in the evening (see \cref{fig:brussels} and todo)
 
@@
 
-%We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
+\subsection{Flanders is polycentric, Wallonia monocentric}
+
+We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
 Look at weekday-mornings: commuting towards Gent, Brussels, Antwerp.
@@
 The reverse occurs on weekday-evenings.
-Local patterns can also be observed and this corresponds to what we know: "cities are the central hubs".
-%Walloon region: everything converges on Brussels
-%I found that whereas there's almost no movement from Antwerp city in the week to West Flanders, there is movement from Antwerp in the weekend to the coast and back. 
+Local patterns can also be observed and this corresponds to what we know: ``cities are the central hubs''.
+Walloon region: everything converges on Brussels
+
+\subsection{Inexplicable results}
+
+We also see weird things happening in Antwerp, where a large amount of people go to the province in the morning on weekdays, while the oposite happens in Liege.
+When looking into the data, the Antwerp pattern appears due to one specific route which is queried each morning.
+Is there a group of colleagues all using the same app taking the same train daily?
+Is it someone who checks this trip over and over again for delays?
+We can only guess\ldots
+
+\begin{figure}
+\centering
+\includegraphics[width=8.1cm,align=center]{brussels}
+\caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.}
+\label{fig:brussels}
+\end{figure}
+
 \begin{figure}
@@
 \includegraphics[width=8.1cm,align=center]{antwerp}
-\caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.
-The biggest transit flows to elsewhere in the province of Antwerp Central station is the hub for the province.}
+\caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.}
 \label{fig:antwerp}
@@
 
-Thoughts on publishing transport data on the web and query logs in the case of GTFS, Linked Connections\cite{lc} (linked data fragments\cite{ldf}) and webservices.
+%\subsection{Intermodal route planning}
 
@@
 
-Query logs are very interesting
+The contribution of this paper is twofold.
+For the first time, we studied query logs to find travel patterns.
+Due to lack of information we could not evaluate our results, yet the visualizations look promising...
+There are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
+For instance, the data only captures an intention of a passenger: it's unsure whether the person actually took the train.
+Furthermore, different requests could be done by one person with an intention to travel.
+On top of that, our sample used for this paper can only represent a maximum of 0.4\% of the daily traffic flow.
+
+The second contribution is our position that we suggest to decouple the query logs from the query execution.
 

### 2015-12-22 paper.tex
remove random

@@
 The query logs of this API has been stored since 2013.
-Our hypothesis is that, while we should study the data with high level of sceptisicm (given enough data, you can find any pattern), our data will be better than random data.
+Our hypothesis is that, while we should study the data with high level of sceptisicm (given enough data, you can find any pattern), \todo{\ldots}
 

### 2015-12-25 paper.tex
Added Alvin input, reworked intro and added new section

@@
 
-Today, public transport data still remains absent from the Linked Open Data cloud\footnote{\url{http://lod-cloud.net/}}.
-Bizarre, one could notice, as many open transport datasets are already available, according to the Global Open Data Index of 2014\footnote{\url{http://index.okfn.org/dataset/timetables/2014/}}.
-We can imagine that, what contributed to this, are the de-facto standard ways to publish public transit data to the Web.
+%Today, public transport data still remains absent from the Linked Open Data cloud\footnote{\url{http://lod-cloud.net/}}.
+%Bizarre, one could notice, as many open transport datasets are already available, according to the Global Open Data Index of 2014\footnote{\url{http://index.okfn.org/dataset/timetables/2014/}}.
+%We can imagine that, what contributed to this, are the de-facto standard ways to publish public transit data to the Web.
 There are currently two ways, as illustrated in \cref{fig:LDFAxis1}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
 
-\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV-files, describing the rules for when a public transit vehicle will pass by on a certain location.
-The specification is a huge success: up to date, it is supported among all current open source route planning software systems, and it is used in products such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
-Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
-The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
+%\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV-files, describing the rules for when a public transit vehicle will pass by on a certain location.
+%The specification is a huge success: up to date, it is supported among all current open source route planning software systems, and it is used in products such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
+%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
+%The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
 
-On the other hand, public transit agencies also publish their data by providing \emph{route planners}.
-These route planner offer a query service on top of the data and expose these over the Web.
-However, when machines should access these, rate limiters are common place: only a limited amount of queries should be done in order for the service to be able to stay online.
+%On the other hand, public transit agencies also publish their data by providing \emph{route planners}.
+%These route planner offer a query service on top of the data and expose these over the Web.
+%However, when machines should access these, rate limiters are common place: only a limited amount of queries should be done in order for the service to be able to stay online.
 
-\input{ldfaxis1}
+%\input{ldfaxis1}
 
@@
 
-As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
-On average, 14,357 of these are valid route planning queries (others include departures in a station, status of a vehicle, or a list of all stations).
-If each query would amount to indeed 1 passenger, this would today be equal to almost 2\% of all passengers.
-As we only gathered a small sample of this data by today, we didn't use the data of 2015, but used the data of 2013 instead, at the time when our servers didn't have caches in place.
-
 As we are dealing with query logs, privacy becomes an issue~\cite{silvestri}.
@@
 \subsection{Processing}
-\todo{Each data record contains the names of both origin and destination (OD) stations, the user agent that made the query todo: Pieter this needs definition, as well as the time when the query was made.
+\todo{Alvin}
 
-As we were specifically interested in examining morning rush hour travel on weekdays, we filtered the dataset to exclude data records created outside the range of xxx hrs and xxx hrs on weekdays, and all data records created over the weekends.
+Each data record contains the names of both origin and destination (OD) stations, the user agent that made the query \todo{Pieter this needs definition}, as well as the time when the query was made.
 
-The iRail API serves a wide range of  applications that vary in terms of software architecture and functionality.
-End user applications such as BeTrains and Railer Pieter xxx and xxx query for time schedules when explicitly directed by the user.
-On the other hand, automated services like xxx make routine queries to gather data for further computation.
-Naturally, such automated queries are detected based on xxx.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 0600 hrs and 1000 hrs as well as 1700 hrs and 2100hrs on weekdays, and all data records created on weekend.
+Queries made by automated user agents like xxx and xxx were also excluded from the dataset to remove biases that may over represent connections between certain stations.
+Similarly, data records with unparsable origin or destination station names due to spelling mistakes or invalid character encoding were removed from the data set.
 
-We exclude all automated queries to refine the dataset and remove potential bias.
+The data is then simplified so that province level flow patterns are emphasised.
+Figure x provides a diagrammatic representation of the procedure.
+First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See Figure xa).
+If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (See Figure xb).
+Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (See Figure xc).
+The following types of flows are observable from the outcome of aggregation:
 
-Queries made by automated user agents like xxx and xxx were also excluded from the dataset to remove biases that may over represent connections between certain stations.
-
-Additionally, data records with unparsable origin or destination station names due to spelling mistakes or invalid character encoding were also removed from the data set.}
+\begin{itemize}
+  \item Travel from any provincial station to a major city.
+  \item Travel between any two major cities.
+  \item Travel between any two provincial stations.
+  \item Travel between any international station to a major city.
+  \item Travel between any international station to a provincial station.
+\end{itemize}
 
@@
 \todo{Alvin}
+
+Visualisation is frequently used to make data analysis tangible, so that the results can be communicated and debated {Robinson, 2008 #121}.
+This is crucial for our research since the questions presented are exploratory in nature {Kraak, 2008 #150} and can be addressed in many ways.
+Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis {Andrienko, 2012 #196}.
+Of these techniques, flow visualisations facilitate the comparison of aggregated movement over space and time.
+There are three types of flow visualisations: Line based representations preserves the complete trajectory while matrix type representations take only the start and end locations into account.
+
+Line based representations indicate movement on a path with the shape of the line.
+The thickness of each line is scaled to the volume of flow along a path.
+The direction of flow is commonly indicated with an arrowhead.
+This representation is detailed and straightforward to understand but becomes cluttered when lines intersect.
+A number of solutions have been proposed for clutter reduction.
+Filtering is the most commonly practiced.
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations {Andrienko, 2007 #198}.
+Hierarchical {Guo, 2009 #199} and density based {Rinzivillo, 2008 #201} clustering have also been introduced for this purpose.
+Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualisation.
+Another solution to reduce clutter is to transform the visual representation.
+This is achieved by rerouting {Phan, 2005 #88} or bundling {Hurter, 2014 #200} lines in close proximity so that they appear grouped.
+Unlike clustering, this solution changes the shape of lines by way of interpolation to obtain an ideal layout.
+In most cases, the interpolation is regulated by conditions to avoid intersections and follow key geographic features such as roads, rivers or coastlines so that the layout appears natural.
+This solution emphasises major flows within the visualisation since large groups gain visual dominance.
+Nonetheless, there are two drawbacks to this solution.
+First, short distance flows that may be important at local scales are de-emphasised or lost in the transformation.
+Next, the lines do not depict actual paths of movement thus this solution may not be suitable for applications that require strict geographic accuracy.
+
+Matrix type representations are frequently used when the origin and destination preside over intermediate locations on the path of movement.
+This type of representation is referred to as origin-destination (OD) matrix.
+Rows and columns correspond to locations while cells are coloured to express the volume of flow.
+OD matrices can be reordered to emphasise connectivity between locations but the lack of geographic context is a distinct disadvantage.
+Several solutions have been developed to address this shortcoming.
+These are based on the notion of small multiples: The matrix is arranged in a geographical order so that the cells correspond to locations on the map.
+Then, a smaller map of the geographical context is nested in each cell to enable more intuitive comparisons {Guo, 2006 #202}.
+The limitation of this solution is similar to other forms or small multiples in general - the layout space allocated to each nested map becomes smaller as the number of cells increase.
+In this instance, nesting detailed geographical maps is problematic since majority of the geographical features become illegible due to scaling.
+OD Map {Wood, 2010 #86} proposes to nests a series of geographically ordered matrices that are more expressive of spatial relations and remain legible even when a small amount of layout space allocated to each cell.
 
@@
 
-\section{Publishing transport data: the impact on query logs}
+\section{Publishing transport data}
 \label{sec:publishing}
 
-Publishing public transport data today happens in two ways: by publishing a data dump of the timetables or by hosting a public route planning service.
+It is in the interest of public transit agencies that the use of their time schedule data is maximized.
+We can imagine that when more people are informed about the agency's offer, there will be a higher amount of people that will use the service.
+There are currently two ways to publish transit data, as illustrated in \cref{fig:LDFAxis2}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
+
+\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
+GTFS is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
+%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
+The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
+
+On the other hand, public transit agencies also publish their data by providing \emph{route planners} such as the French railway company\footnote{\url{http://tech.eu/features/7119/sncf-api-real-time-data/}}, the Dutch railway company\footnote{\url{http://www.ns.nl/reisinformatie/ns-api}} or \ldots\footnote{\url{http://otherapiofadifferentagency}}
+These route planner offer a query service on top of the data and expose these over the Web.
+In this paper we have studied the query logs of such an \emph{expressive} route planning API, which per request offers an interesting log entry.
+
 Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
-These trade-offs can be mapped on an axis, cfr.
-\cref{fig:LDFAxis2}, in which the 2 extremes are identified as: the client does all the processing by downloading all the data at once, versus the server does all the processing by exposing \emph{expressive} APIs.
-Linked Connections~\cite{lc} proposes other ticks on this axis.
-By publishing fragments of the data needed for by the shortest path algorithm, the algorithm can be executed on the client.
-Furthermore, it lowers to load on the server, and it enables clients to federate queries over different servers, including different regions, transit modes and user's need.
+These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
+For route planning services, we call these other ticks Linked Connections~\cite{lc}: 
+by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading data just in time.
+This way, it lowers to load on the server, making it easier to keep the data high available.
+Furhtermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.
 
@@
 
-The expressivity of the server also affects the way logs can be gathered.
+The expressiveness of the server also affects the way logs can be gathered.
 When the server becomes less expressive, e.g., when it only allows to set a departure station and a departure time, then the server doesn't know the arrival station, which makes it harder to get an indication of the travel flow.
+However, from a Web-perspective, if we want to log every HTTP GET request, we expect every HTTP GET to reach the server, and we would have to deliberately disable caches to be able to log every request.
+Disabling this cache for high expressive servers is okay: the chance that two requests are exactly the same for a high expressive server is low, however from a low expressive server, we will have to do more requests to
 
@@
 %\subsection{Intermodal route planning}
+
+\section{Publishing query logs}
+\label{sec:publishingquerylogs}
+
+% Open Anonymized Route Planning Query Logs
+Semantics behind the query logs
+
+As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
+On average, 14,357 of these are valid route planning queries (others include departures in a station, status of a vehicle, or a list of all stations).
+If each query would amount to indeed 1 passenger, this would today be equal to almost 2\% of all passengers.
+As we only gathered a small sample of this data by today, we didn't use the data of 2015, but used the data of 2013 instead, at the time when our servers didn't have caches in place.
 

### 2015-12-26 ldfaxis2.tex
Added LC

@@
   \end{tikzpicture}
-  \caption{This axis extends \cref{fig:LDFAxis1} with Linked Connections: the different ticks on the axis for client effort versus server expressivity.}
+  \caption{This axis extends Linked Data Fragments with Linked Connections. Different ticks on the axis illustrate client effort versus server expressivity.}
   \label{fig:LDFAxis2}

### 2015-12-26 paper.tex
Added LC

@@
 
+%label: fig:LDFAxis2
+\input{ldfaxis2}
+
 Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
@@
 For route planning services, we call these other ticks Linked Connections~\cite{lc}: 
-by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading data just in time.
-This way, it lowers to load on the server, making it easier to keep the data high available.
-Furhtermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.
+by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
+An example of such an implementation is illustrated in \cref{fig:lc}.
+By exploiting the caching mechanisms behind HTTP, it lowers to load on the server, making it easier to keep the data high available.
+Furthermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.
 
-%label: fig:LDFAxis2
-\input{ldfaxis2}
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.48\textwidth]{LC}
+    \caption{An example implementation of a route planning algorithm: fragments of data needed by a route planning algorithm are published in pages.
+The route planning algorithm will need multiple requests instead of just one.
+The fragements of the pages that can be retrieved are however highly cacheable.}
+    \label{fig:lc}
+\end{figure}
 
 The expressiveness of the server also affects the way logs can be gathered.
-When the server becomes less expressive, e.g., when it only allows to set a departure station and a departure time, then the server doesn't know the arrival station, which makes it harder to get an indication of the travel flow.
-However, from a Web-perspective, if we want to log every HTTP GET request, we expect every HTTP GET to reach the server, and we would have to deliberately disable caches to be able to log every request.
-Disabling this cache for high expressive servers is okay: the chance that two requests are exactly the same for a high expressive server is low, however from a low expressive server, we will have to do more requests to
+If we want to log every HTTP GET request, we expect every HTTP GET to reach the server and we would have to deliberately disable caches to be able to log every request.
+The chance that two requests are exactly the same for a high expressive server is low. 
+We have shown in previous sections that these logs are interesting.
+
+A low expressive server that still publishes the data in a queryable way as illustrated in \cref{fig:lc}, exploits the caching mechanism 
+E.g., when it only allows to set a departure station and a departure time, then the server cannot know the arrival station, which makes it harder to get an indication of the travel flow.
 

### 2015-12-27 paper.tex
Rewrote abstract

@@
 The iRail project hosts a route planning API for the Belgian railway company.
-This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
+It is a high expresive webservice, of which the query logs are able to tell us something about the public transit demand at a certain moment.
+%This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
 % Need
 In the field of urban planning, researchers need an indication of how people move between cities. 
-%Yet, getting this data from official sources has proven to be troublesome.
+Yet, getting indications of these travel flows has proven to be troublesome.
 % Task
@@
 % Findings
-We illustrate \todo{X} interesting patterns that correspond to reality and we show an outlier in the data.
+Using data visualizations, we illustrate two interesting patterns that correspond to reality, as confirmed by experts, and we show one that does not.
 % Conclusion
-We could not find a way \todo{to validate \ldots}, yet more research is needed to understand how representative for travel flows they may be.
+More research is needed to understand how representative for travel flows query logs may be.
 % Perspectives
@@
 %We can imagine that, what contributed to this, are the de-facto standard ways to publish public transit data to the Web.
-There are currently two ways, as illustrated in \cref{fig:LDFAxis1}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
+%There are currently two ways, as illustrated in \cref{fig:LDFAxis1}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
 
@@
 The query logs of this API has been stored since 2013.
-Our hypothesis is that, while we should study the data with high level of sceptisicm (given enough data, you can find any pattern), \todo{\ldots}
+We study these query logs by creating a couple of visualizations which illustrate a couple of patterns.
+Two of the documented patterns in this paper correspond to reality, another does not.
 
@@
 These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
-For route planning services, we call these other ticks Linked Connections~\cite{lc}: 
+For route planning systems, we call these other ticks Linked Connections~\cite{lc}: 
 by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
@@
 If we want to log every HTTP GET request, we expect every HTTP GET to reach the server and we would have to deliberately disable caches to be able to log every request.
-The chance that two requests are exactly the same for a high expressive server is low. 
-We have shown in previous sections that these logs are interesting.
+The chance that two requests are exactly the same for a high expressive server is low, as there are more URLs which can be requested. 
+We have shown in previous sections that these logs are interesting: each request contains an entire query which can be used as an indication of travel ``demand''.
 
-A low expressive server that still publishes the data in a queryable way as illustrated in \cref{fig:lc}, exploits the caching mechanism 
+A low expressive server that still publishes the data in a queryable way as illustrated in \cref{fig:lc}, has less interesting results. 
 E.g., when it only allows to set a departure station and a departure time, then the server cannot know the arrival station, which makes it harder to get an indication of the travel flow.
+Furthermore, on the Linked Data Fragments axis, less expressive servers --which have less URLs to be requested-- can exploit the caching mechanism~\cite{ldf}.
+The drawback of having higher cacheability, is that intermediate proxies may have cached the HTTP response somewhere in between and there will not be a log entry.
+
+When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked Connections or a GTFS datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.
+
+\todo{\ldots}
 
@@
 
+As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
+On average, 14,357 of these are valid route planning queries (other queries include requesting departures in a station, status of a vehicle, or a list of all stations).
+If each query would amount to indeed one passenger, this would today be equal to almost 2\% of all passengers.
+
 % Open Anonymized Route Planning Query Logs
 Semantics behind the query logs
-
-As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
-On average, 14,357 of these are valid route planning queries (others include departures in a station, status of a vehicle, or a list of all stations).
-If each query would amount to indeed 1 passenger, this would today be equal to almost 2\% of all passengers.
-As we only gathered a small sample of this data by today, we didn't use the data of 2015, but used the data of 2013 instead, at the time when our servers didn't have caches in place.
 

### 2015-12-27 paper.tex
logs section

@@
 \newcommand{\todo}[1]{\noindent\textcolor{red}{{\bf \{TODO}: #1{\bf \}}}}
+\newdef{definition}{Definition}
+
 \begin{document}
@@
 
-In 2013, we started to use a MySQL table to store each request to the iRail API in a table, including User-Agent string, the timestamp, whether or not the request succeeded, the departure station and the desired station of arrival.
+The iRail API is a plain old XML/JSON HTTP API: one request documented for humans at $api.iRail.be$, will result in an answer that can be used directly in an end-user app.
+It contains four features: a route planning query, a query for the next departures in a certain station, a query for the status of a certain train and a query for a list of all stations.
+In 2013, we started to use a MySQL table to store each log entry to the iRail API in a table.
+The columns of this table are:
+\begin{itemize}
+  \item User-Agent (defined in \cref{def:ua}), 
+  \item the timestamp,
+  \item whether or not the request succeeded,
+  \item the departure station and
+  \item the desired station of arrival.
+\end{itemize}
 From an export of this table, we filtered out only the route plannning queries that succeeded for the entire year.
@@
 
-We cleaned this data by removing the requests done by harvesters or search engine bots.
-Thanks to the user agents, this was a fairly straight-forward operation.
-The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
-Taking into account that 758,370 people take the train each day on a weekday (according to the railway network maintainer Infrabel) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.
+\begin{definition}
+The User-Agent request-header field contains information about the user agent originating the request.
+This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations.
+\label{def:ua}
+\end{definition}
 
@@
 However, we do not publish the IP addresses with each query, which makes it hard to track single users.
-To remove all doubt, we have asked and received the consent of the privacy commission of Belgium.
+To remove all doubt, we have asked and received the consent of the privacy commission of Belgium, which declared we can study the logs without any restrictions.
 
@@
 \subsection{Processing}
-\todo{Alvin}
 
-Each data record contains the names of both origin and destination (OD) stations, the user agent that made the query \todo{Pieter this needs definition}, as well as the time when the query was made.
+We cleaned the logs by removing the requests done by harvesters or search engine bots.
+Thanks to the user agents, this was a fairly straight-forward operation.
+The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
+Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.
 
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 0600 hrs and 1000 hrs as well as 1700 hrs and 2100hrs on weekdays, and all data records created on weekend.
-Queries made by automated user agents like xxx and xxx were also excluded from the dataset to remove biases that may over represent connections between certain stations.
-Similarly, data records with unparsable origin or destination station names due to spelling mistakes or invalid character encoding were removed from the data set.
+Each data record contains the names of both origin and destination (OD) stations, the user agent that made the query, as well as the time when the query was made.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00~hrs and 10:00~hrs as well as 17:00~hrs and 21:00~hrs on weekdays, and all data records created on weekend.
+Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations.
+Similarly, data records with unparsable origin or destination station names due to spelling mistakes or invalid character encoding were removed as well.
 
 The data is then simplified so that province level flow patterns are emphasised.
-Figure x provides a diagrammatic representation of the procedure.
+\cref{fig:brussels} provides a diagrammatic representation of the procedure.
 First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See Figure xa).
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=8.1cm,align=center]{brussels}
+\caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.}
+\label{fig:brussels}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=8.1cm,align=center]{antwerp}
+\caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.}
+\label{fig:antwerp}
+\end{figure}
+
 We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
@@
 We can only guess\ldots
-
-\begin{figure}
-\centering
-\includegraphics[width=8.1cm,align=center]{brussels}
-\caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.}
-\label{fig:brussels}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=8.1cm,align=center]{antwerp}
-\caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.}
-\label{fig:antwerp}
-\end{figure}
 

### 2015-12-30 paper.tex
Added histogram spreadsheets

@@
 Next, we study the iRail query logs to find out whether we can find interesting patterns.
-Finally, we look at Linked Connections~\cite{lc}, a proposed new way of publishing queryable public transit data, and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments~\cite{ldf} approach.
+Finally, we look at Linked~Connections~\cite{lc}, a proposed new way of publishing queryable public transit data, and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments~\cite{ldf} approach.
 
@@
 \begin{itemize}
-  \item User-Agent (defined in \cref{def:ua}), 
+  \item User-Agent as defined by RFC2616\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}: \emph{the User-Agent request-header field contains information about the user agent originating the request.
+This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations};
   \item the timestamp,
@@
 \end{itemize}
+
 From an export of this table, we filtered out only the route plannning queries that succeeded for the entire year.
 The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
-
-\begin{definition}
-The User-Agent request-header field contains information about the user agent originating the request.
-This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations.
-\label{def:ua}
-\end{definition}
 
@@
 These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
-For route planning systems, we call these other ticks Linked Connections~\cite{lc}: 
+For route planning systems, we call these other ticks Linked~Connections~\cite{lc}: 
 by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
@@
 
-When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked Connections or a GTFS datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.
+When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked~Connections or a GTFS~datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.
 

### 2015-12-31 paper.tex
Sync

@@
 \begin{abstract}
-% Context
+
 %Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
-The iRail project hosts a route planning API for the Belgian railway company.
-It is a high expresive webservice, of which the query logs are able to tell us something about the public transit demand at a certain moment.
 %This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
-% Need
+% Context & % Need
 In the field of urban planning, researchers need an indication of how people move between cities. 
-Yet, getting indications of these travel flows has proven to be troublesome.
+Yet, getting statistics of travel flows has proven to be troublesome.
+The iRail project hosts a high expressive route planning API for the Belgian railway company.
 % Task
-In this paper, we analyse the query logs of the iRail API, to see whether it can tell us something about travel flows between cities in Belgium.
+We analyse the query logs of this API, to get an indication of travel flows between cities in Belgium.
+We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which amount to \todo{between $0.3\%$ and $2\%$} of the amount of monthly passengers.
 % Object
-We studied queries for the year 2013, which contains an average of $\sim3000$ interesting queries per day ($\sim0.4\%$ of the amount of passengers per weekday).
+Using data visualizations, we illustrate the commuting patterns in Belgium and show that Brussels, the captial, acts as a central hub. 
 % Findings
-Using data visualizations, we illustrate two interesting patterns that correspond to reality, as confirmed by experts, and we show one that does not.
+The visualizations correspond to reality, as confirmed by experts. 
 % Conclusion
@@
 
+\begin{figure}
+\centering
+\includegraphics[width=193,align=center]{querylogs}
+\caption{Amount of queries on the route planning part of the iRail API between 2012 and 2016.
+In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
+\label{fig:querylogs}
+\end{figure}
+
 From an export of this table, we filtered out only the route plannning queries that succeeded for the entire year.
@@
 To remove all doubt, we have asked and received the consent of the privacy commission of Belgium, which declared we can study the logs without any restrictions.
+
+\begin{figure}
+\centering
+\includegraphics[width=8.3cm,align=center]{avg-all}
+\caption{Distribution of iRail route planning API queries on average per day of the week between 2012 and 2016}
+\label{fig:average}
+\end{figure}
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=8.3cm,align=center]{dec21}
+%\caption{Comparison of route planning queries on December 21st 2015 with the average on a Monday between 2012 and 2016.
+Whilst being a school holiday, there is no notable difference, except for the peek hours being a bit later, }
+%\label{fig:dec21}
+%\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=8.3cm,align=center]{dec24}
+\caption{Comparison of route planning queries on December 24th (Christmass eve) 2015 with the average on a Thursday between 2012 and 2016.
+The illustration shows that people started going home earlier than usual.}
+\label{fig:dec24}
+\end{figure}
 
@@
 \label{sec:results}
-
-\begin{figure}
-\centering
-\includegraphics[width=8.1cm,align=center]{histogram}
-\caption{Histogram of one weekday}
-\label{fig:brussels}
-\end{figure}
 

### 2015-12-31 paper.tex
Abstract rework, EPS files added

@@
 In the field of urban planning, researchers need an indication of how people move between cities. 
-Yet, getting statistics of travel flows has proven to be troublesome.
-The iRail project hosts a high expressive route planning API for the Belgian railway company.
+Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
 % Task
-We analyse the query logs of this API, to get an indication of travel flows between cities in Belgium.
-We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which amount to \todo{between $0.3\%$ and $2\%$} of the amount of monthly passengers.
-% Object
-Using data visualizations, we illustrate the commuting patterns in Belgium and show that Brussels, the captial, acts as a central hub. 
-% Findings
-The visualizations correspond to reality, as confirmed by experts. 
+We analysed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
+We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
+% Object & Findings
+Using data visualizations, we illustrated the commuting patterns in Belgium and show that Brussels, the capital, acts as a central hub.
+The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
+The findings correspond to the real travel demand, say experts of the passenger federation Trein Tram Bus.
 % Conclusion
-More research is needed to understand how representative for travel flows query logs may be.
+We concluded that query logs of route planners are of high importance in getting an indication of the travel flows.
 % Perspectives
-In earlier work, we suggested the Linked Data Fragments axis to discuss new approaches to publish queryable public transit data.
-Now, we suggest a discussion for gathering query logs when using Linked Data Fragments approaches.
+\todo{another sentence about publishing transit data for maximized reuse and how to gather analytics}
 
@@
 
-The iRail API is a plain old XML/JSON HTTP API: one request documented for humans at $api.iRail.be$, will result in an answer that can be used directly in an end-user app.
-It contains four features: a route planning query, a query for the next departures in a certain station, a query for the status of a certain train and a query for a list of all stations.
-In 2013, we started to use a MySQL table to store each log entry to the iRail API in a table.
-The columns of this table are:
+The iRail API is a XML/JSON HTTP API in which one request on \url{http://api.irail.be} will result in a response that can be used directly in an end-user app.
+It contains four features: a route planning query (\url{http://api.irail.be/connections/{?from,to,date,time}}), a query for the next departures in a certain station (\url{http://api.irail.be/liveboard/{?station,date,time}}), a query for the status of a certain train (\url{http://api.irail.be/vehicle/{?id,date}}) and a query for a list of all stations (\url{http://api.irail.be/stations/}).
+We have gathered the \emph{Apache} \emph{access.log} files 
 \begin{itemize}
-  \item User-Agent as defined by RFC2616\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}: \emph{the User-Agent request-header field contains information about the user agent originating the request.
-This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations};
   \item the timestamp,
   \item whether or not the request succeeded,
-  \item the departure station and
-  \item the desired station of arrival.
+  \item the path of the query (e.g., \emph{/connections/?from=Ghent\&to=Antwerp}) and
+  \item the User-Agent (cfr.
+RFC2616\footnote{The User-Agent request-header field contains information about the user agent originating the request.
+This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations -- \url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}})
 \end{itemize}
@@
 \centering
-\includegraphics[width=193,align=center]{querylogs}
+\includegraphics[width=8.6cm,align=center]{querylogs}
 \caption{Amount of queries on the route planning part of the iRail API between 2012 and 2016.
@@
 \centering
-\includegraphics[width=8.3cm,align=center]{avg-all}
+\includegraphics[width=8.6cm,align=center]{avg-all}
 \caption{Distribution of iRail route planning API queries on average per day of the week between 2012 and 2016}
@@
 \centering
-\includegraphics[width=8.3cm,align=center]{dec24}
+\includegraphics[width=8.6cm,align=center]{dec24}
 \caption{Comparison of route planning queries on December 24th (Christmass eve) 2015 with the average on a Thursday between 2012 and 2016.

### 2016-01-02 ldfaxis2.tex
abstract and wrote publishing transport data

--- 2015-12-21 ldfaxis2.tex
@@
       low server availability\\
-      detailled query logs
+      detailed query logs
     };
@@
   \end{tikzpicture}
-  \caption{This axis extends \cref{fig:LDFAxis1} with Linked Connections: the different ticks on the axis for client effort versus server expressivity.}
+  \caption{This axis, first introduced by Linked Data Fragments~\cite{ldf}, illustrates the possibilities within publishing transport data.
+Different ticks on the axis illustrate client effort versus server expressivity: on the far left, data dumps offer high server availability, yet the effort needed by data reusers is high, and query logs do not reach the server.
+Moving to the right, we identify Linked Connections~\cite{lc} as an in-between solution.
+On the far right, route planning services require high server effort, leading to detailed query logs.}
   \label{fig:LDFAxis2}

### 2016-01-02 paper.tex
abstract and wrote publishing transport data

@@
 % Task
-We analysed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
+We analyzed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
 We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
@@
 % Conclusion
-We concluded that query logs of route planners are of high importance in getting an indication of the travel flows.
+We conclude that query logs of route planners are of high importance in getting an indication of the travel flows.
 % Perspectives
-\todo{another sentence about publishing transit data for maximized reuse and how to gather analytics}
+High expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
+In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
 
@@
 Thanks to the user agents, this was a fairly straight-forward operation.
-The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
-Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.
+\todo{The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
+Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.}\footnote{Calculated by using indicators of the Flemish government and extrapolating the numbers for 2015 \url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}
 
@@
 
-It is in the interest of public transit agencies that the use of their time schedule data is maximized.
-We can imagine that when more people are informed about the agency's offer, there will be a higher amount of people that will use the service.
-There are currently two ways to publish transit data, as illustrated in \cref{fig:LDFAxis2}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
+%It is in the interest of public transit agencies that the use of their time schedule data is maximized.
+%We can imagine that when more people are informed about the agency's offer, there will be a higher amount of people that will use the service.
+%There are currently two ways to publish transit data, as illustrated in \cref{fig:LDFAxis2}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
 
-\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
-GTFS is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
+%\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
+%GTFS is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
 %Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
-The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
+%The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
 
-On the other hand, public transit agencies also publish their data by providing \emph{route planners} such as the French railway company\footnote{\url{http://tech.eu/features/7119/sncf-api-real-time-data/}}, the Dutch railway company\footnote{\url{http://www.ns.nl/reisinformatie/ns-api}} or \ldots\footnote{\url{http://otherapiofadifferentagency}}
-These route planner offer a query service on top of the data and expose these over the Web.
-In this paper we have studied the query logs of such an \emph{expressive} route planning API, which per request offers an interesting log entry.
+%On the other hand, public transit agencies also publish their data by providing \emph{route planners} such as the French railway company\footnote{\url{http://tech.eu/features/7119/sncf-api-real-time-data/}}, the Dutch railway company\footnote{\url{http://www.ns.nl/reisinformatie/ns-api}} or \ldots\footnote{\url{http://otherapiofadifferentagency}}
+%These route planner offer a query service on top of the data and expose these over the Web.
+%In this paper we have studied the query logs of such an \emph{expressive} route planning API, which per request offers an interesting log entry.
 
@@
 
-Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
-These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
-For route planning systems, we call these other ticks Linked~Connections~\cite{lc}: 
-by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
-An example of such an implementation is illustrated in \cref{fig:lc}.
-By exploiting the caching mechanisms behind HTTP, it lowers to load on the server, making it easier to keep the data high available.
-Furthermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.
+%Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
+%These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
+%For route planning systems, we call these other ticks Linked~Connections~\cite{lc}: 
+%by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
+%An example of such an implementation is illustrated in \cref{fig:lc}.
+%By exploiting the caching mechanisms behind HTTP, it lowers to load on the server, making it easier to keep the data high available.
+%Furthermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=0.48\textwidth]{LC}
-    \caption{An example implementation of a route planning algorithm: fragments of data needed by a route planning algorithm are published in pages.
+%\begin{figure}[h]
+%    \centering
+%    \includegraphics[width=0.48\textwidth]{LC}
+%    \caption{An example implementation of a route planning algorithm: fragments of data needed by a route planning algorithm are published in pages.
 The route planning algorithm will need multiple requests instead of just one.
 The fragements of the pages that can be retrieved are however highly cacheable.}
-    \label{fig:lc}
-\end{figure}
+%    \label{fig:lc}
+%\end{figure}
+%
 
-The expressiveness of the server also affects the way logs can be gathered.
-If we want to log every HTTP GET request, we expect every HTTP GET to reach the server and we would have to deliberately disable caches to be able to log every request.
-The chance that two requests are exactly the same for a high expressive server is low, as there are more URLs which can be requested. 
-We have shown in previous sections that these logs are interesting: each request contains an entire query which can be used as an indication of travel ``demand''.
+%%% EXPLAIN:
+%%%
+The expressiveness of a server affects the way logs can be gathered.
+We have shown in previous sections that the query logs of a \emph{high expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
+The possibility that two requests are exactly the same is low, as there are many URLs which can be requested.
+Nevertheless, we cannot fully guarantee that each HTTP GET request to the server will trigger a log entry, due to caching mechanisms on the Web~\cite{fielding}, which might lead to a false representation.
 
-A low expressive server that still publishes the data in a queryable way as illustrated in \cref{fig:lc}, has less interesting results. 
-E.g., when it only allows to set a departure station and a departure time, then the server cannot know the arrival station, which makes it harder to get an indication of the travel flow.
-Furthermore, on the Linked Data Fragments axis, less expressive servers --which have less URLs to be requested-- can exploit the caching mechanism~\cite{ldf}.
-The drawback of having higher cacheability, is that intermediate proxies may have cached the HTTP response somewhere in between and there will not be a log entry.
+%intermodality, features and availability
+Hosting a route planning API as the only way to publish transport data comes with three identified limitations, as the server will need to handle the requests from different use cases with different needs:
+\begin{enumerate}
+  \item When an application developer would like a \emph{new feature}, such as taking wheelchair accessibility information into account, the feature would have to be implemented on the server of the data publisher
+  \item Keeping the server \emph{high available} is costly, as any question can be asked by anyone for any purpose.
+  \item Federated querying, which would allow for \emph{intermodal} route planning for route planning APIs, is unexistent up to date.
+\end{enumerate}
 
-When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked~Connections or a GTFS~datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.
+% GTFS
+In order to overcome these limitations, the General Transit Feed Specification (GTFS)\footnote{\url{https://developers.google.com/transit/gtfs/reference}} can be used. 
+GTFS is a compressed ZIP-file containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
+It is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
+%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
+%The goal of the format is to  an exchange format which specific software packages can use to transform it into their format of choice.
+It succesfully enabled reuse for intermodal travel, engineers can rely on the data dumps even if the servers of the transit agency are offline and there is no limitation to the features that can be implemented.
+The logs are however lost.
 
-\todo{\ldots}
-
-\todo{Also in USEWOD2014 there was a paper on LDF and how we could know which SPARQL queries needed to be solved}
-
-%\subsection{Intermodal route planning}
+In \cref{fig:LDFAxis2} we illustrate these two options as two extremes, with other options that are yet to be discovered.
+When a \emph{rather low expressive} server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
+Again, we are not able to rely on the query logs
 
@@
 
-As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+\todo{When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked~Connections or a GTFS~datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.}
+
+\todo{Also in USEWOD2014 there was a paper on LDF and how we could know which SPARQL queries needed to be solved}
+
+\todo{As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
 This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
 On average, 14,357 of these are valid route planning queries (other queries include requesting departures in a station, status of a vehicle, or a list of all stations).
-If each query would amount to indeed one passenger, this would today be equal to almost 2\% of all passengers.
-
-% Open Anonymized Route Planning Query Logs
-Semantics behind the query logs
+If each query would amount to indeed one passenger, this would today be equal to almost 2\% of all passengers.}
 

### 2016-01-03 paper.tex
dot
align center doesn't work
Temporally resolved error returned by "#" & "Package pgfkeys Error: The key '/tikz/text' requires a value."

@@
 \centering
-\includegraphics[width=8.6cm,align=center]{querylogs}
+\includegraphics[width=8.6cm]{querylogs}
 \caption{Amount of queries on the route planning part of the iRail API between 2012 and 2016.
@@
 \centering
-\includegraphics[width=8.6cm,align=center]{avg-all}
+\includegraphics[width=8.6cm]{avg-all}
 \caption{Distribution of iRail route planning API queries on average per day of the week between 2012 and 2016}
@@
 %\centering
-%\includegraphics[width=8.3cm,align=center]{dec21}
+%\includegraphics[width=8.3cm]{dec21}
 %\caption{Comparison of route planning queries on December 21st 2015 with the average on a Monday between 2012 and 2016.
@@
 \centering
-\includegraphics[width=8.6cm,align=center]{dec24}
+\includegraphics[width=8.6cm]{dec24}
 \caption{Comparison of route planning queries on December 24th (Christmass eve) 2015 with the average on a Thursday between 2012 and 2016.
@@
 
-Visualisation is frequently used to make data analysis tangible, so that the results can be communicated and debated {Robinson, 2008 #121}.
-This is crucial for our research since the questions presented are exploratory in nature {Kraak, 2008 #150} and can be addressed in many ways.
-Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis {Andrienko, 2012 #196}.
+Visualisation is frequently used to make data analysis tangible, so that the results can be communicated and debated \todo{Robinson, 2008 -121}.
+This is crucial for our research since the questions presented are exploratory in nature \todo{Kraak, 2008 -150} and can be addressed in many ways.
+Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \todo{Andrienko, 2012 -196}.
 Of these techniques, flow visualisations facilitate the comparison of aggregated movement over space and time.
@@
 Filtering is the most commonly practiced.
-The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations {Andrienko, 2007 #198}.
-Hierarchical {Guo, 2009 #199} and density based {Rinzivillo, 2008 #201} clustering have also been introduced for this purpose.
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \todo{Andrienko, 2007 -198}.
+Hierarchical \todo{Guo, 2009 -199} and density based \todo{Rinzivillo, 2008 -201} clustering have also been introduced for this purpose.
 Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualisation.
 Another solution to reduce clutter is to transform the visual representation.
-This is achieved by rerouting {Phan, 2005 #88} or bundling {Hurter, 2014 #200} lines in close proximity so that they appear grouped.
+This is achieved by rerouting \todo{Phan, 2005 -88} or bundling \todo{Hurter, 2014 -200} lines in close proximity so that they appear grouped.
 Unlike clustering, this solution changes the shape of lines by way of interpolation to obtain an ideal layout.
@@
 These are based on the notion of small multiples: The matrix is arranged in a geographical order so that the cells correspond to locations on the map.
-Then, a smaller map of the geographical context is nested in each cell to enable more intuitive comparisons {Guo, 2006 #202}.
+Then, a smaller map of the geographical context is nested in each cell to enable more intuitive comparisons \todo{Guo, 2006 -202}.
 The limitation of this solution is similar to other forms or small multiples in general - the layout space allocated to each nested map becomes smaller as the number of cells increase.
 In this instance, nesting detailed geographical maps is problematic since majority of the geographical features become illegible due to scaling.
-OD Map {Wood, 2010 #86} proposes to nests a series of geographically ordered matrices that are more expressive of spatial relations and remain legible even when a small amount of layout space allocated to each cell.
+OD Map \todo{Wood, 2010 -86} proposes to nests a series of geographically ordered matrices that are more expressive of spatial relations and remain legible even when a small amount of layout space allocated to each cell.
 
@@
 \centering
-\includegraphics[width=8.1cm,align=center]{brussels}
+\includegraphics[width=8.1cm]{brussels}
 \caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.}
@@
 \centering
-\includegraphics[width=8.1cm,align=center]{antwerp}
+\includegraphics[width=8.1cm]{antwerp}
 \caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.}
@@
 When a \emph{rather low expressive} server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
-Again, we are not able to rely on the query logs
+Again, we are not able to rely on the query logs.
 

### 2016-01-04 paper.tex
Query logs chapter written

--- 2016-01-02 paper.tex
@@
 
-\section{The logs}
+\section{The query logs}
 \label{sec:logs}
@@
 It contains four features: a route planning query (\url{http://api.irail.be/connections/{?from,to,date,time}}), a query for the next departures in a certain station (\url{http://api.irail.be/liveboard/{?station,date,time}}), a query for the status of a certain train (\url{http://api.irail.be/vehicle/{?id,date}}) and a query for a list of all stations (\url{http://api.irail.be/stations/}).
-We have gathered the \emph{Apache} \emph{access.log} files 
+We have gathered the \emph{Apache} \emph{access.log} files from the past years. 
+These log files contain:
 \begin{itemize}
@@
   \item whether or not the request succeeded,
-  \item the path of the query (e.g., \emph{/connections/?from=Ghent\&to=Antwerp}) and
+  \item the path of the request (e.g., \url{/connections/?from=Ghent\&to=Antwerp}) and
   \item the User-Agent (cfr.
@@
 
+For this paper, we were only interested in route planning requests and we filtered out the requests to \emph{connections/}.
+We then parsed the \emph{access.log} path of the request and filtered out the from and to. 
+The resulting dataset, in CSV, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+
+As we are dealing with query logs, privacy becomes an issue~\cite{silvestri}.
+However, we do not publish the IP addresses with each query, which makes it hard to track single users.
+To remove all doubt, we have asked and received the consent of the Belgian privacy commission, which declared we can study the logs without any restrictions.
+
+We researched these three initial questions on this dataset:
+\begin{enumerate}
+  \item How many queries were done monthly and how does this compare with the real number of passengers?
+  \item How are the requests distributed within a day?
+  \item Can we see deviations on an average request distribution on e.g., public holidays?
+\end{enumerate}
+
 \begin{figure}
 \centering
-\includegraphics[width=8.6cm,align=center]{querylogs}
+\includegraphics[width=8.6cm]{querylogs}
 \caption{Amount of queries on the route planning part of the iRail API between 2012 and 2016.
@@
 
-From an export of this table, we filtered out only the route plannning queries that succeeded for the entire year.
-The resulting dataset, used for the rest of this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
-
-As we are dealing with query logs, privacy becomes an issue~\cite{silvestri}.
-However, we do not publish the IP addresses with each query, which makes it hard to track single users.
-To remove all doubt, we have asked and received the consent of the privacy commission of Belgium, which declared we can study the logs without any restrictions.
+In \cref{fig:querylogs} we show the amount of queries per month.
+We can see a sudden grow in April 2015: this is when an official app of the Belgian railway company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}} and many people started looking for an alternative within the apps built on top of the iRail API.
+In 2012, 2013, 2014 and 2015, there were respectively 19.3, 19.4, 19,6 and 19.7 million passengers per month on average\footnote{Calculated by using indicators of the Flemish government and extrapolating the numbers for 2015 \url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}.
+The amount of iRail queries per month, on average in these years are respectively 0.11, 0.17, 0.15 and 0.33 million.
+If we assume that each request is a travel intention, this would result in respectively $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the travels that are captured with the iRail API.
 
@@
 
-%\begin{figure}
-%\centering
-%\includegraphics[width=8.3cm,align=center]{dec21}
-%\caption{Comparison of route planning queries on December 21st 2015 with the average on a Monday between 2012 and 2016.
-Whilst being a school holiday, there is no notable difference, except for the peek hours being a bit later, }
-%\label{fig:dec21}
-%\end{figure}
+Next, we made a distribution of the requests done on each day of the week, which resulted in \cref{fig:average}.
+In this chart we can spot, on Monday until Friday, the morning peak between 5 and 9 and the evening peak between 15 and 18.
+We see a small dent during noon \todo{any explanation for this?}.
+On Fridays we see that the morning peak is stronger than the morning peak, which can be explained by many students that go home to visit their parents for the weekend \todo{Can we give a link to more reading material about this behaviour?
+This might be weird for US people?}.
+On Saturdays, we see almost do not see the peaks, which can be explained by people not going to work during weekends.
+On Sundays, we see most requests happen in the evening, which can be explained by students going back to their student homes.
 
@@
 \centering
-\includegraphics[width=8.6cm,align=center]{dec24}
-\caption{Comparison of route planning queries on December 24th (Christmass eve) 2015 with the average on a Thursday between 2012 and 2016.
+\includegraphics[width=8.6cm]{dec24}
+\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between 2012 and 2016.
 The illustration shows that people started going home earlier than usual.}
@@
 
+Finally, we compared the average distribution of Thursdays with the distribution on Thursday the 24th of December 2015 (Christmass Eve).
+Our hypothesis that we would now see an earlier evening peak can be confirmed by \cref{fig:dec24}.
+
+We conclude that the timing of the queries look interesting for flow analysis. 
+Can we also find interesting patterns when we study where people are heading between certain time intervals?
+
 \section{Method}
 \label{sec:method}
+
+For flow analysis, we need to analyse where people are going on what part of the day.
+In this section we first discuss how we processed the data in order to find the locations people are heading to and then discuss how we visualise the data in order to show proper results.
 
@@
 \todo{The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
-Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.}\footnote{Calculated by using indicators of the Flemish government and extrapolating the numbers for 2015 \url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}
+Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.}
 
@@
 \label{sec:publishingquerylogs}
+
+The iRail project is a mediator: it is a not for profit organization to stimulate digital creativity concerning mobility.
+The interface that takes care of the route planning interface is open source and we suggested a change to the system in which the query logs become open\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
+As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
 
@@
 
-\section{Conclusion and discussion}
+\section{Conclusion and future work}
 \label{sec:conclusion}
@@
 
+In future work, iRail will try to use these to predict train congestion.
 % Fix spacing after References header (as line 1308 of sig-alternate.cls breaks it)

### 2016-01-04 paper.tex
Cleaned method section

@@
 
-We cleaned the logs by removing the requests done by harvesters or search engine bots.
-Thanks to the user agents, this was a fairly straight-forward operation.
-\todo{The remainder of the file delivers an average of 3000 route planning queries a day, which we can use an an indication of travel intensities.
-Taking into account that 758,370 people take the train each day on a weekday (according to official number of the Belgian Railway company in 2013\footnote{\url{http://www.treintrambus.be/images/OpstappendeReizigers_2013_2.pdf}}) and assuming that each query is a trip done by a certain passenger, this sample equals $\sim0.4\%$ of the railway users in Belgium.}
+We processed the the data so that for both every origin and destination we have a unique identifier and the longitude and latitude.
+We have done this using the open-source reconciliator at \url{https://github.com/irail/stations}.
 
-Each data record contains the names of both origin and destination (OD) stations, the user agent that made the query, as well as the time when the query was made.
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00~hrs and 10:00~hrs as well as 17:00~hrs and 21:00~hrs on weekdays, and all data records created on weekend.
+Each data record now contains the station object of both origin and destination (OD), the user agent that made the query, as well as the time when the query was made.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of \todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 06:00~hrs and 10:00~hrs as well as 17:00~hrs and 21:00~hrs on weekdays, and all data records created on weekend.
 Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations.

### 2016-01-04 paper.tex

@@
 
-We researched these three initial questions on this dataset:
+We established three preliminary questions to  investigate the potential of this dataset:
 \begin{enumerate}
@@
 This might be weird for US people?}.
-On Saturdays, we see almost do not see the peaks, which can be explained by people not going to work during weekends.
-On Sundays, we see most requests happen in the evening, which can be explained by students going back to their student homes.
+On Saturdays, we almost do not see the peaks, which can be explained by people not going to work on weekends.
+On Sundays, we see most requests happen in the evening, which can be explained by students returning to college accomodations.
 

### 2016-01-04 paper.tex
Modified description of dataset.

@@
 
-The iRail API is a XML/JSON HTTP API in which one request on \url{http://api.irail.be} will result in a response that can be used directly in an end-user app.
+The iRail API is a XML/JSON HTTP API where each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
 It contains four features: a route planning query (\url{http://api.irail.be/connections/{?from,to,date,time}}), a query for the next departures in a certain station (\url{http://api.irail.be/liveboard/{?station,date,time}}), a query for the status of a certain train (\url{http://api.irail.be/vehicle/{?id,date}}) and a query for a list of all stations (\url{http://api.irail.be/stations/}).
-We have gathered the \emph{Apache} \emph{access.log} files from the past years. 
-These log files contain:
+We have gathered the \emph{Apache} \emph{access.log} files generated between 2012 and 2015 containing: 
 \begin{itemize}
-  \item the timestamp,
-  \item whether or not the request succeeded,
-  \item the path of the request (e.g., \url{/connections/?from=Ghent\&to=Antwerp}) and
-  \item the User-Agent (cfr.
+  \item Timestamp,
+  \item HTTP Status Code,
+  \item Request Path (e.g., \url{/connections/?from=Ghent\&to=Antwerp})
+  \item User-Agent (cfr.
 RFC2616\footnote{The User-Agent request-header field contains information about the user agent originating the request.
@@
 
-For this paper, we were only interested in route planning requests and we filtered out the requests to \emph{connections/}.
-We then parsed the \emph{access.log} path of the request and filtered out the from and to. 
-The resulting dataset, in CSV, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+We are particularly intersted in studying the potential application of route planning queries for flow analysis thus  only requests to (\url{http://api.irail.be/connections/{?from,to,date,time}}) were extracted for further investigation.
+The resulting dataset described in this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 

### 2016-01-04 paper.tex
Modified iRail API feature section.

@@
 The iRail API is a XML/JSON HTTP API where each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
-It contains four features: a route planning query (\url{http://api.irail.be/connections/{?from,to,date,time}}), a query for the next departures in a certain station (\url{http://api.irail.be/liveboard/{?station,date,time}}), a query for the status of a certain train (\url{http://api.irail.be/vehicle/{?id,date}}) and a query for a list of all stations (\url{http://api.irail.be/stations/}).
+It contains four features: 
+\begin{itemize}
+\item Route Planning Feature (\url{http://api.irail.be/connections/{?from,to,date,time}}).
+\item Next Departures Feature (\url{http://api.irail.be/liveboard/{?station,date,time}}).
+\item Train Status Feature query (\url{http://api.irail.be/vehicle/{?id,date}}).
+\item List All Stations Feature (\url{http://api.irail.be/stations/}).
+\end{itemize}
+
+\todo{what are the intended functions of each feature?
+We should mention that the route planning feature is perhaps the most popular among lay users?
+Thus is also has good potential to be used as a proxy for analysis...}
+
+We are particularly intersted in studying the potential application of route planning queries for flow analysis thus  only requests to (\url{http://api.irail.be/connections/{?from,to,date,time}}) were extracted for further investigation.
+The resulting dataset described in this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+
 We have gathered the \emph{Apache} \emph{access.log} files generated between 2012 and 2015 containing: 
@@
 \end{itemize}
-
-We are particularly intersted in studying the potential application of route planning queries for flow analysis thus  only requests to (\url{http://api.irail.be/connections/{?from,to,date,time}}) were extracted for further investigation.
-The resulting dataset described in this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 

### 2016-01-04 paper.tex
Modified "Method" introduction

@@
 It contains four features:
+
 \begin{itemize}
-\item Route Planning Feature (\url{http://api.irail.be/connections/{?from,to,date,time}}).
-\item Next Departures Feature (\url{http://api.irail.be/liveboard/{?station,date,time}}).
-\item Train Status Feature query (\url{http://api.irail.be/vehicle/{?id,date}}).
-\item List All Stations Feature (\url{http://api.irail.be/stations/}).
+\item Route Planning (\url{http://api.irail.be/connections/{?from,to,date,time}}).
+\item Next Departures (\url{http://api.irail.be/liveboard/{?station,date,time}}).
+\item Train Status (\url{http://api.irail.be/vehicle/{?id,date}}).
+\item List All Stations (\url{http://api.irail.be/stations/}).
 \end{itemize}
@@
 
-We have gathered the \emph{Apache} \emph{access.log} files generated between 2012 and 2015 containing: 
+This dataset is compiled from the \emph{Apache} \emph{access.log} files generated between 2012 and 2015. Each data entry contains a: 
 \begin{itemize}
@@
 
-For flow analysis, we need to analyse where people are going on what part of the day.
-In this section we first discuss how we processed the data in order to find the locations people are heading to and then discuss how we visualise the data in order to show proper results.
+Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame.
+Data visualisation is typically employed to facilate in this process thus the input data must be transformed into a format that expresses the quantity and directional flow of human movment.
+In this section we describe the data transformation procedure and data visualisation technique  employed in our case study.
 
@@
 
-We processed the the data so that for both every origin and destination we have a unique identifier and the longitude and latitude.
-We have done this using the open-source reconciliator at \url{https://github.com/irail/stations}.
+We processed the the data so that for both every origin and destination we have a unique identifier and the geographic coordinates.
+This is facilitaed with the open-source reconciliator \todo{one liner that describes how this works} situated at \url{https://github.com/irail/stations}.
 

### 2016-01-04 paper.tex
merged

@@
 The iRail API is a XML/JSON HTTP API where each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
+
 It contains four features: 
+\begin{enumerate}
+\item \emph{Route Planning} (\url{http://api.irail.be/connections/{?from,to,date,time}}) plans a route from one station to another, taking into account a preferred start time.
+\item \emph{Next Departures} (\url{http://api.irail.be/liveboard/{?station,date,time}}) shows the next train departures in a certain station, useful for creating quick overviews.
+\item \emph{Trip Status} (\url{http://api.irail.be/vehicle/{?id,date}}) returns the stations this train passes by on the day of the request.
+\item \emph{List of all stations} (\url{http://api.irail.be/stations/}) gives an overview of all stations Belgian trains can arrive at.
+\end{enumerate}
 
-\begin{itemize}
-\item Route Planning (\url{http://api.irail.be/connections/{?from,to,date,time}}).
-\item Next Departures (\url{http://api.irail.be/liveboard/{?station,date,time}}).
-\item Train Status (\url{http://api.irail.be/vehicle/{?id,date}}).
-\item List All Stations (\url{http://api.irail.be/stations/}).
-\end{itemize}
-
-\todo{what are the intended functions of each feature?
-We should mention that the route planning feature is perhaps the most popular among lay users?
-Thus is also has good potential to be used as a proxy for analysis...}
-
+All except the Route Planning 
 We are particularly intersted in studying the potential application of route planning queries for flow analysis thus only requests to (\url{http://api.irail.be/connections/{?from,to,date,time}}) were extracted for further investigation.
@@
 When a \emph{rather low expressive} server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
-Again, we are not able to rely on the query logs.
+We would be able to fully rely on the query logs if the expressivity would be maximal (extreme right) and caching would be turned off.
+Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the Web, it will not scale up.\todo{PC: Just being devil's advocate here: is caching really necessary to scale sufficiently?}
 
@@
 
-The iRail project is a mediator: it is a not for profit organization to stimulate digital creativity concerning mobility.
-The interface that takes care of the route planning interface is open source and we suggested a change to the system in which the query logs become open\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
+IRail is a not for profit organization that stimulates digital creativity concerning mobility.
+The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
 As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
 
-\todo{When the expressiveness of the servers is lower in order to maximize the reuse of the data, such as in the case of Linked~Connections or a GTFS~datadump, we suggest to, instead of logging each GET-request, to have an analytics server that gathers all the logs.}
+In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
+We call this a \emph{JSON-stream}, as the JSON objects can be read one per one.
+Unix commands such as \emph{grep} work well with this format.
+We chose this over CSV as we have more flexibility towards changing the object properties.
 

### 2016-01-04 paper.tex

@@
 Ruben Verborgh\\
-\affaddr{\email{\texttt{ruben.verborg@ugent.be}}}
+\affaddr{\email{\texttt{ruben.verborgh@ugent.be}}}
 \and
@@
 
-All except the Route Planning 
-We are particularly intersted in studying the potential application of route planning queries for flow analysis thus only requests to (\url{http://api.irail.be/connections/{?from,to,date,time}}) were extracted for further investigation.
+We are particularly intersted in studying the potential application of route planning queries for flow analysis.
+The other queries are commonly used by digital signage providers or status boards, which poll the data on a regular basis.
+We extracted the route planning requests (\url{http://api.irail.be/connections/{?from,to,date,time}}) for further investigation.
 The resulting dataset described in this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 
-This dataset is compiled from the \emph{Apache} \emph{access.log} files generated between 2012 and 2015. Each data entry contains a: 
-\begin{itemize}
-  \item Timestamp,
-  \item HTTP Status Code,
-  \item Request Path (e.g., \url{/connections/?from=Ghent\&to=Antwerp})
-  \item User-Agent (cfr.
+This dataset is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015.
+Each data entry contains a timestamp of when the request was executed, the HTTP status code of the response, the request path (e.g., \url{/connections/?from=Ghent\&to=Antwerp}) and the User-Agent (cfr.
 RFC2616\footnote{The User-Agent request-header field contains information about the user agent originating the request.
-This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations -- \url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}})
-\end{itemize}
+This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations -- \url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}).
 
 As we are dealing with query logs, privacy becomes an issue~\cite{silvestri}.
-However, we do not publish the IP addresses with each query, which makes it hard to track single users.
+However, we do not publish or process the IP addresses with each query, which makes it hard to track single users.
 To remove all doubt, we have asked and received the consent of the Belgian privacy commission, which declared we can study the logs without any restrictions.
@@
 \includegraphics[width=8.6cm]{querylogs}
-\caption{Amount of queries on the route planning part of the iRail API between 2012 and 2016.
+\caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
 In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
@@
 \includegraphics[width=8.6cm]{avg-all}
-\caption{Distribution of iRail route planning API queries on average per day of the week between 2012 and 2016}
+\caption{Distribution of iRail route planning API queries on average per day of the week between October 2012 and December 2015}
 \label{fig:average}
@@
 \includegraphics[width=8.6cm]{dec24}
-\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between 2012 and 2016.
+\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
 The illustration shows that people started going home earlier than usual.}

### 2016-01-04 paper.tex

@@
 % Perspectives
-High expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
-In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
+Nevertheless, we should be cautious about what we measure as GET requests on the Web are cacheable.
+In order to have a query log for studying travel demand that works for all publishing strategies, we suggest future work to send anonymized POST requests from the end-user applications to an analytics server.
+%High expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
+%In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
 

### 2016-01-05 paper.tex
Modified method section

--- 2016-01-04 paper.tex
@@
 
-The iRail API is a XML/JSON HTTP API where each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
-
-It contains four features: 
+The iRail API is a XML/JSON HTTP API where each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app. It contains four features: 
 \begin{enumerate}
@@
 
-We are particularly intersted in studying the potential application of route planning queries for flow analysis.
-The other queries are commonly used by digital signage providers or status boards, which poll the data on a regular basis.
+We are particularly intersted in studying the potential application of route planning queries for flow analysis as this feature is of greatest relevance to human users.
+The other features are commonly used by digital signage providers or status boards, which poll the data on a regular basis.
 We extracted the route planning requests (\url{http://api.irail.be/connections/{?from,to,date,time}}) for further investigation.
@@
 This dataset is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015. 
-Each data entry contains a timestamp of when the request was executed, the HTTP status code of the response, the request path (e.g., \url{/connections/?from=Ghent\&to=Antwerp}) and the User-Agent (cfr.
+Each data entry contains a unique identifier for database management, a timestamp of the request  executed, the HTTP status code of the response, the request path (e.g., \url{/connections/?from=Ghent\&to=Antwerp}) and the User-Agent (cfr.
 RFC2616\footnote{The User-Agent request-header field contains information about the user agent originating the request.
@@
 
-As we are dealing with query logs, privacy becomes an issue~\cite{silvestri}.
-However, we do not publish or process the IP addresses with each query, which makes it hard to track single users.
-To remove all doubt, we have asked and received the consent of the Belgian privacy commission, which declared we can study the logs without any restrictions.
+Privacy preservation is paramount~\cite{silvestri} when dealing with query logs.
+To protect privacy of our users, IP addresses have been removed from the queries.
+Additionally, the Belgian privacy commission has granted us the right to study these query logs without any restrictions.
 
@@
 Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
-Data visualisation is typically employed to facilate in this process thus the input data must be transformed into a format that expresses the quantity and directional flow of human movment.
+In most cases, the analytical objective is to identify trends that occur in journeys made  between pairs of locations and allow for their sensitivites to be explored. 
+Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded.
 In this section we describe the data transformation procedure and data visualisation technique employed in our case study.
@@
 
-We processed the the data so that for both every origin and destination we have a unique identifier and the geographic coordinates.
+We processed the the data so that data records with matching pairs of origin and destination (OD) station names are aggregated. To do so, several technical
+
  This is facilitaed with the open-source reconciliator \todo{one liner that describes how this works} situated at \url{https://github.com/irail/stations}.
+
+To address this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and correct errors.
 

### 2016-01-05 paper.tex
Modified Section 4.1

--- 2016-01-05 paper.tex
@@
 
-We processed the the data so that data records with matching pairs of origin and destination (OD) station names are aggregated. To do so, several technical
+The data is formatted in a manner where data records with matching pairs of origin and destination (OD) station names are aggregated. 
+To arrive at this outcome several technical challenges must be addressed.
+One distinctive challenge is the substantial amount of inconsistency in the manner where origin and destination (OD) station names are defined. 
+Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of aggregation.
+To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and correct identifiable errors.
 
- This is facilitaed with the open-source reconciliator \todo{one liner that describes how this works} situated at \url{https://github.com/irail/stations}.
-
-To address this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and correct errors.
-
-Each data record now contains the station object of both origin and destination (OD), the user agent that made the query, as well as the time when the query was made.
 As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of \todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 06:00~hrs and 10:00~hrs as well as 17:00~hrs and 21:00~hrs on weekdays, and all data records created on weekend.

### 2016-01-05 paper.tex
Section 4.1 Complete (figures not ready yet)

--- 2016-01-05 paper.tex
@@
 
-We processed the the data so that data records with matching pairs of origin and destination (OD) station names are aggregated. To do so, several technical
+Matching pairs of origin and destination (OD) stations must be aggregated prior to visualisation.
+Several stages of data processing are required to arrive at this outcome.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
+\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
+Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
+Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
+One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
+Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualisation.
+To overcome this challenge, an open-source reconciliator (https:// github.com/ irail/ stations) was employed to reduce spelling variations and correct identifiable errors.
 
- This is facilitaed with the open-source reconciliator \todo{one liner that describes how this works} situated at \url{https://github.com/irail/stations}.
-
-To address this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and correct errors.
-
-Each data record now contains the station object of both origin and destination (OD), the user agent that made the query, as well as the time when the query was made.
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of \todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 06:00~hrs and 10:00~hrs as well as 17:00~hrs and 21:00~hrs on weekdays, and all data records created on weekend.
-Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations.
-Similarly, data records with unparsable origin or destination station names due to spelling mistakes or invalid character encoding were removed as well.
-
-The data is then simplified so that province level flow patterns are emphasised.
+The data is then aggregated so that province level flow patterns are emphasised.
 \cref{fig:brussels} provides a diagrammatic representation of the procedure.
-First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See Figure xa).
-If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (See Figure xb).
-Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (See Figure xc).
+First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See \cref{fig:brussels}a).
+If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (See \cref{fig:brussels}b).
+Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (See \cref{fig:brussels}c).
 The following types of flows are observable from the outcome of aggregation:

### 2016-01-05 paper.tex
Moved existing text on "Visualisation" to Related Work
Shifted text back to "Visualisation"

@@
 Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
-In most cases, the analytical objective is to identify trends that occur in journeys made  between pairs of locations and allow for their sensitivites to be explored. 
+In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivites to be explored. 
 Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded.
@@
 
-\todo{Alvin}
+Visualization is crucial for making the analysis tangible, so that results can be communicated and debated \todo{Robinson, 2008 -121}. 
+This is key for our research since the questions presented are exploratory in nature \todo{Kraak, 2008 -150} and can be addressed in many ways. 
+Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \todo{Andrienko, 2012 -196}. 
+Of these techniques, flow visualisations are commonly employed to facilitate the comparison of aggregated movement. 
+Flow visualisations can be broadly categorized into two groups.
+Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
+Since our data solely consists of OD information, we narrow our description to the latter group of visualisation techniques.
 
-Visualisation is frequently used to make data analysis tangible, so that the results can be communicated and debated \todo{Robinson, 2008 -121}.
-This is crucial for our research since the questions presented are exploratory in nature \todo{Kraak, 2008 -150} and can be addressed in many ways.
-Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \todo{Andrienko, 2012 -196}.
-Of these techniques, flow visualisations facilitate the comparison of aggregated movement over space and time.
-There are three types of flow visualisations: Line based representations preserves the complete trajectory while matrix type representations take only the start and end locations into account.
-
-Line based representations indicate movement on a path with the shape of the line.
-The thickness of each line is scaled to the volume of flow along a path.
-The direction of flow is commonly indicated with an arrowhead.
-This representation is detailed and straightforward to understand but becomes cluttered when lines intersect.
-A number of solutions have been proposed for clutter reduction.
-Filtering is the most commonly practiced.
-The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \todo{Andrienko, 2007 -198}.
-Hierarchical \todo{Guo, 2009 -199} and density based \todo{Rinzivillo, 2008 -201} clustering have also been introduced for this purpose.
-Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualisation.
-Another solution to reduce clutter is to transform the visual representation.
-This is achieved by rerouting \todo{Phan, 2005 -88} or bundling \todo{Hurter, 2014 -200} lines in close proximity so that they appear grouped.
-Unlike clustering, this solution changes the shape of lines by way of interpolation to obtain an ideal layout.
-In most cases, the interpolation is regulated by conditions to avoid intersections and follow key geographic features such as roads, rivers or coastlines so that the layout appears natural.
-This solution emphasises major flows within the visualisation since large groups gain visual dominance.
-Nonetheless, there are two drawbacks to this solution.
-First, short distance flows that may be important at local scales are de-emphasised or lost in the transformation.
-Next, the lines do not depict actual paths of movement thus this solution may not be suitable for applications that require strict geographic accuracy.
-
-Matrix type representations are frequently used when the origin and destination preside over intermediate locations on the path of movement.
-This type of representation is referred to as origin-destination (OD) matrix.
+OD type representations are frequently used when the origin and destination preside over intermediate locations on the path of movement. 
+The OD Matrix is most common visualization technique in this type of representation. 
 Rows and columns correspond to locations while cells are coloured to express the volume of flow.
-OD matrices can be reordered to emphasise connectivity between locations but the lack of geographic context is a distinct disadvantage.
-Several solutions have been developed to address this shortcoming.
-These are based on the notion of small multiples: The matrix is arranged in a geographical order so that the cells correspond to locations on the map.
-Then, a smaller map of the geographical context is nested in each cell to enable more intuitive comparisons \todo{Guo, 2006 -202}.
-The limitation of this solution is similar to other forms or small multiples in general - the layout space allocated to each nested map becomes smaller as the number of cells increase.
-In this instance, nesting detailed geographical maps is problematic since majority of the geographical features become illegible due to scaling.
-OD Map \todo{Wood, 2010 -86} proposes to nests a series of geographically ordered matrices that are more expressive of spatial relations and remain legible even when a small amount of layout space allocated to each cell.
 

### 2016-01-05 paper.tex
Description of "Node-Link" representations

@@
 Rows and columns correspond to locations while cells are coloured to express the volume of flow.
+OD matrices can be reordered to emphasise connectivity between locations.
+While this feature is particularly useful for identifying trends in the data, the tabular display is challenging for lay users to interpret and  less expressive of connections between locations than representations based on the node-link metaphor.
 

### 2016-01-05 paper.tex
Description of chord diagram completed

--- 2016-01-05 paper.tex
@@
 
-OD type representations are frequently used when the origin and destination preside over intermediate locations on the path of movement. 
+OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
 The OD Matrix is most common visualization technique in this type of representation. 
 Rows and columns correspond to locations while cells are coloured to express the volume of flow.
+OD matrices can be reordered to emphasise connectivity between locations.
+While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
+In recent years, such representations have become increasingly prominent for  displaying movement in scientific publications as well as popular culture.
+Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
+
+The chord diagram is a visualisation technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
+Each node is represented by an arc where its length is proportional to the total volume of incoming and outging flows. 
+Chords or curved line segements are drawn to connect nodes. 
+The width at the head or tail of each chord indicates the amount of movement that occured between a certain location to another at the other end.
 

### 2016-01-05 paper.tex
Added chord diagram interaction & colour coding details

@@
 Chords or curved line segements are drawn to connect nodes. 
-The width at the head or tail of each chord indicates the amount of movement that occured between a certain location to another at the other end.
+The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
+In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
+Interactive filtering is introduced to simplify the visualisation and details are provided in popup dialogue when upon mouse-over.
 

### 2016-01-05 paper.tex
Section 3

--- 2016-01-05 paper.tex
@@
 
+Visualisation is frequently used to make data analysis tangible, so that the results can be communicated and debated \todo{Robinson, 2008 -121}.
+This is crucial for our research since the questions presented are exploratory in nature \todo{Kraak, 2008 -150} and can be addressed in many ways.
+Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \todo{Andrienko, 2012 -196}.
+Of these techniques, flow visualisations facilitate the comparison of aggregated movement over space and time.
+There are three types of flow visualisations: Line based representations preserves the complete trajectory while matrix type representations take only the start and end locations into account.
+
+Line based representations indicate movement on a path with the shape of the line.
+The thickness of each line is scaled to the volume of flow along a path.
+The direction of flow is commonly indicated with an arrowhead.
+This representation is detailed and straightforward to understand but becomes cluttered when lines intersect.
+A number of solutions have been proposed for clutter reduction.
+Filtering is the most commonly practiced.
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \todo{Andrienko, 2007 -198}.
+Hierarchical \todo{Guo, 2009 -199} and density based \todo{Rinzivillo, 2008 -201} clustering have also been introduced for this purpose.
+Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualisation.
+Another solution to reduce clutter is to transform the visual representation.
+This is achieved by rerouting \todo{Phan, 2005 -88} or bundling \todo{Hurter, 2014 -200} lines in close proximity so that they appear grouped.
+Unlike clustering, this solution changes the shape of lines by way of interpolation to obtain an ideal layout.
+In most cases, the interpolation is regulated by conditions to avoid intersections and follow key geographic features such as roads, rivers or coastlines so that the layout appears natural.
+This solution emphasises major flows within the visualisation since large groups gain visual dominance.
+Nonetheless, there are two drawbacks to this solution.
+First, short distance flows that may be important at local scales are de-emphasised or lost in the transformation.
+Next, the lines do not depict actual paths of movement thus this solution may not be suitable for applications that require strict geographic accuracy.
+
+Matrix type representations are frequently used when the origin and destination preside over intermediate locations on the path of movement.
+This type of representation is referred to as origin-destination (OD) matrix.
+Rows and columns correspond to locations while cells are coloured to express the volume of flow.
+OD matrices can be reordered to emphasise connectivity between locations but the lack of geographic context is a distinct disadvantage.
+Several solutions have been developed to address this shortcoming.
+These are based on the notion of small multiples: The matrix is arranged in a geographical order so that the cells correspond to locations on the map.
+Then, a smaller map of the geographical context is nested in each cell to enable more intuitive comparisons \todo{Guo, 2006 -202}.
+The limitation of this solution is similar to other forms or small multiples in general - the layout space allocated to each nested map becomes smaller as the number of cells increase.
+In this instance, nesting detailed geographical maps is problematic since majority of the geographical features become illegible due to scaling.
+OD Map \todo{Wood, 2010 -86} proposes to nests a series of geographically ordered matrices that are more expressive of spatial relations and remain legible even when a small amount of layout space allocated to each cell.
+
 \section{The query logs}
@@
 Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
-In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivites to be explored. 
+In most cases, the analytical objective is to identify trends that occur in journeys made  between pairs of locations and allow for their sensitivites to be explored. 
 Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded.
@@
 
-Visualization is crucial for making the analysis tangible, so that results can be communicated and debated \todo{Robinson, 2008 -121}. 
-This is key for our research since the questions presented are exploratory in nature \todo{Kraak, 2008 -150} and can be addressed in many ways. 
-Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \todo{Andrienko, 2012 -196}. 
-Of these techniques, flow visualisations are commonly employed to facilitate the comparison of aggregated movement. 
-Flow visualisations can be broadly categorized into two groups.
-Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
-Since our data solely consists of OD information, we narrow our description to the latter group of visualisation techniques.
-
-OD type representations are frequently used when the origin and destination preside over intermediate locations on the path of movement. 
-The OD Matrix is most common visualization technique in this type of representation. 
-Rows and columns correspond to locations while cells are coloured to express the volume of flow.
-OD matrices can be reordered to emphasise connectivity between locations.
-While this feature is particularly useful for identifying trends in the data, the tabular display is challenging for lay users to interpret and  less expressive of connections between locations than representations based on the node-link metaphor.
+\todo{Alvin}
 
@@
 GTFS is a compressed ZIP-file containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
-It is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
+It is supported among all current open-source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
 %Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
@@
 It succesfully enabled reuse for intermodal travel, engineers can rely on the data dumps even if the servers of the transit agency are offline and there is no limitation to the features that can be implemented.
-The logs are however lost.
+The public transit agency however has no access any longer to an indication of travel demand.
 
@@
 IRail is a not for profit organization that stimulates digital creativity concerning mobility.
-The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
+The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
 As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.

### 2016-01-05 paper.tex
Modified section 3, paragraph 1

@@
 
-We are particularly intersted in studying the potential application of route planning queries for flow analysis as this feature is of greatest relevance to human users.
-The other features are commonly used by digital signage providers or status boards, which poll the data on a regular basis.
-We extracted the route planning requests (\url{http://api.irail.be/connections/{?from,to,date,time}}) for further investigation.
-The resulting dataset described in this paper, can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+We are particularly interested in studying the potential application of route planning queries as an information resource for flow analysis due to the input it captures from human users.
+The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards.
+The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015.
+A copy of this dataset in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+Each data entry contains a unique identifier, a timestamp of the request executed, the HTTP status code of the response indicating a successful or failed operation, the request path between two stations and information of the user agent (cfr.
+RFC2616) including details like the type of device, protocol and ip address.
+User agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
+Automated recognition of user agents also enables our API to provide tailored responses certain services such as \todo{Pieter}.
+Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users.
 
-This dataset is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015. 
-Each data entry contains a unique identifier for database management, a timestamp of the request  executed, the HTTP status code of the response, the request path (e.g., \url{/connections/?from=Ghent\&to=Antwerp}) and the User-Agent (cfr.
-RFC2616\footnote{The User-Agent request-header field contains information about the user agent originating the request.
-This is for statistical purposes, the tracing of protocol violations, and automated recognition of user agents for the sake of tailoring responses to avoid particular user agent limitations -- \url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}).
+Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be utilised in scientific studies without further restrictions.
 
-Privacy preservation is paramount~\cite{silvestri} when dealing with query logs.
-To protect privacy of our users, IP addresses have been removed from the queries.
-Additionally, the Belgian privacy commission has granted us the right to study these query logs without any restrictions.
-
-We established three preliminary questions to  investigate the potential of this dataset:
+We established three research questions to investigate the potential of this dataset:
 \begin{enumerate}
@@
 The data is then aggregated so that province level flow patterns are emphasised.
+\todo{Alvin: Figure must be drawn}
 \cref{fig:brussels} provides a diagrammatic representation of the procedure.

### 2016-01-05 paper.tex

@@
 A copy of this dataset in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
-Each data entry contains a unique identifier, a timestamp of the request executed, the HTTP status code of the response indicating a successful or failed operation, the request path between two stations and information of the user agent (cfr.
+Each data entry contains a unique identifier, a timestamp of the request executed, the HTTP status code of the response indicating a successful or failed operation, the request path between two stations and information of the user-agent (cfr.
 RFC2616) including details like the type of device, protocol and ip address.
-User agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
-Automated recognition of user agents also enables our API to provide tailored responses certain services such as \todo{Pieter}.
+User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
+Automated recognition of user-agents also enables our API to provide tailored responses certain services such as \todo{Pieter}.
 Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
-
 Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be utilised in scientific studies without further restrictions.
-
-We established three research questions to investigate the potential of this dataset:
-\begin{enumerate}
-  \item How many queries were done monthly and how does this compare with the real number of passengers?
-  \item How are the requests distributed within a day?
-  \item Can we see deviations on an average request distribution on e.g., public holidays?
-\end{enumerate}
 
@@
 
-In \cref{fig:querylogs} we show the amount of queries per month.
-We can see a sudden grow in April 2015: this is when an official app of the Belgian railway company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}} and many people started looking for an alternative within the apps built on top of the iRail API.
+Results from preliminary analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
+Three observations in particular prompted further analysis of the data in greater detail. 
+First, there has been a steady increase in the number of queries made between 2012 and 2015. 
+\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
+A sharp rise in the amount of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
 In 2012, 2013, 2014 and 2015, there were respectively 19.3, 19.4, 19,6 and 19.7 million passengers per month on average\footnote{Calculated by using indicators of the Flemish government and extrapolating the numbers for 2015 \url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}.

### 2016-01-05 paper.tex

@@
 \cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
-A sharp rise in the amount of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
-In 2012, 2013, 2014 and 2015, there were respectively 19.3, 19.4, 19,6 and 19.7 million passengers per month on average\footnote{Calculated by using indicators of the Flemish government and extrapolating the numbers for 2015 \url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}.
-The amount of iRail queries per month, on average in these years are respectively 0.11, 0.17, 0.15 and 0.33 million.
-If we assume that each request is a travel intention, this would result in respectively $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the travels that are captured with the iRail API.
+A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
+Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
+Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
+Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
+This suggests that \todo{Pieter: what does this mean?
+The numbers appear rather small though}.
 
@@
 
-Next, we made a distribution of the requests done on each day of the week, which resulted in \cref{fig:average}.
-In this chart we can spot, on Monday until Friday, the morning peak between 5 and 9 and the evening peak between 15 and 18.
-We see a small dent during noon \todo{any explanation for this?}.
-On Fridays we see that the morning peak is stronger than the morning peak, which can be explained by many students that go home to visit their parents for the weekend \todo{Can we give a link to more reading material about this behaviour?
-This might be weird for US people?}.
-On Saturdays, we almost do not see the peaks, which can be explained by people not going to work on weekends.
-On Sundays, we see most requests happen in the evening, which can be explained by students returning to college accomodations.
+Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
+\cref{fig:average} depicts an break down of the queries that occur on an hourly basis for each day of the week.
+Several trends can be observed in this chart. 
+Morning and evening peak periods in particular, are clearly distinguishable.
+In this regard, the peak hours on weekdays are distinctively different from those of the weekends. 
+Similarly, the frequency of queries during the morning peak period on Fridays appear to be substantially lower than those that occur between Monday and Thursday.
+The discrepancy maybe explained by a decrease in the number students who travel home from college over the weekends.
+\todo{Can we give a link to more reading material about this behaviour?
+This might be weird for US people?
+--> not sure if this is really case.
+We need to explain this in a clear way}.
+The absence of a clear peak on Saturdays can be attributed to lack of journey to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
 
@@
 \end{figure}
-
-Finally, we compared the average distribution of Thursdays with the distribution on Thursday the 24th of December 2015 (Christmass Eve).
-Our hypothesis that we would now see an earlier evening peak can be confirmed by \cref{fig:dec24}.
-
-We conclude that the timing of the queries look interesting for flow analysis. 
-Can we also find interesting patterns when we study where people are heading between certain time intervals?
 

### 2016-01-05 paper.tex
Future work started

@@
 User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
-Automated recognition of user-agents also enables our API to provide tailored responses certain services such as \todo{Pieter}.
 Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
@@
 Unix commands such as \emph{grep} work well with this format.
-We chose this over CSV as we have more flexibility towards changing the object properties.
+We chose this over CSV as we have more flexibility towards changing the object properties in the future.
+Next, we chose to add an object to the stream which only contains an object with key \emph{@context}.
+This is the JSON-LD context of all the objects in the, JSON-stream.
+Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.
 
-\todo{Also in USEWOD2014 there was a paper on LDF and how we could know which SPARQL queries needed to be solved}
+%As a new version of the logs are put at \url{http://api.irail.be/logs/} only each second, we have added cache-headers to the document with max-age equal to 1 second.
+%\todo{Also in USEWOD2014 there was a paper on LDF and how we could know which SPARQL queries needed to be solved}
 
-\todo{As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-This amounts today to an average of 86k queries to the API per day, as the API gained popularity thanks to apps like \emph{BeTrains}\footnote{\url{https://play.google.com/store/apps/details?id=tof.cv.mpp}} and \emph{Railer}\footnote{\url{http://railer.be/}}.
-On average, 14,357 of these are valid route planning queries (other queries include requesting departures in a station, status of a vehicle, or a list of all stations).
-If each query would amount to indeed one passenger, this would today be equal to almost 2\% of all passengers.}
+\section{Future Work}
 
-\section{Conclusion and future work}
+Query logs are thus an indication of travel intentions. 
+Yet we can identify some hiatuses in query logs as travel intentions:
+\begin{enumerate}
+  \item Caching may mask peak hours, as \ldots
+  \item Multiple requests may be needed to cover a travel intention
+  \item \ldots
+\end{enumerate}
+
+In order to create a better representation of travel intentions, we would recommend apps to do a POST-request instead of a
+
+\section{Conclusion}
 \label{sec:conclusion}

### 2016-01-06 paper.tex
Modified section 3, paragraph 4

--- 2016-01-05 paper.tex
@@
 Morning and evening peak periods in particular, are clearly distinguishable.
-In this regard, the peak hours on weekdays are distinctively different from those of the weekends. 
+Peak hours on weekdays are also distinctively different from those of weekends.
 Similarly, the frequency of queries during the morning peak period on Fridays appear to be substantially lower than those that occur between Monday and Thursday.
@@
 The absence of a clear peak on Saturdays can be attributed to lack of journey to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
+
+Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
+Figure 3 provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
+As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
+For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.
 

### 2016-01-06 paper.tex
Missing line from section 3, paragraph 4
Added visualisation references

@@
 As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
-For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.
+For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis. \todo{where people are heading between certain time intervals?}
 
@@
 
-Visualization is crucial for making the analysis tangible, so that results can be communicated and debated \todo{Robinson, 2008 -121}. 
-This is key for our research since the questions presented are exploratory in nature \todo{Kraak, 2008 -150} and can be addressed in many ways. 
-Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \todo{Andrienko, 2012 -196}. 
+Visualization is crucial for making the analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
+This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
+Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
 Of these techniques, flow visualisations are commonly employed to facilitate the comparison of aggregated movement. 
@@
 Reading both diagrams in this manner reveals the complexity of movement on the rail system and the significance of cities in daily travel.
-Brussels serves as the principal centre of rail activity in general, yet its centrality is more distinctive for the walloon region than Flanders.
+Brussels serves as the principal centre of rail activity in general, yet its function as a centre is more distinctive for the walloon region than Flanders.
 Queries from Walloonia with exception to Liege, are generally made from provincial stations towards Brussels instead of their respective provincial capitals.

### 2016-01-06 paper.tex

--- 2016-01-05 paper.tex
@@
 The absence of a clear peak on Saturdays can be attributed to lack of journey to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
+
+Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
+Figure 3 provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
+As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
+For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis. \todo{where people are heading between certain time intervals?}
 

### 2016-01-06 paper.tex
Modified section 5.1, paragraph 1

--- 2016-01-06 paper.tex
@@
 This suggests that \todo{Pieter: what does this mean?
-The numbers appear rather small though}.
+The numbers appear rather small though.
+Ideally this should be reflected in figure 1}.
 
@@
 
-Visualization is crucial for making the analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
+Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
 This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
@@
 In this chapter, we report on the results of studying visualizations with experts of Trein Tram Bus, a not for profit passenger federation.
+\todo{Trein Tram Bus needs to be introduced in the introduction}
 
-\subsection{Commuting pattern}
+\subsection{Structure of flows in belgium}
 
-We see people leaving in the morning and return in the evening (see \cref{fig:brussels} and todo)
-
-The pair of chord diagrams in \cref{fig:brussels,fig:antwerp} capture the aggregated number of queries made between any two stations on weekdays and weekends between 6 to 9 in the morning.
-The chords that link two distinct stations, are coloured coded by region and proportionately scaled to the number of queries from a station of origin to a destination.
-For instance, approximately 3,530 queries were made from Leuven to Brussels than in the opposite direction (approximately 1,530) on weekdays.
-Reading both diagrams in this manner reveals the complexity of movement on the rail system and the significance of cities in daily travel.
-Brussels serves as the principal centre of rail activity in general, yet its function as a centre is more distinctive for the walloon region than Flanders.
-Queries from Walloonia with exception to Liege, are generally made from provincial stations towards Brussels instead of their respective provincial capitals.
-Queries in Flanders, on the hand, tend to be distributed among major cities otherwise known as the Flemish Diamond, a network of four metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
-The difference between both patterns appear to correspond with existing measures of population density, providing valuable insight and alternative perspectives into the function of cities in rural and urban settings.
-
-\subsection{Flanders is polycentric, Wallonia monocentric}
+The chord diagrams shown in \cref{fig:brussels} and \cref{fig:antwerp} depict the aggregated number of queries made between any two stations in the morning and evening respectively. 
+Visualising our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
+Brussels stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
+On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
+With exception to Liege, queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
+Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
+The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
 

### 2016-01-06 paper.tex
Shifted around chapters 6 and 7, extended conclusion with future work.

@@
 
-The iRail API is a XML/JSON HTTP API where each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app. It contains four features: 
+\subsection{Historical logs}
+
+The iRail API is an XML/JSON HTTP API: each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
+It contains four features: 
 \begin{enumerate}
@@
 \end{figure}
+
+\subsection{Real-time logs}
+\label{sec:realtimelogs}
+
+IRail is a not for profit organization that stimulates digital creativity concerning mobility.
+The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
+As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+
+In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
+The JSON objects can be read one per one, so unix commands such as \emph{grep} work well with this format.
+We chose this over CSV as we have more flexibility towards changing the object properties in the future.
+Next, we chose to add an object to the stream which only contains an object with key \emph{@context}.
+This is the JSON-LD context of all the objects in the, JSON-stream.
+Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.
+
+Using this data, everyone can build their own historic query log dataset.
+The server costs for iRail to host this dataset stay limited.
+Furthermore, this way of publishing data enables research for real-time predictions of travel intentions.
+For this paper, we will focus on using the historic logs.
+Later, the research can be reproduced with more data.
 
@@
 %%%
-The expressiveness of a server affects the way logs can be gathered.
+The expressiveness of a server affects the way logs can be gathered~\cite{usewod2015}.
 We have shown in previous sections that the query logs of a \emph{high expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
@@
 We would be able to fully rely on the query logs if the expressivity would be maximal (extreme right) and caching would be turned off.
-Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the Web, it will not scale up.\todo{PC: Just being devil's advocate here: is caching really necessary to scale sufficiently?}
+Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the open Web, caching helps decrease the CPU load on servers on  peak moments.
 
-\section{Publishing query logs}
-\label{sec:publishingquerylogs}
-
-IRail is a not for profit organization that stimulates digital creativity concerning mobility.
-The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
-As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-
-In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
-We call this a \emph{JSON-stream}, as the JSON objects can be read one per one.
-Unix commands such as \emph{grep} work well with this format.
-We chose this over CSV as we have more flexibility towards changing the object properties in the future.
-Next, we chose to add an object to the stream which only contains an object with key \emph{@context}.
-This is the JSON-LD context of all the objects in the, JSON-stream.
-Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.
-
-%As a new version of the logs are put at \url{http://api.irail.be/logs/} only each second, we have added cache-headers to the document with max-age equal to 1 second.
-%\todo{Also in USEWOD2014 there was a paper on LDF and how we could know which SPARQL queries needed to be solved}
-
-\section{Future Work}
-
-Query logs are thus an indication of travel intentions. 
-Yet we can identify some hiatuses in query logs as travel intentions:
-\begin{enumerate}
-  \item Caching may mask peak hours, as \ldots
-  \item Multiple requests may be needed to cover a travel intention
-  \item \ldots
-\end{enumerate}
-
-In order to create a better representation of travel intentions, we would recommend apps to do a POST-request instead of a
-
-\section{Conclusion}
+\section{Conclusion and future work}
 \label{sec:conclusion}
 
-The contribution of this paper is twofold.
-For the first time, we studied query logs to find travel patterns.
-Due to lack of information we could not evaluate our results, yet the visualizations look promising...
-There are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
-For instance, the data only captures an intention of a passenger: it's unsure whether the person actually took the train.
-Furthermore, different requests could be done by one person with an intention to travel.
-On top of that, our sample used for this paper can only represent a maximum of 0.4\% of the daily traffic flow.
+First we studied query logs to find travel patterns in Belgium and can conclude this is up to now, the best representation of travel flows over the railway network in Belgium.
+%, as no alternatives are available.
+We studied data which represent a maximum of 1.66\% of the passengers.
+\todo{ALVIN: Visualizing our dataset reveals the complexity of movement on the Belgian rail system and the significance of cities in weekday travel.
+Furthermore, we found evidence that Flanders is polycentric, while Walloon traffic converges on Brussels.}
 
-The second contribution is our position that we suggest to decouple the query logs from the query execution.
+Next, we also concluded that there are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
+We identified a couple of gaps:
+\begin{enumerate}
+  \item The data only captures an intention of a passenger: it's unsure whether the person actually took the train.
+  \item Multiple user queries may be needed to cover a travel intention.
+  \item Caching may mask peak hours, as many of the same requests in one minute is only stored in the logs once.
+\end{enumerate}
+Nevertheless, we published the query logs as open data at \url{http://api.irail.be/logs} for other scientists to continue to research possibilities with this dataset.
+For instance, predicting trip congestion on the basis of this data, looks promising. %the iRail project is planning to use this data to predict trip congestion on the basis of this data.
 
-In future work, iRail will try to use these to predict train congestion.
+Related work~\cite{verborgh2014lonesome} suggested to add additional metadata in the HTTP headers in order to track the user's intention.
+While this is an interesting suggestion, it does not overcome the problem of caching.
+In order to create a better representation of travel intentions, future work can gather logs by POST-requests with the only purpose to gather analytics of travel intention.
+This is a similar approach than what happens on the Web of Documents today with \emph{Piwik} or \emph{Google Analytics}.
+
 % Fix spacing after References header (as line 1308 of sig-alternate.cls breaks it)

### 2016-01-06 paper.tex
Fixes in Section 3.1

@@
 Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
-This suggests that \todo{Pieter: what does this mean?
-The numbers appear rather small though.
-Ideally this should be reflected in figure 1}.
 
@@
 Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
-\cref{fig:average} depicts an break down of the queries that occur on an hourly basis for each day of the week.
+\cref{fig:average} depicts a break down of the queries that occur, on average, on an hourly basis for each day of the week.
 Several trends can be observed in this chart. 
 Morning and evening peak periods in particular, are clearly distinguishable.
+At noon, a small dent is noticable, which may be attributed to part time workers and meetings on location after or before noon.
 Peak hours on weekdays are also distinctively different from those of weekends.
-Similarly, the frequency of queries during the morning peak period on Fridays appear to be substantially lower than those that occur between Monday and Thursday.
-The discrepancy maybe explained by a decrease in the number students who travel home from college over the weekends.
-\todo{Can we give a link to more reading material about this behaviour?
-This might be weird for US people?
---> not sure if this is really case.
-We need to explain this in a clear way}.
-The absence of a clear peak on Saturdays can be attributed to lack of journey to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
-
-Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
-Figure 3 provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
-As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
-For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis. \todo{where people are heading between certain time intervals?}
+Similarly, the frequency of queries during the evening peak period on Fridays appear to be substantially higher than those that occur between Monday and Thursday.
+The discrepancy maybe explained by students that travel home from their student homes over the weekends. 
+The absence of a clear peak on Saturdays can be attributed to lack of journeys to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
 
@@
 \end{figure}
+
+Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
+\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
+As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
+For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
 
@@
 \todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
-Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
+Queries made by automated user agents like search enginebots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
 Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
@@
 Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualisation.
-To overcome this challenge, an open-source reconciliator (https:// github.com/ irail/ stations) was employed to reduce spelling variations and correct identifiable errors.
+To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
 

### 2016-01-06 paper.tex
revert space removal

@@
 \todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
-Queries made by automated user agents like search enginebots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
+Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
 Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
@@
 
-\subsection{Structure of flows in belgium}
+\subsection{Structure of flows in Belgium}
 

### 2016-01-06 paper.tex
Modified section 5.1

@@
 Visualising our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
+Turning our attention to morning commutes, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
+The reverse occurs in the evening when trips are made towards provincial stations.
 Brussels stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
@@
 Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
+These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with   Brussels as a major centre.
 The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
@@
 \end{figure}
-
-We can also observe the effect of "density" in Flanders where people move from the rural towards cities.
-Look at weekday-mornings: commuting towards Gent, Brussels, Antwerp.
-This is not as obvious in the Walloon region (except for Liege).
-The reverse occurs on weekday-evenings.
-Local patterns can also be observed and this corresponds to what we know: ``cities are the central hubs''.
-Walloon region: everything converges on Brussels
 
@@
 We would be able to fully rely on the query logs if the expressivity would be maximal (extreme right) and caching would be turned off. 
-Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the open Web, caching helps decrease the CPU load on servers on  peak moments.
+Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the open Web, caching helps decrease the CPU load on servers at peak moments.
 

### 2016-01-06 paper.tex

@@
 Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded.
-In this section we describe the data transformation procedure and data visualisation technique employed in our case study.
+In this section we describe the data transformation procedure and data visualization technique employed in our case study.
 
@@
 
-Matching pairs of origin and destination (OD) stations must be aggregated prior to visualisation.
+Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
 Several stages of data processing are required to arrive at this outcome.
@@
 One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
-Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualisation.
+Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
 To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
@@
 This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
-Movement data has received substantial attention from the visualisation community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
-Of these techniques, flow visualisations are commonly employed to facilitate the comparison of aggregated movement. 
-Flow visualisations can be broadly categorized into two groups.
+Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
+Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
+Flow visualizations can be broadly categorized into two groups.
 Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
-Since our data solely consists of OD information, we narrow our description to the latter group of visualisation techniques.
+Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
 
@@
 
-The chord diagram is a visualisation technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
+The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
 Each node is represented by an arc where its length is proportional to the total volume of incoming and outging flows. 
@@
 In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
-Interactive filtering is introduced to simplify the visualisation and details are provided in popup dialogue when upon mouse-over.
+Interactive filtering is introduced to simplify the visualization and details are provided in popup dialogue when upon mouse-over.
 
@@
 The chord diagrams shown in \cref{fig:brussels} and \cref{fig:antwerp} depict the aggregated number of queries made between any two stations in the morning and evening respectively. 
-Visualising our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
+Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
 Turning our attention to morning commutes, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
@@
 
-We also see weird things happening in Antwerp, where a large amount of people go to the province in the morning on weekdays, while the oposite happens in Liege.
-When looking into the data, the Antwerp pattern appears due to one specific route which is queried each morning.
-Is there a group of colleagues all using the same app taking the same train daily?
-Is it someone who checks this trip over and over again for delays?
-We can only guess\ldots
+There are several inconsistent patterns in the data that do not correspond with existing understanding of flows in Belgium.
+In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
+Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API or a high concentration people who utilise route planning applications on a regular basis.
+Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
 

### 2016-01-06 paper.tex
Added figures in full pages
Review abstract.

@@
 \newcommand{\todo}[1]{\noindent\textcolor{red}{{\bf \{TODO}: #1{\bf \}}}}
+\newcommand{\rv}[1]{\noindent\textcolor{red}{{\bf \{RV}: #1{\bf \}}}}
 \newdef{definition}{Definition}
@@
 % Task
-We analyzed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
+In order to get an indication of travel flows between cities in Belgium,
+we analyzed the query logs of the iRail API, a highly expressive route planning API for the Belgian railways.
 We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
 % Object & Findings
-Using data visualizations, we illustrated the commuting patterns in Belgium and show that Brussels, the capital, acts as a central hub.
+Using data visualizations, we illustrate the commuting patterns in Belgium and confirm that Brussels, the capital, acts as a central hub.
 The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
-The findings correspond to the real travel demand, say experts of the passenger federation Trein Tram Bus.
+The findings correspond to the real travel demand, according to experts of the passenger federation Trein Tram Bus.
 % Conclusion
-We conclude that query logs of route planners are of high importance in getting an indication of the travel flows.
+We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
 % Perspectives
-High expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
+\rv{These last sentences don't fit well. They could be removed.}
+Highly expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
 In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
@@
 
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=8.1cm]{brussels}
+\includegraphics[width=15cm]{brussels}
 \caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.}
 \label{fig:brussels}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \centering
-\includegraphics[width=8.1cm]{antwerp}
+\includegraphics[width=15cm]{antwerp}
 \caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.}
 \label{fig:antwerp}
-\end{figure}
+\end{figure*}
 

### 2016-01-06 paper.tex
Review Section 3.

@@
 \subsection{Historical logs}
-
 The iRail API is an XML/JSON HTTP API: each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
-It contains four features: 
+It exposes four features: 
 \begin{enumerate}
@@
 We are particularly interested in studying the potential application of route planning queries as an information resource for flow analysis due to the input it captures from human users.
-The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards.
+The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards,
+and thus do not reflect direct human demands.
 The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015.
+\rv{Perhaps say \emph{we} made this copy available?}
 A copy of this dataset in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+\rv{I would set the following in a list, preferably one line per thing, but an inline enumeration would also work.}
 Each data entry contains a unique identifier, a timestamp of the request executed, the HTTP status code of the response indicating a successful or failed operation, the request path between two stations and information of the user-agent (cfr.
-RFC2616) including details like the type of device, protocol and ip address.
+RFC2616) including details like the type of device, protocol and IP address.
 User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
@@
 
+\rv{Here, you start with results. This seems to be a new section---I would recommend to start one.}
 Results from preliminary analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
@@
 First, there has been a steady increase in the number of queries made between 2012 and 2015. 
-\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
+\cref{fig:querylogs} provides a breakdown of the numbers on a monthly basis. 
 A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
@@
 \includegraphics[width=8.6cm]{dec24}
-\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
+\caption{Comparison of route planning queries on December 24th (Christmas Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
 The illustration shows that people started going home earlier than usual.}
@@
 Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
-\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
-As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
+\cref{fig:dec24} provides a breakdown of queries per hour on 24th December 2015 (Christmas Eve, on which Belgian families typically gather) in comparison to an average Thursday of the same year. 
+As illustrated, the evening peak on Christmas Eve occurs earlier than that of an average Thursday. 
 For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
@@
 
-IRail is a not for profit organization that stimulates digital creativity concerning mobility.
+iRail is a not for profit organization that stimulates digital creativity concerning mobility.
 The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
@@
 In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
-The JSON objects can be read one per one, so unix commands such as \emph{grep} work well with this format.
+The JSON objects can be read one per one, so UNIX commands such as \emph{grep} work well with this format.
 We chose this over CSV as we have more flexibility towards changing the object properties in the future.

### 2016-01-06 paper.tex
Review Section 4.

@@
 Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
-In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivites to be explored. 
-Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded.
+In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivities to be explored. 
+Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
 In this section we describe the data transformation procedure and data visualization technique employed in our case study.
@@
 Several stages of data processing are required to arrive at this outcome.
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays,
+\rv{why? and is it just you, or also the reader?}
+we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
 \todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
-Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
+Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may overrepresent connections between certain stations. 
+\rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
 Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
 One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
-Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
+Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
 To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
@@
 Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
+\rv{OD, I'm assuming origin--destination, or something else? Is it necessary to use the abbreviation the whole time?}
 Flow visualizations can be broadly categorized into two groups.
@@
 
+\rv{Starting here, this section isn't really going anywhere, at least not to a point. Consider removing the next two paragraphs and start talking about your data.}
 OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
@@
 
-The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
-Each node is represented by an arc where its length is proportional to the total volume of incoming and outging flows. 
-Chords or curved line segements are drawn to connect nodes. 
+The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
+Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 
+Chords or curved line segments are drawn to connect nodes. 
 The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
 In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
-Interactive filtering is introduced to simplify the visualization and details are provided in popup dialogue when upon mouse-over.
+Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.
 

### 2016-01-06 paper.tex
Review Section 5.

@@
 
-There are several inconsistent patterns in the data that do not correspond with existing understanding of flows in Belgium.
+There are several inconsistent patterns in the data that do not correspond with an existing understanding of flows in Belgium.
 In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
 Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API or a high concentration people who utilise route planning applications on a regular basis.
+\rv{Earlier, you said you filtered bots. Is this still the case?}
 Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
@@
 The expressiveness of a server affects the way logs can be gathered~\cite{usewod2015}.
-We have shown in previous sections that the query logs of a \emph{high expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
+We have shown in previous sections that the query logs of a \emph{highly expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
 The possibility that two requests are exactly the same is low, as there are many URLs which can be requested.

### 2016-01-06 paper.tex
Review Section 6.

@@
 \begin{enumerate}
-  \item When an application developer would like a \emph{new feature}, such as taking wheelchair accessibility information into account, the feature would have to be implemented on the server of the data publisher
-  \item Keeping the server \emph{high available} is costly, as any question can be asked by anyone for any purpose.
-  \item Federated querying, which would allow for \emph{intermodal} route planning for route planning APIs, is unexistent up to date.
+  \item When an application developer would like a \emph{new feature}, such as taking wheelchair accessibility information into account, the feature would have to be implemented on the server of the data publisher.
+  \item Keeping the server \emph{highly available} is costly, as any question can be asked by anyone for any purpose.
+  \item Federated querying, which would allow for \emph{intermodal} route planning for route planning APIs, is inexistent up to date.
 \end{enumerate}
@@
 In \cref{fig:LDFAxis2} we illustrate these two options as two extremes, with other options that are yet to be discovered.
+\rv{\emph{low expressive} is not correct, but I don't know what the right term is. (It is also high\emph{ly} expressive, I corrected that.)}
 When a \emph{rather low expressive} server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
 We would be able to fully rely on the query logs if the expressivity would be maximal (extreme right) and caching would be turned off. 
-Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the open Web, caching helps decrease the CPU load on servers at peak moments.
+Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the open Web, caching helps decrease load on servers at peak moments.
 

### 2016-01-06 paper.tex
Review Section 7.

@@
 While this is an interesting suggestion, it does not overcome the problem of caching.
-In order to create a better representation of travel intentions, future work can gather logs by POST-requests with the only purpose to gather analytics of travel intention.
+In order to create a better representation of travel intentions, future work can gather logs by POST requests with the only purpose to gather analytics of travel intention.
 This is a similar approach than what happens on the Web of Documents today with \emph{Piwik} or \emph{Google Analytics}.

### 2016-01-06 paper.tex
restructured the text

--- 2016-01-06 paper.tex
@@
 \newcommand{\todo}[1]{\noindent\textcolor{red}{{\bf \{TODO}: #1{\bf \}}}}
-\newcommand{\rv}[1]{\noindent\textcolor{red}{{\bf \{RV}: #1{\bf \}}}}
 \newdef{definition}{Definition}
@@
 % Task
-In order to get an indication of travel flows between cities in Belgium,
-we analyzed the query logs of the iRail API, a highly expressive route planning API for the Belgian railways.
+We analyzed the query logs of the iRail API, a high expressive route planning API for the Belgian railways, to get an indication of travel flows between cities in Belgium.
 We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
 % Object & Findings
-Using data visualizations, we illustrate the commuting patterns in Belgium and confirm that Brussels, the capital, acts as a central hub.
+Using data visualizations, we illustrated the commuting patterns in Belgium and show that Brussels, the capital, acts as a central hub.
 The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
-The findings correspond to the real travel demand, according to experts of the passenger federation Trein Tram Bus.
+The findings correspond to the real travel demand, say experts of the passenger federation Trein Tram Bus.
 % Conclusion
-We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
+We conclude that query logs of route planners are of high importance in getting an indication of the travel flows.
 % Perspectives
-\rv{These last sentences don't fit well. They could be removed.}
-Highly expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
+High expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
 In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
@@
 
-\subsection{Historical logs}
+%\subsection{Historical logs}
+
 The iRail API is an XML/JSON HTTP API: each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
-It exposes four features: 
+It contains four features: 
 \begin{enumerate}
@@
 
+%IRail is a not for profit organization that stimulates digital creativity concerning mobility.
+This interface is open source and we were able to suggest changes to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
+As of the 17th of December 2015, iRail publishes the most recent 1,000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
+In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
+The JSON objects can be read one per one, so unix commands such as \emph{grep} work well with this format.
+We chose this over CSV as we have more flexibility towards changing the object properties in the future.
+Next, we chose to add an HTTP header with a link to the query log context according to the JSON-LD specification\footnote{\url{}} to turn it into Linked Data.
+%This is the JSON-LD context of all the objects in the, JSON-stream.
+Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.
+
+Using this data, everyone can build their own historic query log dataset.
+The server costs for iRail to host this dataset stay limited.
+Furthermore, this way of publishing data enables research for real-time predictions of travel intentions.
+Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
+Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be published as open data.
+
 We are particularly interested in studying the potential application of route planning queries as an information resource for flow analysis due to the input it captures from human users.
-The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards,
-and thus do not reflect direct human demands.
-The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015.
-\rv{Perhaps say \emph{we} made this copy available?}
+The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards.
+The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
 A copy of this dataset in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
-\rv{I would set the following in a list, preferably one line per thing, but an inline enumeration would also work.}
-Each data entry contains a unique identifier, a timestamp of the request executed, the HTTP status code of the response indicating a successful or failed operation, the request path between two stations and information of the user-agent (cfr.
-RFC2616) including details like the type of device, protocol and IP address.
-User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
-Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
-Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be utilised in scientific studies without further restrictions.
+%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
+Each row within this dataset contains a timestamp of when the request was executed, the request path between two stations with longitudes and latitudes, the user-agent string cfr.
+RFC2616, the operating system of the user-agent and whether or not the user-agent is a bot.
+%After analysis, it appeared 10\% of the request where performed by Google bots.
+%User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
+%\todo{add glue to Method section?}
+
+\section{Method}
+\label{sec:method}
+
+Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
+In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivites to be explored. 
+Data visualization is typically employed to facilate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movment to be visually encoded.
+In this section we describe the data transformation procedure and data visualization technique employed in our case study.
+
+\subsection{Processing}
+
+Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
+Several stages of data processing are required to arrive at this outcome.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
+\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?} 
+Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
+Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
+One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
+Spelling mistakes as well as the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
+To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
+
+The data is then aggregated so that province level flow patterns are emphasized.
+\todo{Alvin: Figure must be drawn}
+\cref{fig:brussels} provides a diagrammatic representation of the procedure.
+First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See \cref{fig:brussels}a).
+If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (See \cref{fig:brussels}b).
+Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (See \cref{fig:brussels}c).
+The following types of flows are observable from the outcome of aggregation:
+
+\begin{itemize}
+  \item Travel from any provincial station to a major city.
+  \item Travel between any two major cities.
+  \item Travel between any two provincial stations.
+  \item Travel between any international station to a major city.
+  \item Travel between any international station to a provincial station.
+\end{itemize}
+
+\subsection{Visualization}
+
+Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
+This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
+Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
+Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
+Flow visualizations can be broadly categorized into two groups.
+Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
+Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
+
+OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
+The OD Matrix is most common visualization technique in this type of representation. 
+Rows and columns correspond to locations while cells are coloured to express the volume of flow.
+OD matrices can be reordered to emphasise connectivity between locations.
+While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
+In recent years, such representations have become increasingly prominent for  displaying movement in scientific publications as well as popular culture.
+Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
+
+The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a cirle. 
+Each node is represented by an arc where its length is proportional to the total volume of incoming and outging flows. 
+Chords or curved line segements are drawn to connect nodes. 
+The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
+In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
+Interactive filtering is introduced to simplify the visualization and details are provided in popup dialogue when upon mouse-over.
+
+\section{Results}
+\label{sec:results}
+
+In this chapter, we report on the results of studying visualizations with experts of Trein Tram Bus, a not for profit passenger federation.
+\todo{Trein Tram Bus needs to be introduced in the introduction}
+
+\subsection{Time based analysis}
 
@@
 
-\rv{Here, you start with results. This seems to be a new section---I would recommend to start one.}
-Results from preliminary analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
+Results from the analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
 Three observations in particular prompted further analysis of the data in greater detail. 
 First, there has been a steady increase in the number of queries made between 2012 and 2015. 
-\cref{fig:querylogs} provides a breakdown of the numbers on a monthly basis. 
+\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
 A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
@@
 \includegraphics[width=8.6cm]{dec24}
-\caption{Comparison of route planning queries on December 24th (Christmas Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
+\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
 The illustration shows that people started going home earlier than usual.}
@@
 Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
-\cref{fig:dec24} provides a breakdown of queries per hour on 24th December 2015 (Christmas Eve, on which Belgian families typically gather) in comparison to an average Thursday of the same year. 
-As illustrated, the evening peak on Christmas Eve occurs earlier than that of an average Thursday. 
+\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
+As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
 For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
-
-\subsection{Real-time logs}
-\label{sec:realtimelogs}
-
-iRail is a not for profit organization that stimulates digital creativity concerning mobility.
-The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
-As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-
-In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
-The JSON objects can be read one per one, so UNIX commands such as \emph{grep} work well with this format.
-We chose this over CSV as we have more flexibility towards changing the object properties in the future.
-Next, we chose to add an object to the stream which only contains an object with key \emph{@context}.
-This is the JSON-LD context of all the objects in the, JSON-stream.
-Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.
-
-Using this data, everyone can build their own historic query log dataset.
-The server costs for iRail to host this dataset stay limited.
-Furthermore, this way of publishing data enables research for real-time predictions of travel intentions.
-For this paper, we will focus on using the historic logs.
-Later, the research can be reproduced with more data.
-
-\section{Method}
-\label{sec:method}
-
-Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
-In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivities to be explored. 
-Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
-In this section we describe the data transformation procedure and data visualization technique employed in our case study.
-
-\subsection{Processing}
-
-Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
-Several stages of data processing are required to arrive at this outcome.
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays,
-\rv{why? and is it just you, or also the reader?}
-we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
-\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
-Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may overrepresent connections between certain stations. 
-\rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
-Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
-One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
-Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
-To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
-
-The data is then aggregated so that province level flow patterns are emphasised.
-\todo{Alvin: Figure must be drawn}
-\cref{fig:brussels} provides a diagrammatic representation of the procedure.
-First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See \cref{fig:brussels}a).
-If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (See \cref{fig:brussels}b).
-Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (See \cref{fig:brussels}c).
-The following types of flows are observable from the outcome of aggregation:
-
-\begin{itemize}
-  \item Travel from any provincial station to a major city.
-  \item Travel between any two major cities.
-  \item Travel between any two provincial stations.
-  \item Travel between any international station to a major city.
-  \item Travel between any international station to a provincial station.
-\end{itemize}
-
-\subsection{Visualization}
-
-Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
-This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
-Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
-Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
-\rv{OD, I'm assuming origin--destination, or something else? Is it necessary to use the abbreviation the whole time?}
-Flow visualizations can be broadly categorized into two groups.
-Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
-Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
-
-\rv{Starting here, this section isn't really going anywhere, at least not to a point. Consider removing the next two paragraphs and start talking about your data.}
-OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
-The OD Matrix is most common visualization technique in this type of representation. 
-Rows and columns correspond to locations while cells are coloured to express the volume of flow.
-OD matrices can be reordered to emphasise connectivity between locations.
-While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
-In recent years, such representations have become increasingly prominent for  displaying movement in scientific publications as well as popular culture.
-Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
-
-The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
-Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 
-Chords or curved line segments are drawn to connect nodes. 
-The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another. 
-In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
-Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.
-
-\section{Results}
-\label{sec:results}
-
-In this chapter, we report on the results of studying visualizations with experts of Trein Tram Bus, a not for profit passenger federation.
-\todo{Trein Tram Bus needs to be introduced in the introduction}
 
@@
 
-There are several inconsistent patterns in the data that do not correspond with an existing understanding of flows in Belgium.
+There are several inconsistent patterns in the data that do not correspond with existing understanding of flows in Belgium.
 In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
 Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API or a high concentration people who utilise route planning applications on a regular basis.
-\rv{Earlier, you said you filtered bots. Is this still the case?}
 Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
@@
 The expressiveness of a server affects the way logs can be gathered~\cite{usewod2015}.
-We have shown in previous sections that the query logs of a \emph{highly expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
+We have shown in previous sections that the query logs of a \emph{high expressive server}, such as the iRail API, are interesting: each request contains an entire query which can be interpreted as a travel intention.
 The possibility that two requests are exactly the same is low, as there are many URLs which can be requested.
@@
 
-First we studied query logs to find travel patterns in Belgium and can conclude this is up to now, the best representation of travel flows over the railway network in Belgium.
-%, as no alternatives are available.
+Firstly, we studied query logs to find travel patterns in Belgium.
+We conclude that this is up to now, the best representation of travel flows over the Belgian railway network.%, as no alternatives are available.
 We studied data which represent a maximum of 1.66\% of the passengers.
-\todo{ALVIN: Visualizing our dataset reveals the complexity of movement on the Belgian rail system and the significance of cities in weekday travel.
-Furthermore, we found evidence that Flanders is polycentric, while Walloon traffic converges on Brussels.}
+By visualizing the average distribution of requests on workdays and Saturday and Sunday, we were able to recognize the patterns we would expect: a.o.
+a morning and evening peak on workdays and a bigger evening peak on Friday and Sunday.
+More interestingly, we were able to see that the evening peak on Christmass Eve, Thursday the 24th of December 2015, started earlier than average on Thursdays.
+Visualizing our dataset using the origin and destination reveals the complexity of movement on the Belgian rail system and the significance of cities in weekday travel.
+Furthermore, we found evidence that Flanders is polycentric, while Walloon traffic is monocentric.
 
-Next, we also concluded that there are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
+Secondly, we also concluded that there are obvious caveats associated with the use of such data as proxy for actual statistical counts. 
 We identified a couple of gaps:
 \begin{enumerate}
-  \item The data only captures an intention of a passenger: it's unsure whether the person actually took the train.
+  \item The data only captures an intention of an end-user: it's unsure whether the person actually took the train.
   \item Multiple user queries may be needed to cover a travel intention.

### 2016-01-06 paper.tex

--- 2016-01-06 paper.tex
@@
 
-\subsection{Historical logs}
 The iRail API is an XML/JSON HTTP API: each request on \url{http://api.irail.be} returns a response that can be used directly in an end-user app.
@@
 
-We are particularly interested in studying the potential application of route planning queries as an information resource for flow analysis due to the input it captures from human users.
-The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards,
-and thus do not reflect direct human demands.
-The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015.
-\rv{Perhaps say \emph{we} made this copy available?}
-A copy of this dataset in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
-\rv{I would set the following in a list, preferably one line per thing, but an inline enumeration would also work.}
-Each data entry contains a unique identifier, a timestamp of the request executed, the HTTP status code of the response indicating a successful or failed operation, the request path between two stations and information of the user-agent (cfr.
-RFC2616) including details like the type of device, protocol and IP address.
-User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
-Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
-Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be utilised in scientific studies without further restrictions.
-
-\begin{figure}
-\centering
-\includegraphics[width=8.6cm]{querylogs}
-\caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
-In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
-\label{fig:querylogs}
-\end{figure}
-
-\rv{Here, you start with results. This seems to be a new section---I would recommend to start one.}
-Results from preliminary analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
-Three observations in particular prompted further analysis of the data in greater detail. 
-First, there has been a steady increase in the number of queries made between 2012 and 2015. 
-\cref{fig:querylogs} provides a breakdown of the numbers on a monthly basis. 
-A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
-Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
-Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
-Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
-
-\begin{figure}
-\centering
-\includegraphics[width=8.6cm]{avg-all}
-\caption{Distribution of iRail route planning API queries on average per day of the week between October 2012 and December 2015}
-\label{fig:average}
-\end{figure}
-
-Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
-\cref{fig:average} depicts a break down of the queries that occur, on average, on an hourly basis for each day of the week.
-Several trends can be observed in this chart. 
-Morning and evening peak periods in particular, are clearly distinguishable.
-At noon, a small dent is noticable, which may be attributed to part time workers and meetings on location after or before noon.
-Peak hours on weekdays are also distinctively different from those of weekends.
-Similarly, the frequency of queries during the evening peak period on Fridays appear to be substantially higher than those that occur between Monday and Thursday.
-The discrepancy maybe explained by students that travel home from their student homes over the weekends. 
-The absence of a clear peak on Saturdays can be attributed to lack of journeys to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
-
-\begin{figure}
-\centering
-\includegraphics[width=8.6cm]{dec24}
-\caption{Comparison of route planning queries on December 24th (Christmas Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
-The illustration shows that people started going home earlier than usual.}
-\label{fig:dec24}
-\end{figure}
-
-Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
-\cref{fig:dec24} provides a breakdown of queries per hour on 24th December 2015 (Christmas Eve, on which Belgian families typically gather) in comparison to an average Thursday of the same year. 
-As illustrated, the evening peak on Christmas Eve occurs earlier than that of an average Thursday. 
-For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
-
-\subsection{Real-time logs}
-\label{sec:realtimelogs}
-
-iRail is a not for profit organization that stimulates digital creativity concerning mobility.
-The interface that takes care of the route planning interface is open source, and we could easily suggest a change to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
-As of the 17th of December 2015, iRail publishes the most recent 1000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
-
+%IRail is a not for profit organization that stimulates digital creativity concerning mobility.
+This interface is open source and we were able to suggest changes to the system in which the query logs become open data\footnote{\url{https://github.com/iRail/iRail/pull/138}}.
+As of the 17th of December 2015, iRail publishes the most recent 1,000 requests done on the API as open data, each second, at \url{http://api.irail.be/logs/}.
 In order to publish this data, we chose a newline-based format: each new line contains a new query object encoded in JSON. 
@@
 We chose this over CSV as we have more flexibility towards changing the object properties in the future.
-Next, we chose to add an object to the stream which only contains an object with key \emph{@context}.
-This is the JSON-LD context of all the objects in the, JSON-stream.
+Next, we chose to add an HTTP header with a link to the query log context according to the JSON-LD specification\footnote{\url{}} to turn it into Linked Data.
+%This is the JSON-LD context of all the objects in the, JSON-stream.
 Subsequentially, we call this a \emph{JSONLD-stream} and this has been formalized at \emph{https://github.com/pietercolpaert/jsonld-stream}.
@@
 Furthermore, this way of publishing data enables research for real-time predictions of travel intentions.
-For this paper, we will focus on using the historic logs.
-Later, the research can be reproduced with more data.
+Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
+Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be published as open data.
+
+We are particularly interested in studying the potential application of route planning queries as an information resource for flow analysis due to the input it captures from human users.
+The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
+The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
+We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+%CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
+Each row within this dataset contains:
+\begin{enumerate}
+\item A timestamp of when the request was executed,
+\item The request path between two stations with longitudes and latitudes,
+\item The user-agent string cfr. RFC2616,
+\item The operating system of the user-agent and
+\item Whether or not the user-agent is a bot.
+\end{enumerate}
+%After analysis, it appeared 10\% of the request where performed by Google bots.
+%User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
+%\todo{add glue to Method section?}
 
@@
 Several stages of data processing are required to arrive at this outcome.
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays,
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
 \rv{why? and is it just you, or also the reader?}
-we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
-\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found above} 
-Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may overrepresent connections between certain stations. 
+\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?} 
+Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
 \rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
@@
 
-The data is then aggregated so that province level flow patterns are emphasised.
+The data is then aggregated so that province level flow patterns are emphasized.
 \todo{Alvin: Figure must be drawn}
@@
 \todo{Trein Tram Bus needs to be introduced in the introduction}
+
+\subsection{Time based analysis}
+
+\begin{figure}
+\centering
+\includegraphics[width=8.6cm]{querylogs}
+\caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
+In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
+\label{fig:querylogs}
+\end{figure}
+
+Results from the analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
+Three observations in particular prompted further analysis of the data in greater detail. 
+First, there has been a steady increase in the number of queries made between 2012 and 2015. 
+\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
+A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
+Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
+Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
+Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
+
+\begin{figure}
+\centering
+\includegraphics[width=8.6cm]{avg-all}
+\caption{Distribution of iRail route planning API queries on average per day of the week between October 2012 and December 2015}
+\label{fig:average}
+\end{figure}
+
+Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
+\cref{fig:average} depicts a break down of the queries that occur, on average, on an hourly basis for each day of the week.
+Several trends can be observed in this chart. 
+Morning and evening peak periods in particular, are clearly distinguishable.
+At noon, a small dent is noticable, which may be attributed to part time workers and meetings on location after or before noon.
+Peak hours on weekdays are also distinctively different from those of weekends.
+Similarly, the frequency of queries during the evening peak period on Fridays appear to be substantially higher than those that occur between Monday and Thursday.
+The discrepancy maybe explained by students that travel home from their student homes over the weekends. 
+The absence of a clear peak on Saturdays can be attributed to lack of journeys to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
+
+\begin{figure}
+\centering
+\includegraphics[width=8.6cm]{dec24}
+\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
+The illustration shows that people started going home earlier than usual.}
+\label{fig:dec24}
+\end{figure}
+
+Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
+\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
+As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
+For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
 

### 2016-01-06 paper.tex
Acted on RV's comments

@@
 
+\rv{Starting here, this section isn't really going anywhere, at least not to a point. Consider removing the next two paragraphs and start talking about your data.}
+OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
+The OD Matrix is most common visualization technique in this type of representation. 
+Rows and columns correspond to locations while cells are coloured to express the volume of flow.
+OD matrices can be reordered to emphasise connectivity between locations.
+While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
+In recent years, such representations have become increasingly prominent for  displaying movement in scientific publications as well as popular culture.
+Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
+
 \section{The query logs}
@@
 Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be published as open data.
+
+\todo{Maybe a sub header here?}
 
@@
 \rv{why? and is it just you, or also the reader?}
-\todo{PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?} 
-Queries made by automated user agents like search engine bots and data harvesters were also excluded from the dataset to remove biases that may over represent connections between certain stations. 
+\todo{Alvin: no longer a problem since i changed the text to remove any mention of exact hours.
+PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?}
+While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may may over represent connections between certain stations.
+In this regard, such queries have been excluded to improve the accuracy of our analysis.
 \rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
@@
 
-\rv{Starting here, this section isn't really going anywhere, at least not to a point. Consider removing the next two paragraphs and start talking about your data.}
-OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
-The OD Matrix is most common visualization technique in this type of representation. 
-Rows and columns correspond to locations while cells are coloured to express the volume of flow.
-OD matrices can be reordered to emphasise connectivity between locations.
-While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
-In recent years, such representations have become increasingly prominent for  displaying movement in scientific publications as well as popular culture.
-Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
-
 The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
@@
 The width at the head or tail of each chord indicates the amount of movement relative frequency of movement from a certain location to another.
-In our implementation, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
+In our implementation of the chord diagram, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
 Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.
@@
 \cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
-A dinstinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
+A distinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
 Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
@@
 In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
-Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API or a high concentration people who utilise route planning applications on a regular basis.
+Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
 \rv{Earlier, you said you filtered bots. Is this still the case?}

### 2016-01-06 paper.tex
some comments processed

--- 2016-01-06 paper.tex
@@
 we analyzed the query logs of the iRail API, a highly expressive route planning API for the Belgian railways.
-We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.5\%$ and $1.7\%$ of the amount of monthly passengers.
+We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.56\%$ and $1.66\%$ of the amount of monthly passengers.
 % Object & Findings
-Using data visualizations, we illustrate the commuting patterns in Belgium and confirm that Brussels, the capital, acts as a central hub.
+Using data visualizations, we illustrate that expected patterns such as 
+Moreover, using origin-destination visualizations, we can confirm that Brussels, the capital, acts as a central hub.
 The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
@@
 % Conclusion
-We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
+We conclude that query logs of route planners are of high importance in getting an indication of travel flows, yet better analytics would be acquirable using dedicated write-requests to capture travel intentions.
 % Perspectives
-\rv{These last sentences don't fit well. They could be removed.}
-Highly expressive transport data publishing methods such as route planning API exist, as well as low expressive data dumps or data fragments.
-In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
+%Highly expressive transport data publishing methods such as route planning API exist, as well as data dumps or data fragments.
+%In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
 
@@
 The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
-We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+We made a copy of this dataset in CSV format available at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 %CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
@@
 In this chapter, we report on the results of studying visualizations with experts of Trein Tram Bus, a not for profit passenger federation.
-\todo{Trein Tram Bus needs to be introduced in the introduction}
+%\todo{Trein Tram Bus needs to be introduced in the introduction}
 
@@
 In \cref{fig:LDFAxis2} we illustrate these two options as two extremes, with other options that are yet to be discovered.
-\rv{\emph{low expressive} is not correct, but I don't know what the right term is. (It is also high\emph{ly} expressive, I corrected that.)}
-When a \emph{rather low expressive} server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
+When a server only allows to set e.g., a departure station and a departure time, then the server cannot log the arrival station, yet the client is still able to plan a route by executing the algorithm on the client-side~\cite{lc}.
 We would be able to fully rely on the query logs if the expressivity would be maximal (extreme right) and caching would be turned off. 

### 2016-01-07 paper.tex

--- 2016-01-06 paper.tex
@@
 
-\rv{Starting here, this section isn't really going anywhere, at least not to a point. Consider removing the next two paragraphs and start talking about your data.}
 OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
@@
 As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
-\rv{why? and is it just you, or also the reader?}
-\todo{Alvin: no longer a problem since i changed the text to remove any mention of exact hours.
-PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?}
 While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may may over represent connections between certain stations.
 In this regard, such queries have been excluded to improve the accuracy of our analysis.
-\rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
 Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
@@
 In particular, the large number of trips made between Antwerp city and other stations within the province of Antwerp in the morning, is not observable in other major cities in Flanders.
-Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
-\rv{Earlier, you said you filtered bots. Is this still the case?}
+Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots.
+Alternatively the large number of trips may have originated from users who repeatedly request for updates on delayed trains or from a concentration of people who utilise route planning applications on a regular basis.
 Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
+\todo{not sure if we should move this somewhere else}
+As with any form of analysis based on appropriation of data, we acknowledge that more work is required to identify caveats and understand the validity of the findings yielded.
 

### 2016-01-07 paper.tex
processed feedback and comments

--- 2016-01-06 paper.tex
@@
 % Object & Findings
-Using data visualizations, we illustrate that expected patterns such as 
+We illustrate that we can see morning and evening peaks in the query logs.
 Moreover, using origin-destination visualizations, we can confirm that Brussels, the capital, acts as a central hub.
@@
 % Conclusion
-We conclude that query logs of route planners are of high importance in getting an indication of travel flows, yet better analytics would be acquirable using dedicated write-requests to capture travel intentions.
+We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
 % Perspectives
+However, better analytics would be acquirable using dedicated HTTP POST requests to capture travel intentions.
+%%Redundant according to RV:
 %Highly expressive transport data publishing methods such as route planning API exist, as well as data dumps or data fragments.
@@
 
-\todo{Rework introduction at the end}
+%\todo{Rework introduction at the end}
 
-%Today, public transport data still remains absent from the Linked Open Data cloud\footnote{\url{http://lod-cloud.net/}}.
-%Bizarre, one could notice, as many open transport datasets are already available, according to the Global Open Data Index of 2014\footnote{\url{http://index.okfn.org/dataset/timetables/2014/}}.
-%We can imagine that, what contributed to this, are the de-facto standard ways to publish public transit data to the Web.
-%There are currently two ways, as illustrated in \cref{fig:LDFAxis1}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
-
-%\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV-files, describing the rules for when a public transit vehicle will pass by on a certain location.
-%The specification is a huge success: up to date, it is supported among all current open source route planning software systems, and it is used in products such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
-%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
-%The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
-
-%On the other hand, public transit agencies also publish their data by providing \emph{route planners}.
-%These route planner offer a query service on top of the data and expose these over the Web.
-%However, when machines should access these, rate limiters are common place: only a limited amount of queries should be done in order for the service to be able to stay online.
-
-%\input{ldfaxis1}
+The \emph{iRail} project\footnote{\url{https://hello.irail.be}} started in 2008 to make the data of the Belgian railway accessible for developers.
+Ever since, the project offers developer both a data dump for third party apps ($\sim$~August~2015) and a route planning API ($\sim$~October~2010).
+The query logs of the latter has been stored since November 2012 (requests before that period got lost due to server migrations).
 
 Getting indications of people flows through public transit networks is a challenge.
-The data is tedious to get, mainly with public transit systems where passengers don't have to check-in and check-out.
-Nevertheless, people are calculating their intended routes by using the Web.
+Today, the Belgian railway company (SNCB) takes samples, manually, of people getting on and off trains in stations. 
+To date, this is still the SNCB's only source to determine people flows through their network\footnote{\url{http://www.belgianrail.be/nl/corporate/in-de-kijker/comptages_voyageurs_2014.aspx}}.
+Also within other transit systems the data is tedious to get, certainly when passengers do not have to check-in and check-out.
+
 \emph{Can we get an indication of these transit flows by studying the query logs of Web-services?}
 
-The \emph{iRail} project\footnote{\url{http://hello.irail.be}} started in 2008 to make the data of the Belgian railway accessible for developers.
-Ever since, the project offers developer both a GTFS data dump for third party apps and a route planning API.
-The query logs of this API has been stored since 2013.
+In this paper, we give a small overview of related work to gather data for flow analysis.
 We study these query logs by creating a couple of visualizations which illustrate a couple of patterns.
-Two of the documented patterns in this paper correspond to reality, another does not.
-
-In this paper, we give a small overview of related work to gather data for flow analysis.
 Next, we study the iRail query logs to find out whether we can find interesting patterns.
-Finally, we look at Linked~Connections~\cite{lc}, a proposed new way of publishing queryable public transit data, and whether we would still be able to have an indication of the transit flows with a Linked Data Fragments~\cite{ldf} approach.
+Finally, we look at public transit data is being published, and how we would be able to get a better .
 
@@
 
-\rv{Starting here, this section isn't really going anywhere, at least not to a point. Consider removing the next two paragraphs and start talking about your data.}
-OD type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
+In order to study data representing data flows, visualizing it is paramount.
+Origin Destination (OD) type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
 The OD Matrix is most common visualization technique in this type of representation. 
@@
 While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
-In recent years, such representations have become increasingly prominent for  displaying movement in scientific publications as well as popular culture.
+In recent years, such representations have become increasingly prominent for displaying movement in scientific publications as well as popular culture.\todo{citation needed?}
 Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
@@
 Furthermore, this way of publishing data enables research for real-time predictions of travel intentions.
-Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and IP addresses have been removed to protect the privacy of our users. 
-Additionally, the integrity of this dataset has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be published as open data.
+Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and, therefore, IP addresses have been removed to protect the privacy of our users. 
+Additionally, the integrity of this dataset, without IP addresses, has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be published as open data.
 
-\todo{Maybe a sub header here?}
-
-We are particularly interested in studying the potential application of route planning queries as an information resource for flow analysis due to the input it captures from human users.
-The remaining features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
+%\todo{Maybe a sub header here?} - PC: I wouldn't do a subheader...
+We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
+The other features are less relevant for this purpose as they are generally polled by digital signage providers or status boards, and thus do not reflect direct human demands.
 The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
@@
 \end{enumerate}
+
+\begin{figure}
+\centering
+\includegraphics[width=8.6cm]{querylogs}
+\caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
+In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
+\label{fig:querylogs}
+\end{figure}
+
+There has been an increase in the number of queries made between 2012 and 2015. 
+\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
+A distinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
+Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
+Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
+Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
+
 %After analysis, it appeared 10\% of the request where performed by Google bots.
-%User-agent information is important for statistical reasons and is essential for API management operations like tracing protocol violations or imposing rate limitations\footnote{\url{http://www.w3.org/Protocols/rfc2616/rfc2616-sec14.html}}.
 %\todo{add glue to Method section?}
@@
 Several stages of data processing are required to arrive at this outcome.
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
-\rv{why? and is it just you, or also the reader?}
-\todo{Alvin: no longer a problem since i changed the text to remove any mention of exact hours.
-PC: we will have to explain ourselves why we'd choose different hours than what we've found further on: could you change the intervals int he visualizations to what matches \cref{fig:average}?}
-While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may may over represent connections between certain stations.
+While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may over represent connections between certain stations.
 In this regard, such queries have been excluded to improve the accuracy of our analysis. 
-\rv{Be a bit more precise here: they are valid traffic, but not helpful for prediction of human behavior.}
 Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
@@
 Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
-To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) in PHP was employed to reduce spelling variations and link this with a geographical location.
+To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and link this with a geographical location.
 
@@
 Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
-This is key for our research since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
+This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
 Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
 Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
-\rv{OD, I'm assuming origin--destination, or something else? Is it necessary to use the abbreviation the whole time?}
 Flow visualizations can be broadly categorized into two groups.
-Representations that preserve the complete trajectory and OD type representations that take only the start and end locations into account.
+Representations that preserve the complete trajectory and Origin Destination type representations that take only the start and end locations into account.
 Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
@@
 
-In this chapter, we report on the results of studying visualizations with experts of Trein Tram Bus, a not for profit passenger federation.
-%\todo{Trein Tram Bus needs to be introduced in the introduction}
+In this chapter, we report on the results of studying visualizations with experts of \emph{Trein Tram Bus}\footnote{\url{http://treintrambus.be}}, a not for profit passenger federation.
+%Isn't this sufficient ↑? \todo{Trein Tram Bus needs to be introduced in the introduction}
 
 \subsection{Time based analysis}
-
-\begin{figure}
-\centering
-\includegraphics[width=8.6cm]{querylogs}
-\caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
-In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
-\label{fig:querylogs}
-\end{figure}
-
-Results from the analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
-Three observations in particular prompted further analysis of the data in greater detail. 
-First, there has been a steady increase in the number of queries made between 2012 and 2015. 
-\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
-A distinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
-Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
-Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
-Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
 
@@
 
-Secondly, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
+Results from the analysis of this dataset reveal trends in route planning queries that may provide insights into rail travel demand. 
+First, the daily and hourly distribution of queries appear to correspond with known commuting patterns. 
 \cref{fig:average} depicts a break down of the queries that occur, on average, on an hourly basis for each day of the week.
@@
 
-Finally, the distribution of queries on public holidays is observed to deviate from that of an average day. 
-\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, 24th December 2015, in comparison to an average Thursday of the same year. 
+Moreover, the distribution of queries on public holidays is observed to deviate from that of an average day. 
+\cref{fig:dec24} provides a breakdown of queries per hour on Christmas eve, December 24th 2015, in comparison to an average Thursday of the same year. 
 As illustrated, the evening peak on Christmas eve occurs earlier than that of an average Thursday. 
-For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
+%For these reasons, we believe that route planning query data maybe a viable source of information for flow analysis.%\todo{where people are heading between certain time intervals?}
 
@@
 
-\subsection{Inexplicable results}
+\subsection{Counterintuitive results}
 
@@
 Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
-\rv{Earlier, you said you filtered bots. Is this still the case?}
-Accordingly, the reverse is identified in Liege where an unusually large connection originating from a suburb towards Liege city prompted speculation as well as discussions on how such outliers should be treated or what may have led to their existence.
+Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liege city.
+This prompted speculation, as well as discussions, on how such outliers should be treated or what may have led to their existence.
 
@@
 
-%It is in the interest of public transit agencies that the use of their time schedule data is maximized.
-%We can imagine that when more people are informed about the agency's offer, there will be a higher amount of people that will use the service.
-%There are currently two ways to publish transit data, as illustrated in \cref{fig:LDFAxis2}: publishing data using the General Transit Feed Specification (GTFS) and offering a route planning service.
-
-%\emph{GTFS}, on the one hand, is a data dump format: it is a compressed ZIP-file, containing a couple of CSV files, describing the rules for when a public transit vehicle will pass by on a certain location.
-%GTFS is supported among all current open source route planning software systems and it is used in products/apps such as \emph{CityMapper}, \emph{Ally}, \emph{Navitia.io}, \emph{Google Maps} and \emph{Bing Maps}.
-%Persistence of the identifiers used within these datasets, is not a requirement, neither is it an ambition of the format\footnote{\url{https://groups.google.com/forum/#!msg/gtfs-changes/Z8Mf31MaZms/8Hc9F4psAQAJ}}.
-%The goal of GTFS is to create an exchange format which specific software packages can use to transform it into their format of choice.
-
-%On the other hand, public transit agencies also publish their data by providing \emph{route planners} such as the French railway company\footnote{\url{http://tech.eu/features/7119/sncf-api-real-time-data/}}, the Dutch railway company\footnote{\url{http://www.ns.nl/reisinformatie/ns-api}} or \ldots\footnote{\url{http://otherapiofadifferentagency}}
-%These route planner offer a query service on top of the data and expose these over the Web.
-%In this paper we have studied the query logs of such an \emph{expressive} route planning API, which per request offers an interesting log entry.
-
 %label: fig:LDFAxis2
@@
 
-%Linked Data Fragments~\cite{ldf} is a way to think about data publishing as a set of trade-offs between client and servers.
-%These trade-offs can be visualized on the axis in \cref{fig:LDFAxis2}: ticks in between the extremes indicate different client/server options.
-%For route planning systems, we call these other ticks Linked~Connections~\cite{lc}: 
-%by publishing fragments of the data needed by an earliest arrival time algorithm, the algorithm can be executed by a user agent while the downloading the data.
-%An example of such an implementation is illustrated in \cref{fig:lc}.
-%By exploiting the caching mechanisms behind HTTP, it lowers to load on the server, making it easier to keep the data high available.
-%Furthermore, it enables clients to federate queries over different servers representing different regions, transit modes or different user requirements such as low criminality rates or wheelchair accessibility.
-
-%\begin{figure}[h]
-%    \centering
-%    \includegraphics[width=0.48\textwidth]{LC}
-%    \caption{An example implementation of a route planning algorithm: fragments of data needed by a route planning algorithm are published in pages.
-The route planning algorithm will need multiple requests instead of just one.
-The fragements of the pages that can be retrieved are however highly cacheable.}
-%    \label{fig:lc}
-%\end{figure}
-%
-
-%%% EXPLAIN:
-%%%
 The expressiveness of a server affects the way logs can be gathered~\cite{usewod2015}.
@@
 \begin{enumerate}
-  \item The data only captures an intention of an end-user: it's unsure whether the person actually took the train.
+  \item The data only captures an intention of an end-user: it is unsure whether the person actually took the train.
   \item Multiple user queries may be needed to cover a travel intention.

### 2016-01-07 paper.tex

@@
 % Object & Findings
-We illustrate that we can see morning and evening peaks in the query logs.
-Moreover, using origin-destination visualizations, we can confirm that Brussels, the capital, acts as a central hub.
+Using data visualizations, we illustrate the commuting patterns in Belgium and confirm that Brussels, the capital, acts as a central hub. 
 The Flemish region appears to be polycentric, while in the Walloon region, everything converges on Brussels.
@@
 We chose this over CSV as we have more flexibility towards changing the object properties in the future.
-Next, we chose to add an HTTP header with a link to the query log context according to the JSON-LD specification\footnote{\url{}} to turn it into Linked Data.
+Next, we chose to add an HTTP header with a link to the query log context according to the JSON-LD specification\footnote{\url{http://www.w3.org/TR/json-ld/}} to turn it into Linked Data.
 %This is the JSON-LD context of all the objects in the, JSON-stream.
@@
 The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
-We made a copy of this dataset in CSV format available at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
 %CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
@@
 %After analysis, it appeared 10\% of the request where performed by Google bots.
-%\todo{add glue to Method section?}
 
@@
 Several stages of data processing are required to arrive at this outcome.
+%As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
 While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may over represent connections between certain stations.
@@
 Further investigation led to the identification of an isolated connection originating from Antwerp city towards a suburb, suggesting anomalous usage of the iRail API that were not flagged as bots or a high concentration people who utilise route planning applications on a regular basis.
+Alternatively the large number of trips may have originated from users who repeatedly request for updates on delayed trains or from a concentration of people who utilise route planning applications on a regular basis.
 Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liege city.
 This prompted speculation, as well as discussions, on how such outliers should be treated or what may have led to their existence.
+As with any form of analysis based on appropriation of data, we acknowledge that more work is required to identify caveats and understand the validity of the findings presented.
 

### 2016-01-07 paper.tex

@@
 %\todo{Rework introduction at the end}
+Today, the Belgian railway company (SNCB) takes manual samples of people getting on and off trains in stations. 
+To date, this is still the SNCB's only source to determine people flows through their network\footnote{\url{http://www.belgianrail.be/nl/corporate/in-de-kijker/comptages_voyageurs_2014.aspx}}.
+Also within other transit systems the data is tedious to get, certainly when passengers do not have to check-in and check-out.
+Nonetheless, indications of how people move within your transit network is of uttermost importance for, amongst others, taking demand-driven policy decisions concerning mobility.
 
@@
 Getting indications of people flows through public transit networks is a challenge. 
-Today, the Belgian railway company (SNCB) takes samples, manually, of people getting on and off trains in stations. 
-To date, this is still the SNCB's only source to determine people flows through their network\footnote{\url{http://www.belgianrail.be/nl/corporate/in-de-kijker/comptages_voyageurs_2014.aspx}}.
-Also within other transit systems the data is tedious to get, certainly when passengers do not have to check-in and check-out.
-
-\emph{Can we get an indication of these transit flows by studying the query logs of Web-services?}
+Yet, \emph{can we not get an indication of these transit flows by studying the query logs of Web-services?}
 
 In this paper, we give a small overview of related work to gather data for flow analysis.
-We study these query logs by creating a couple of visualizations which illustrate a couple of patterns.
-Next, we study the iRail query logs to find out whether we can find interesting patterns.
-Finally, we look at public transit data is being published, and how we would be able to get a better .
+Next, we introduce the iRail query logs, which we study by the means of visualizations.
+Finally, we look at public transit data is being published, and how we would be able to get a better indication of travel flows than just mere query logs.
 

### 2016-01-07 paper.tex
Relwork

@@
 
-%\todo{Rework introduction at the end}
 Today, the Belgian railway company (SNCB) takes manual samples of people getting on and off trains in stations. 
@@
 Also within other transit systems the data is tedious to get, certainly when passengers do not have to check-in and check-out.
-Nonetheless, indications of how people move within your transit network is of uttermost importance for, amongst others, taking demand-driven policy decisions concerning mobility.
+Nonetheless, indications of how people move within a transit network is of uttermost importance for, amongst others, making demand-driven policy decisions concerning mobility.
 
 The \emph{iRail} project\footnote{\url{https://hello.irail.be}} started in 2008 to make the data of the Belgian railway accessible for developers.
-Ever since, the project offers developer both a data dump for third party apps ($\sim$~August~2015) and a route planning API ($\sim$~October~2010).
+The project offers developers both a data dump for third party apps (since $\sim$~August~2015) and a route planning API (since $\sim$~October~2010).
 The query logs of the latter has been stored since November 2012 (requests before that period got lost due to server migrations).
@@
 
-\todo{Write relwork at the end}
-
 \emph{Flow analysis} is a topic of theoretical interest and practical importance in various disciplines. 
@@
 Smart card data from public transport systems~\cite{roth,beecham}, taxi journeys~\cite{ferreira} as well as cellular call data~\cite{sevtsuk} have provided planners with new opportunities to develop greater understanding of mobility patterns in urban environments~\cite{batty}.
+Yet query logs were, to our knowledge, never used for flow analysis so far.
 
-In order to study data representing data flows, visualizing it is paramount.
+On highly expressive servers, such as \sparql-endpoints, query logs can play an important role in fine-tuning the underlying datastores~\cite{arias2011empirical}.
+E.g., they determined that in 2011 in the context of the graph query language \sparql, most of the queries are simple and include few triple patterns and joins.
+Furthermore, the graph patterns are usually star-shaped and despite triple pattern chains exist, they are generally short.
+Other, more cacheable interfaces more to the left on the Linked~Data~Fragments axis~\cite{ldf}, also benefit from query logs in order to e.g., know who is using what clients to query the Web, yet the user intention is lost~\cite{usewod2015}.
+This was called ``a blessing for privacy'', yet adding the user query as an HTTP header was suggested.
+
+Visualizations are crucial to understanding the patterns in data representing transit flows.
 Origin Destination (OD) type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
@@
 While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
-In recent years, such representations have become increasingly prominent for displaying movement in scientific publications as well as popular culture.\todo{citation needed?}
+In recent years, such representations have become increasingly prominent for displaying movement in scientific publications as well as popular culture.
 Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced.
+\todo{PC: Alvin, can we have a citation for this paragraph? Maybe you can cite one your own papers here?}
 

### 2016-01-07 paper.tex

@@
 \usetikzlibrary{arrows,positioning,shapes.misc}
+%\let\subcaption\relax
 \usepackage{graphicx}
+%\usepackage[caption=false]{subfig}
+%\usepackage{caption}
+\usepackage{subcaption}
+\captionsetup[figure]{labelfont=bf,textfont=bf}
+\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont}
 \definecolor{lightgrey}{RGB}{170, 170, 170}
@@
 
+\begin{figure*}
+  \centering
+  \begin{subfigure}{.475\textwidth}
+    \includegraphics[width=8cm]{method-a.pdf}\\
+    \subcaption{Spatial aggregation based on provincial administrative boundaries}
+    \label{fig:method-a}
+  \end{subfigure}%
+  \begin{subfigure}{.475\textwidth}
+    \includegraphics[width=8cm]{method-b.pdf}\\
+    \subcaption{\todo{Alvin: caption to explain what this is}}
+    \label{fig:method-b}
+  \end{subfigure}%
+  \hfill%
+  \begin{subfigure}{0.9\textwidth}
+    \includegraphics[width=15cm]{method-c.pdf}\\
+    \subcaption{Yet another subfigure \todo{Alvin: caption to explain what this is}}
+    \label{fig:method-c}
+  \end{subfigure}%
+  \caption{General caption of the image \todo{Alvin: caption to explain what this is}}
+  \label{fig:method}
+\end{figure*}
+
 The data is then aggregated so that province level flow patterns are emphasized.
-\todo{Alvin: Figure must be drawn}
-\cref{fig:brussels} provides a diagrammatic representation of the procedure.
-First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (See \cref{fig:brussels}a).
-If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (See \cref{fig:brussels}b).
-Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (See \cref{fig:brussels}c).
+\cref{fig:method} provides a diagrammatic representation of the procedure.
+First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (see \cref{fig:method-a}).
+Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (see \cref{fig:method-b}).
+If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (see \cref{fig:method-c}).
 The following types of flows are observable from the outcome of aggregation:

### 2016-01-07 paper.tex
Figure 3 and 4 on same page
Figure placement rework

@@
 
+\begin{figure*}
+  \centering
+  \includegraphics[width=14cm]{method-c.pdf}\\
+  \caption{Yet another figure \todo{Alvin: caption to explain what this is}}
+  \label{fig:method-c}
+\end{figure*}
+
 Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
@@
 
-\begin{figure*}
+\begin{figure}[h]
   \centering
-  \begin{subfigure}{.475\textwidth}
-    \includegraphics[width=8cm]{method-a.pdf}\\
-    \subcaption{Spatial aggregation based on provincial administrative boundaries}
+  \begin{subfigure}{.2375\textwidth}
+    \includegraphics[width=4cm]{method-a.pdf}\\
+    \subcaption{Lines and dots\todo{Alvin}} % 4 words max
     \label{fig:method-a}
   \end{subfigure}%
-  \begin{subfigure}{.475\textwidth}
-    \includegraphics[width=8cm]{method-b.pdf}\\
-    \subcaption{\todo{Alvin: caption to explain what this is}}
+  \begin{subfigure}{.2375\textwidth}
+    \includegraphics[width=4cm]{method-b.pdf}\\
+    \subcaption{arrows and dots\todo{Alvin}} % 4words max
     \label{fig:method-b}
   \end{subfigure}%
-  \hfill%
-  \begin{subfigure}{0.9\textwidth}
-    \includegraphics[width=15cm]{method-c.pdf}\\
-    \subcaption{Yet another subfigure \todo{Alvin: caption to explain what this is}}
-    \label{fig:method-c}
-  \end{subfigure}%
-  \caption{General caption of the image \todo{Alvin: caption to explain what this is}}
-  \label{fig:method}
-\end{figure*}
+  \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+  \caption{With the arrow and dots and the lines and dots you can see that TODO \todo{Alvin: can you write a caption here?}}
+\end{figure}
 
@@
 \subsection{Time based analysis}
-
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 \label{fig:average}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=8.6cm]{dec24}
+\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
+The illustration shows that people started going home earlier than usual.}
+\label{fig:dec24}
 \end{figure}
@@
 The absence of a clear peak on Saturdays can be attributed to lack of journeys to work queries while the evening peak on Sundays may emerge from students returning to college accommodations.
-
-\begin{figure}
-\centering
-\includegraphics[width=8.6cm]{dec24}
-\caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
-The illustration shows that people started going home earlier than usual.}
-\label{fig:dec24}
-\end{figure}
 

### 2016-01-07 paper.tex
Better last sentence of abstract
Modified section 4.1

@@
 Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
-In this section we describe the data transformation procedure and data visualization technique employed in our case study.
+Several stages of data processing are required to arrive at this outcome.
 
-\subsection{Processing}
+\subsection{Data Processing}
+
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
+Queries made by automated user-agents like search engine bots and data harvesters can be considered as valid traffic but may over represent connections between certain stations.
+We have excluded such queries since we are specifically interested in studying human travel behavior.
+Station names tend to contain be highly inconsistent since any text information can be parsed to the API.
+This may include invalid character encoding, spelling mistakes and the use of unstandardized abbreviations.
+We employ an open-source reconciliator (\url{https://github.com/irail/stations}) to reduce spelling variations and remove data records with missing information. 
+Additionally, station names are automatically linked to their correspodinging geographical coordinates through this process.
 
@@
   \includegraphics[width=14cm]{method-c.pdf}\\
-  \caption{Yet another figure \todo{Alvin: caption to explain what this is}}
+  \caption{Spatial aggregation based on provincial administrative boundaries.
+Each station is color coded to indicate the province it will be aggregated to.
+Stations in major cities as well as stations outside of Belgium are aggregated to seperate distinct groups.}
   \label{fig:method-c}
 \end{figure*}
-
-Matching pairs of origin and destination (OD) stations must be aggregated prior to visualization.
-Several stages of data processing are required to arrive at this outcome.
-%As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
-While queries made by automated user agents like search engine bots and data harvesters can be considered as valid traffic, they may over represent connections between certain stations.
-In this regard, such queries have been excluded to improve the accuracy of our analysis. 
-Similarly, data records with unparsable origin or destination station names due to invalid character encoding were removed. 
-One distinctive challenge is the substantial amount of inconsistency in the way station names are logged. 
-Spelling mistakes, and the use of unstandardized abbreviations in particular, reduce the quality and accuracy of the resulting visualization.
-To overcome this challenge, an open-source reconciliator (\url{https://github.com/irail/stations}) was employed to reduce spelling variations and link this with a geographical location.
 
@@
 
-The data is then aggregated so that province level flow patterns are emphasized.
+Next, the data is aggregated so that province level flow patterns are emphasized.
 \cref{fig:method} provides a diagrammatic representation of the procedure.

### 2016-01-07 paper.tex
Modified section 4 + figures 2 & 3

@@
 \item A timestamp of when the request was executed,
-\item The request path between two stations with longitudes and latitudes,
+\item The request path between two stations and their associated geographic coordinates,
 \item The user-agent string cfr. RFC2616,
 \item The operating system of the user-agent and
-\item Whether or not the user-agent is a bot.
+\item A flag for whether the user-agent is a bot.
 \end{enumerate}
@@
 
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hrs and 10:00 hrs as well as 17:00 hrs and 21:00 hrs on weekdays, and all data records created on weekend.
+As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hours and 10:00 hours as well as 17:00 hours and 21:00 hours on weekdays, and all data records created on weekend.
 Queries made by automated user-agents like search engine bots and data harvesters can be considered as valid traffic but may over represent connections between certain stations.
-We have excluded such queries since we are specifically interested in studying human travel behavior.
+Such queries have been excluded since we are specifically interested in studying human travel behavior.
 Station names tend to contain be highly inconsistent since any text information can be parsed to the API.
@@
 We employ an open-source reconciliator (\url{https://github.com/irail/stations}) to reduce spelling variations and remove data records with missing information. 
-Additionally, station names are automatically linked to their correspodinging geographical coordinates through this process.
+Additionally, it automatically links station names to their corresponding geographical coordinates.
 
@@
   \includegraphics[width=14cm]{method-c.pdf}\\
-  \caption{Spatial aggregation based on provincial administrative boundaries.
-Each station is color coded to indicate the province it will be aggregated to.
-Stations in major cities as well as stations outside of Belgium are aggregated to seperate distinct groups.}
-  \label{fig:method-c}
+  \caption{Each station is color coded to indicate the province it will be aggregated to. Stations in major cities as well as stations outside of Belgium are aggregated to seperate groups.}
+  \label{fig:method-b}
 \end{figure*}
@@
     \includegraphics[width=4cm]{method-a.pdf}\\
-    \subcaption{Lines and dots\todo{Alvin}} % 4 words max
-    \label{fig:method-a}
+    \subcaption{} % 4 words max
+    %\label{fig:method-a}
   \end{subfigure}%
@@
     \includegraphics[width=4cm]{method-b.pdf}\\
-    \subcaption{arrows and dots\todo{Alvin}} % 4words max
-    \label{fig:method-b}
+    \subcaption{} % 4words max
+    %\label{fig:method-b}
   \end{subfigure}%
   \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-  \caption{With the arrow and dots and the lines and dots you can see that TODO \todo{Alvin: can you write a caption here?}}
+  \label{fig:method-a}
+  \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
+(a) Individual data records are depicted geographically with arrows connecting origins to their respective destinations.
+(b) The result of aggregation is a set of weighted connections between provinces.}
 \end{figure}
@@
 Next, the data is aggregated so that province level flow patterns are emphasized.
-\cref{fig:method} provides a diagrammatic representation of the procedure.
-First, both origin and destination stations for each data record are spatially aggregated based on provincial administrative boundaries (see \cref{fig:method-a}).
-Train stations in major cities are excluded from aggregation so that the volume of flow between provinces and major cities are comparable (see \cref{fig:method-b}).
-If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group (see \cref{fig:method-c}).
+\todo{Pieter: I can't seem to link the figure here} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
+To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities are aggregated seperately. 
+If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group.
+A diagrammtic breakdown is depicted in \todo{Pieter: Same problem here...} where color and shape is used to demarcate each group.
 The following types of flows are observable from the outcome of aggregation:

### 2016-01-07 paper.tex
Added figure references

@@
 % Perspectives
-However, better analytics would be acquirable using dedicated HTTP POST requests to capture travel intentions.
+However, better travel intentions would be acquirable using dedicated HTTP POST requests.
 %%Redundant according to RV:
@@
   \caption{Each station is color coded to indicate the province it will be aggregated to. Stations in major cities as well as stations outside of Belgium are aggregated to seperate groups.}
-  \label{fig:method-b}
+  \label{fig:method-c}
 \end{figure*}
@@
     \subcaption{} % 4 words max
-    %\label{fig:method-a}
+    \label{fig:method-a}
   \end{subfigure}%
@@
     \subcaption{} % 4words max
-    %\label{fig:method-b}
+    \label{fig:method-b}
   \end{subfigure}%
   \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-  \label{fig:method-a}
+  \label{fig:method}
   \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
@@
 Next, the data is aggregated so that province level flow patterns are emphasized.
-\todo{Pieter: I can't seem to link the figure here} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
+\cref{fig:method} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
 To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities are aggregated seperately. 
 If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group. 
-A diagrammtic breakdown is depicted in \todo{Pieter: Same problem here...} where color and shape is used to demarcate each group.
+A diagrammatic breakdown is depicted in \cref{fig:method-c} where color and shape is used to demarcate each group.
 The following types of flows are observable from the outcome of aggregation:

### 2016-01-08 paper.tex
Fixed label
Figures (no captions)

@@
 
-\subsection{Time based analysis}
+\subsection{Time Based Analysis}
 \begin{figure}[h]
@@
 
-\subsection{Structure of flows in Belgium}
+\subsection{Structure of Flows in Belgium}
 
@@
   \centering
-\includegraphics[width=15cm]{brussels}
-\caption{Pair of chord visualization of the city of Brussels: transit flows towards Brussels from elsewhere in Belgium and vice versa.}
-\label{fig:brussels}
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=13cm]{weekdays_morning.pdf}\\
+    \subcaption{Weekdays} % 4 words max
+    \label{fig:weekdays_morning}
+  \end{subfigure}
+  \begin{subfigure}{\textwidth}
+    \centering
+    \includegraphics[width=13cm]{weekends_morning.pdf}\\
+    \subcaption{Weekends} % 4words max
+    \label{fig:weekends_morning}
+  \end{subfigure}
+  \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+  \label{fig:mornings}
+  \caption{\todo{Needs a caption!}}
 \end{figure*}
 
-\begin{figure*}
+\begin{figure*}[t]
+\begin{tabular}{ccc}
+  \includegraphics[width=.33\textwidth]{wd-brussels.pdf}& 
+  \includegraphics[width=.33\textwidth]{wd-antwerp.pdf}& 
+  \includegraphics[width=.33\textwidth]{wd-gent.pdf}\\
+  (a) Brussels City&(b) Antwerp City&(c) Gent\\[6pt]
+  \includegraphics[width=.33\textwidth]{wd-leuven.pdf}& 
+  \includegraphics[width=.33\textwidth]{wd-liege.pdf}& 
+  \includegraphics[width=.33\textwidth]{wd-mons.pdf}\\ 
+  (d) Leuven&(e) Liege City&(f) Mons\\
+\end{tabular}
+\label{fig:cities}
 \centering
-\includegraphics[width=15cm]{antwerp}
-\caption{Pair of chord visualization of the city of Antwerp: transit flows towards Antwerp from elsewhere in Belgium and vice versa.}
-\label{fig:antwerp}
+\caption{\todo{Needs a caption!}}
 \end{figure*}
 
-\subsection{Counterintuitive results}
+\subsection{Counter Intuitive Results}
 

### 2016-01-08 paper.tex
Modified related work

@@
 
-Visualizations are crucial to understanding the patterns in data representing transit flows.
-Origin Destination (OD) type representations are used when the origin and destination preside over intermediate locations on the path of movement. 
-The OD Matrix is most common visualization technique in this type of representation. 
+%Visualizations are crucial for understanding the patterns in data and representing transit flows.
+Visualizations are crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
+This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
+Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support flow analysis \cite{andrienko2012visual}. 
+Flow visualizations can be broadly categorized into two groups. 
+Representations that display the complete trajectory and Origin Destination (OD) type representations that take only the start and end locations into account.
+Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
+OD type representations are commonly employed when the origin and destination preside over intermediate locations on the path of movement.
+
+The OD Matrix is most common visualization technique based on this type of representation. 
 Rows and columns correspond to locations while cells are coloured to express the volume of flow.
@@
 While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
-In recent years, such representations have become increasingly prominent for displaying movement in scientific publications as well as popular culture.
+In recent years, such representations have become increasingly prominent for displaying movement in science as well as popular culture.
 Locations are depicted as nodes and arrows are drawn between them to indicate movement in a certain direction, allowing for exact and individual relationships to be traced. 
-\todo{PC: Alvin, can we have a citation for this paragraph? Maybe you can cite one your own papers here?}
+This representation is detailed and straightforward to understand but becomes cluttered when multiple lines intersect. 
+A number of solutions have been proposed for clutter reduction. 
+Filtering is the most commonly practiced. 
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \todo{Andrienko, 2007 -198}. 
+Another solution is to transform the visual representation. 
+Flow maps \todo{Phan, 2005 -88} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that emphasize key geographic features such as roads, rivers or coastlines. 
+\todo{Alvin: Will expand abit more to describe why chord diagrams were chosen}
 
@@
 
-%\todo{Maybe a sub header here?} - PC: I wouldn't do a subheader...
 We are particularly interested in studying the potential application of \emph{route planning} queries as an information resource for flow analysis due to the input it captures from human users.
@@
 \item The operating system of the user-agent and
-\item A flag for whether the user-agent is a bot.
+\item A flag stating if the user-agent is a bot.
 \end{enumerate}
@@
 Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
-Several stages of data processing are required to arrive at this outcome.
+The data must undergo several stages of data processing to arrive at this outcome.
 
@@
 
-As we were specifically interested in studying morning and evening rush hour travel that occur on weekdays, we filtered the dataset to exclude data records created outside the time range of 06:00 hours and 10:00 hours as well as 17:00 hours and 21:00 hours on weekdays, and all data records created on weekend.
+We were specifically interested in understanding how travel during the morning rush hour on weekdays differ from weekends.
+In this regard, the dataset was filtered to exclude data records created beyond the time range of 06:00 hours and 10:00 hours.
 Queries made by automated user-agents like search engine bots and data harvesters can be considered as valid traffic but may over represent connections between certain stations.
-Such queries have been excluded since we are specifically interested in studying human travel behavior.
+Accordingly, such queries were excluded since they do not explicitly represent human travel behavior.
 Station names tend to contain be highly inconsistent since any text information can be parsed to the API.
 This may include invalid character encoding, spelling mistakes and the use of unstandardized abbreviations.
-We employ an open-source reconciliator (\url{https://github.com/irail/stations}) to reduce spelling variations and remove data records with missing information. 
-Additionally, it automatically links station names to their corresponding geographical coordinates.
+We employed an open-source reconciliator (\url{https://github.com/irail/stations}) to reduce spelling variations and remove data records with missing information. 
+Through this process, station names were also linked to their corresponding geographical coordinates.
 
@@
 \cref{fig:method} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
-To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities are aggregated seperately. 
-If an origin or destination station is a located outside of Belgium, it is considered international travel and aggregated in a distinct group. 
+To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities were aggregated into seperate groups. 
+Simiarly, stations located outside of Belgium were aggregated to distinct group. 
 A diagrammatic breakdown is depicted in \cref{fig:method-c} where color and shape is used to demarcate each group.
@@
 \subsection{Visualization}
-
-Visualization is crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
-This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
-Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support analysis \cite{andrienko2012visual}. 
-Of these techniques, flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
-Flow visualizations can be broadly categorized into two groups.
-Representations that preserve the complete trajectory and Origin Destination type representations that take only the start and end locations into account.
-Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
-
+%Flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
 The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
@@
 
-The chord diagrams shown in \cref{fig:brussels} and \cref{fig:antwerp} depict the aggregated number of queries made between any two stations in the morning and evening respectively. 
-Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in weekday travel.
-Turning our attention to morning commutes, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
-The reverse occurs in the evening when trips are made towards provincial stations.
-Brussels stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
+The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (See \cref{fig:weekdays_morning}) and weekends (See \cref{fig:weekends_morning}) respectively.
+Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.
+%We can observe that Brussels, Antwerp, Gent and Leuven are the dominant destinations.
+%Turning our attention to weekday travel, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
+%The reverse occurs in the evening when trips are made towards provincial stations.
+Brussels City stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity (see \cref{fig:mornings}a).
 On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
-With exception to Liege, queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
-Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Ghent, Brussels, Antwerp and Leuven.
+With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries.
+Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Gent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
 These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels as a major centre.

### 2016-01-08 paper.tex

@@
 Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Gent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
-These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels as a major centre.
+These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels city as a major centre.
 The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
@@
   \label{fig:mornings}
-  \caption{\todo{Needs a caption!}}
+  \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
 \end{figure*}
@@
 \centering
-\caption{\todo{Needs a caption!}}
+\caption{Comparison of morning rush hour travel flows on weekdays originating from or towards six major cities in Belgium.}
 \end{figure*}

### 2016-01-08 paper.tex
Merged and added dist with paper compilation in it
Modified related work

@@
 Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support flow analysis \cite{andrienko2012visual}. 
-Flow visualizations can be broadly categorized into two groups. 
+Visualizations of flows can be broadly categorized into two groups. 
 Representations that display the complete trajectory and Origin Destination (OD) type representations that take only the start and end locations into account.
 Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
-OD type representations are commonly employed when the origin and destination preside over intermediate locations on the path of movement.
 
-The OD Matrix is most common visualization technique based on this type of representation. 
+OD type representations are employed when the origin and destination preside over intermediate locations on the path of movement.
+The OD Matrix is the most common visualization technique based on this type of representation.
 Rows and columns correspond to locations while cells are coloured to express the volume of flow.
@@
 Another solution is to transform the visual representation. 
-Flow maps \todo{Phan, 2005 -88} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that emphasize key geographic features such as roads, rivers or coastlines. 
-\todo{Alvin: Will expand abit more to describe why chord diagrams were chosen}
+Flow maps \todo{Phan, 2005 -88} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that do not obscure key geographic features such as roads, rivers or coastlines.
+However, this solution is optimal only if strict geographic accuracy is not required since the arrows do not depict actual paths of movement
 

### 2016-01-08 paper.tex
Modified figure 7 caption

@@
   \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-  \label{fig:method}
   \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
@@
 (b) The result of aggregation is a set of weighted connections between provinces.}
+  \label{fig:method-ab}
 \end{figure}
@@
 Next, the data is aggregated so that province level flow patterns are emphasized.
-\cref{fig:method} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
+\cref{fig:method-ab} provides a diagrammatic representation of spatial aggregation on the basis of provincial administrative boundaries.
 To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities were aggregated into seperate groups. 
@@
   \centering
-  \begin{subfigure}{\textwidth}
+  \begin{subfigure}{.95\textwidth}
     \centering
@@
   \end{subfigure}
-  \begin{subfigure}{\textwidth}
+  \begin{subfigure}{.95\textwidth}
     \centering
@@
   \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+  \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
   \label{fig:mornings}
-  \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
 \end{figure*}
@@
 \centering
-\caption{Comparison of morning rush hour travel flows on weekdays originating from or towards six major cities in Belgium.}
+\caption{Comparison of morning rush hour travel flows on weekdays originating from or destined for six major cities in Belgium.}
 \end{figure*}

### 2016-01-08 paper.tex
small fixes

@@
 Next, we introduce the iRail query logs, which we study by the means of visualizations.
-Finally, we look at public transit data is being published, and how we would be able to get a better indication of travel flows than just mere query logs.
+Finally, we look at how public transit data is published and how we would be able to get a better indication of travel flows.
 
@@
 Over the past decade, large datasets have become increasingly commonplace due to the proliferation of sensor networks and portable devices like smartphones.
-Termed ``Big Data'' due to the large volume of data records that emerge from real-time sensing\cite{kitchin}, such datasets typically contain information of activities or processes linked to the space and time where they occur.
+Termed ``Big Data'' due to the large volume of data records that emerge from real-time sensing~\cite{kitchin}, such datasets typically contain information of activities or processes linked to the space and time where they occur.
 In the domain of ``Smart City'' research, much has been accomplished with the use of ``Big Data'' to monitor human movement.

### 2016-01-08 paper.tex
Related Work + Bibs Updated

--- 2016-01-08 paper.tex
@@
 For instance, interest in modelling traffic flows emerged from the strain placed on urban transportation systems during peak hours~\cite{roth,ferreira}.
-Likewise, insight into routine travel patterns is crucial for the conceptualisation of functional urban areas~\cite{servillo,sykora}, urban hierarchies~\cite{christaller} and other territorial structures.
+Likewise, insight into routine travel patterns is crucial for the conceptualisation of functional urban areas~\cite{servillo2013town}, urban hierarchies~\cite{christaller1966central} and other territorial structures.
 
@@
 This representation is detailed and straightforward to understand but becomes cluttered when multiple lines intersect.
+
 A number of solutions have been proposed for clutter reduction. 
 Filtering is the most commonly practiced. 
-The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \todo{Andrienko, 2007 -198}. 
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \cite{andrienko2007visual} . 
+Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualization.
 Another solution is to transform the visual representation. 
-Flow maps \todo{Phan, 2005 -88} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that do not obscure key geographic features such as roads, rivers or coastlines.
-However, this solution is optimal only if strict geographic accuracy is not required since the arrows do not depict actual paths of movement
+Flow maps \cite{phan2005flow} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that align with key geographic features like roads, rivers or coastlines. 
+However, a certain degree of ambiguity is introduced to the visualization since the arrows do not depict actual paths of movement and maybe misleading to interpret.
+
+Non-geographic representation is a viable alternative to this challenge in this specific context of visualizing flows on the rail network since the locations of stations and the exact routes between them are known and have been pre-determined.
+Thus the analytical focus and goal of visualization is to depict differences in the volume of flows between different stations. 
+While many visualization techniques fulfil this purpose, we find chord diagrams \cite{krzywinski2009circos} particularly appropriate for delivering information in a compact form, allowing for comparison across multiple plots in a single view.
+A further description of chord diagrams is provided in \cref{sec:vis}.
 
@@
 
-\subsection{Visualization}
+\subsection{Visualization}\label{sec:vis}
 %Flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 

### 2016-01-08 paper.tex

@@
 
-The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (See \cref{fig:weekdays_morning}) and weekends (See \cref{fig:weekends_morning}) respectively.
+The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively.
 Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.
-%We can observe that Brussels, Antwerp, Gent and Leuven are the dominant destinations.
-%Turning our attention to weekday travel, we can observe that Antwerp, Brussels and Gent are the dominant destinations.
+%We can observe that Brussels, Antwerp, Ghent and Leuven are the dominant destinations.
+%Turning our attention to weekday travel, we can observe that Antwerp, Brussels and Ghent are the dominant destinations.
 %The reverse occurs in the evening when trips are made towards provincial stations.
@@
 With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries.
-Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Gent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
+Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Ghent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
 These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels city as a major centre.
@@
   \centering
-  \begin{subfigure}{.95\textwidth}
+  \begin{subfigure}{1\textwidth}
     \centering
@@
   \end{subfigure}
-  \begin{subfigure}{.95\textwidth}
+  \begin{subfigure}{1\textwidth}
     \centering
@@
   \end{subfigure}
-  \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+  %\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
   \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
@@
 \begin{figure*}[t]
+\centering
 \begin{tabular}{ccc}
@@
   \includegraphics[width=.33\textwidth]{wd-gent.pdf}\\
-  (a) Brussels City&(b) Antwerp City&(c) Gent\\[6pt]
+  (a) Brussels City&(b) Antwerp City&(c) Ghent\\[6pt]
   \includegraphics[width=.33\textwidth]{wd-leuven.pdf}& 
@@
   \includegraphics[width=.33\textwidth]{wd-mons.pdf}\\ 
-  (d) Leuven&(e) Liege City&(f) Mons\\
+  (d) Leuven&(e) Liège City&(f) Mons\\
 \end{tabular}
+\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+\caption{Comparison of morning rush hour travel flows on weekdays originating from or destined for six major cities in Belgium.}
 \label{fig:cities}
-\centering
-\caption{Comparison of morning rush hour travel flows on weekdays originating from or destined for six major cities in Belgium.}
 \end{figure*}
@@
 Alternatively the large number of trips may have originated from users who repeatedly request for updates on delayed trains or from a concentration of people who utilise route planning applications on a regular basis.
-Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liege city.
+Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liège city.
 This prompted speculation, as well as discussions, on how such outliers should be treated or what may have led to their existence.
@@
 \label{sec:publishing}
-
-%label: fig:LDFAxis2
-\input{ldfaxis2}
 
@@
 Turning off caching may work in a private setting where you know how much and what kind of queries you can expect, yet on the open Web, caching helps decrease load on servers at peak moments.
+
+%label: fig:LDFAxis2
+\input{ldfaxis2}
 

### 2016-01-08 paper.tex

@@
 For instance, interest in modelling traffic flows emerged from the strain placed on urban transportation systems during peak hours~\cite{roth,ferreira}.
-Likewise, insight into routine travel patterns is crucial for the conceptualisation of functional urban areas~\cite{servillo2013town}, urban hierarchies~\cite{christaller1966central} and other territorial structures.
+Likewise, insight into routine travel patterns is crucial for the conceptualization of functional urban areas~\cite{servillo2013town}, urban hierarchies~\cite{christaller1966central} and other territorial structures.
 
@@
 The dataset described in this paper is compiled from the \emph{Apache} \emph{access.log} files generated between October 2012 and December 2015, filtered by ``connections'' entries.
-We made a copy of this dataset available in CSV format can be downloaded at \todo{\url{http://datawijs.be/apilog.tar.gz}}.
+We made a copy of this dataset available in CSV format can be downloaded at \url{https://github.com/pietercolpaert/irail-route-planning-query-logs}.
 %CSV header of the final logs: datetime,requested_datetime,from_id,from_name,from_long,from_lat,to_id,to_name,to_long,to_lat,user_agent,operating_system,is_bot
@@
 In this chapter, we report on the results of studying visualizations with experts of \emph{Trein Tram Bus}\footnote{\url{http://treintrambus.be}}, a not for profit passenger federation.
-%Isn't this sufficient ↑? \todo{Trein Tram Bus needs to be introduced in the introduction}
 
@@
 
-The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively.
+The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively\footnote{An interactive version of this visualization can be viewed at \url{http://rxd.architectuur.kuleuven.be/irail/weekdays.html} and \url{http://rxd.architectuur.kuleuven.be/irail/weekends.html}}.
 Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.

### 2016-01-08 paper.tex
Footnote to in line

@@
 
-The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively\footnote{An interactive version of this visualization can be viewed at \url{http://rxd.architectuur.kuleuven.be/irail/weekdays.html} and \url{http://rxd.architectuur.kuleuven.be/irail/weekends.html}}.
+The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively.
+An interactive version of this visualization can be viewed at \url{http://rxd.architectuur.kuleuven.be/irail/weekdays.html} and \url{http://rxd.architectuur.kuleuven.be/irail/weekends.html}. 
 Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.

### 2016-01-08 paper.tex
Modified related work

@@
 
-\emph{Flow analysis} is a topic of theoretical interest and practical importance in various disciplines. 
-Flow analysis is conventionally conducted to study spatial dynamics and identify routine patterns in the movement of people.
+\emph{Flow analysis} is a topic of theoretical interest and practical importance conventionally conducted to study spatial dynamics and identify routine patterns in the movement of people.
 For instance, interest in modelling traffic flows emerged from the strain placed on urban transportation systems during peak hours~\cite{roth,ferreira}.
@@
 Filtering is the most commonly practiced. 
-The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \cite{andrienko2007visual} . 
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \cite{andrienko2007visual}. 
 Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualization.
+
 Another solution is to transform the visual representation. 
@@
 However, a certain degree of ambiguity is introduced to the visualization since the arrows do not depict actual paths of movement and maybe misleading to interpret.
+A computationally cost effective alternative is to draw links as arcs instead of arrows \cite{chua2014flowsampler}.
+Yet, this visual encoding may intensify the visual complexity of the representation.
 

### 2016-01-08 paper.tex
Title changed and subsections introduced in relwork
Chord diagram, added reference in section 4.2

@@
 
-\title{Studying public transit API query logs\\ to get an indication of travel flows}
+\title{What public transit API logs tell us about travel flows}
 \numberofauthors{6}
@@
 \label{sec:relwork}
+%\subsection{Flow analysis}
 
@@
 
+\subsection{Studying Query Logs}
+
 On highly expressive servers, such as \sparql-endpoints, query logs can play an important role in fine-tuning the underlying datastores~\cite{arias2011empirical}.
@@
 This was called ``a blessing for privacy'', yet adding the user query as an HTTP header was suggested.
+
+\subsection{Visualizing Travel Flows}
 
@@
 %Flow visualizations are commonly employed to facilitate the comparison of aggregated movement. 
-The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle. 
+The chord diagram is a visualization technique based on the node-link metaphor that arranges nodes along the circumference of a circle \cite{krzywinski2009circos}. 
 Each node is represented by an arc where its length is proportional to the total volume of incoming and outgoing flows. 

### 2016-01-14 paper.tex
Citation added

@@
 %Visualizations are crucial for understanding the patterns in data and representing transit flows.
-Visualizations are crucial for making data analysis tangible, so that results can be communicated and debated \cite{robinson2008collaborative}. 
+Visualizations are crucial for making data analysis tangible \cite{thomas2006visual}, so that results can be communicated and debated \cite{robinson2008collaborative}. 
 This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
@@
 OD matrices can be reordered to emphasise connectivity between locations.
-While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor.
+While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor \cite{holten2011extended}.
 In recent years, such representations have become increasingly prominent for displaying movement in science as well as popular culture.

### 2016-01-14 paper.tex
Citations added

@@
 Another solution is to transform the visual representation. 
-Flow maps \cite{phan2005flow} reduces clutter by rerouting the arrows to minimize intersection and obtain layouts that align with key geographic features like roads, rivers or coastlines. 
-However, a certain degree of ambiguity is introduced to the visualization since the arrows do not depict actual paths of movement and maybe misleading to interpret.
+This can be achieved by bundling lines in close proximity to minimize intersections \cite{hurter2014bundled}.
+Flow maps \cite{phan2005flow} follow the same principle to reduce clutter but yield layouts that align with key geographic features like roads, rivers or coastlines. 
+However, a certain degree of ambiguity is introduced to the visualization by both techniques since the arrows do not depict actual paths of movement and maybe misleading to interpret.
+Transition diagrams are a type of visulization that summarizes large numbers of trajectories while minimizing arbitary transformations to the layout \cite{andrienko2012visual}. 
+The idea is to divide each trajectory into segments, so that movement on a path is approximated by a sequence of transitions between adjacent discrete locations. 
+Locations are either predefined or obtained by various space partitioning algorithms. 
+Arrows depict flows and are scaled to the aggregate number of transitions between pairs of adjacent locations. 
+An advantage of this representation is that the topology of flows is captured as a directed graph, allowing for quantitative metrics like centrality to be computed and visually compared. 
+Nonetheless, the solution requires substantial data processing prior to visualization and analysis.
 A computationally cost effective alternative is to draw links as arcs instead of arrows \cite{chua2014flowsampler}.

### 2016-01-15 paper.tex

@@
 Flow analysis involves the comparison of aggregated movement between distinct locations within a specific time frame. 
-In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for the sensitivities to be explored. 
+In most cases, the analytical objective is to identify trends that occur in journeys made between pairs of locations and allow for their differences to be explored. 
 Data visualization is typically employed to facilitate in this process thus the input data must be transformed into a format that allows for the quantity and directional flow of movement to be visually encoded.
@@
 To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities were aggregated into seperate groups. 
-Simiarly, stations located outside of Belgium were aggregated to distinct group. 
+Similarly, stations located outside of Belgium were aggregated to distinct group. 
 A diagrammatic breakdown is depicted in \cref{fig:method-c} where color and shape is used to demarcate each group.

### 2016-01-15 paper.tex
Merge with Alvin, better formatting of authors and reviewed Andrew's comments in introduction

@@
 \alignauthor
-Pieter Colpaert\\
-\affaddr{\email{\texttt{pieter.colpaert@ugent.be}}}
+Pieter Colpaert
+\textsuperscript{*}
 \and
 \alignauthor
-Alvin Chua\\
-\affaddr{\email{\texttt{alvin.chua@asro.kuleuven.be}}}
+Alvin Chua\textsuperscript{\#}
 \and
 \alignauthor
-Ruben Verborgh\\
-\affaddr{\email{\texttt{ruben.verborgh@ugent.be}}}
+Ruben Verborgh\textsuperscript{*}
 \and
 \alignauthor
-Erik Mannens\\
-\affaddr{\email{\texttt{erik.mannens@ugent.be}}}
+Erik Mannens\textsuperscript{*}
 \and
 \alignauthor
-Rik Van de Walle\\
-\affaddr{\email{\texttt{rik.vandewalle@ugent.be}}}
+Rik Van de Walle\textsuperscript{*}
 \and
 \alignauthor
-Andrew Vande Moere\\
-\affaddr{\email{\texttt{andrew.vandemoere\\@asro.kuleuven.be}}}\\
+Andrew Vande Moere\textsuperscript{\#}
+\sharedaffiliation
+\affaddr{\textsuperscript{*} iMinds -- Ghent University -- MMLab\\
+\email{\texttt{firstname.lastname@ugent.be}}}\\
+\affaddr{\textsuperscript{\#} Research[x]Design -- KU Leuven\\
+\email{\texttt{firstname.lastname@kuleuven.be}}}
 }
@@
 \begin{abstract}
-
 %Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
@@
 
-The \emph{iRail} project\footnote{\url{https://hello.irail.be}} started in 2008 to make the data of the Belgian railway accessible for developers.
+The \emph{iRail} project\footnote{\url{https://hello.irail.be}} started in 2008 to make route planning data of the Belgian railway accessible for developers.
 The project offers developers both a data dump for third party apps (since $\sim$~August~2015) and a route planning API (since $\sim$~October~2010).
@@
 
-Getting indications of people flows through public transit networks is a challenge. 
-Yet, \emph{can we not get an indication of these transit flows by studying the query logs of Web-services?}
+Getting an indication of how people move within a transit network is of high importance within the field of \emph{smart cities}, \emph{urban planning} as well as \emph{demand driven mobility}.
+Furthermore, it may benefit travelers to know in real-time whether the upcoming trip is in popular demand, as it may trigger the traveler to take the next, less congested, trip.
+In this paper, we study the query logs of route planning web-services as an indication of people flows within transit networks.
 
-In this paper, we give a small overview of related work to gather data for flow analysis.
+We give a small overview of related work to gather data for flow analysis.
 Next, we introduce the iRail query logs, which we study by the means of visualizations.
@@
 %Visualizations are crucial for understanding the patterns in data and representing transit flows.
-Visualizations are crucial for making data analysis tangible \cite{thomas2006visual}, so that results can be communicated and debated \cite{robinson2008collaborative}. 
-This is key for our research, since the questions presented are exploratory in nature \cite{kraak2008exploratory} and can be addressed in many ways. 
-Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support flow analysis \cite{andrienko2012visual}. 
+Visualizations are crucial for making data analysis tangible~\cite{thomas2006visual}, so that results can be communicated and debated~\cite{robinson2008collaborative}. 
+This is key for our research, since the questions presented are exploratory in nature~\cite{kraak2008exploratory} and can be addressed in many ways. 
+Movement data has received substantial attention from the visualization community and a range of techniques has been developed to support flow analysis~\cite{andrienko2012visual}. 
 Visualizations of flows can be broadly categorized into two groups. 
@@
 OD matrices can be reordered to emphasise connectivity between locations.
-While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor \cite{holten2011extended}.
+While this feature is particularly useful for identifying trends and outliers in the data, the tabular display is challenging for lay users to interpret and less expressive of connections between locations than representations based on the node-link metaphor~\cite{holten2011extended}.
 In recent years, such representations have become increasingly prominent for displaying movement in science as well as popular culture.
@@
 Filtering is the most commonly practiced. 
-The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations \cite{andrienko2007visual}. 
+The same goal can be achieved by clustering lines that have similar properties like common start, end or intermediate locations~\cite{andrienko2007visual}. 
 Both solutions are effective at reducing clutter yet it should be noted that excessive aggregation or filtering removes a significant amount of information from the visualization.
@@
 Another solution is to transform the visual representation. 
-This can be achieved by bundling lines in close proximity to minimize intersections \cite{hurter2014bundled}.
-Flow maps \cite{phan2005flow} follow the same principle to reduce clutter but yield layouts that align with key geographic features like roads, rivers or coastlines. 
+This can be achieved by bundling lines in close proximity to minimize intersections~\cite{hurter2014bundled}.
+Flow maps~\cite{phan2005flow} follow the same principle to reduce clutter but yield layouts that align with key geographic features like roads, rivers or coastlines. 
 However, a certain degree of ambiguity is introduced to the visualization by both techniques since the arrows do not depict actual paths of movement and maybe misleading to interpret.
-Transition diagrams are a type of visulization that summarizes large numbers of trajectories while minimizing arbitary transformations to the layout \cite{andrienko2012visual}. 
+Transition diagrams are a type of visulization that summarizes large numbers of trajectories while minimizing arbitary transformations to the layout~\cite{andrienko2012visual}. 
 The idea is to divide each trajectory into segments, so that movement on a path is approximated by a sequence of transitions between adjacent discrete locations. 

### 2016-01-15 paper.tex
Not our data and small abstract change

@@
 % Context & % Need
-In the field of urban planning, researchers need an indication of how people move between cities. 
+In the field of smart cities, researchers need an indication of how people move in and between cities.
 Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
 % Task
-In order to get an indication of travel flows between cities in Belgium,
+In order to get an indication of public transit travel flows in Belgium,
 we analyzed the query logs of the iRail API, a highly expressive route planning API for the Belgian railways.
@@
 Representations that display the complete trajectory and Origin Destination (OD) type representations that take only the start and end locations into account.
-Since our data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
+Since iRail's query log data solely consists of OD information, we narrow our description to the latter group of visualization techniques.
 
@@
 Furthermore, this way of publishing data enables research for real-time predictions of travel intentions.
-Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and, therefore, IP addresses have been removed to protect the privacy of our users. 
+Privacy preservation is paramount~\cite{silvestri} when dealing with query logs and, therefore, IP addresses have been removed to protect the privacy of iRail's users. 
 Additionally, the integrity of this dataset, without IP addresses, has been vetted by the Belgian Privacy Commission and permission has been granted for these query logs to be published as open data.
@@
 Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
-Assuming that each request reciprocates with an intention to travel by rail, our data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
-
+Assuming that each request reciprocates with an intention to travel by rail, iRail's query log data captures a respective $0.56\%$, 0.88\%, 0.77\% and 1.66\% of the actual journeys that have occurred.
 %After analysis, it appeared 10\% of the request where performed by Google bots.

### 2016-01-15 paper.tex
Feedback of Andrew processed

@@
 \cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
-A distinctive increase in the number of queries can be observed in April 2015 when the official route planning service provided by of the Belgian rail company was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
+A distinctive increase in the number of queries can be observed in April 2015 when one of the official route planning apps provided by the Belgian rail company (RailTime) was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
 Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.

### 2016-01-15 paper.tex
Feedback PH processed

@@
 Furthermore, the graph patterns are usually star-shaped and despite triple pattern chains exist, they are generally short.
-Other, more cacheable interfaces more to the left on the Linked~Data~Fragments axis~\cite{ldf}, also benefit from query logs in order to e.g., know who is using what clients to query the Web, yet the user intention is lost~\cite{usewod2015}.
+Other, more cacheable interfaces, as introduced by Linked~Data~Fragments~\cite{ldf}, also benefit from query logs in order to e.g., know who is using what clients to query the Web, yet the user intention is lost~\cite{usewod2015}.
 This was called ``a blessing for privacy'', yet adding the user query as an HTTP header was suggested.

### 2016-02-18 paper.tex
Trying to get everything to work

@@
 
-\usepackage{listings}
-
 % Typography
@@
 \usepackage[normalem]{ulem}
-\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=2,
+\PassOptionsToPackage{hyphens}{url}\usepackage[pdftex,bookmarks,bookmarksopen,bookmarksdepth=2,
             urlcolor=black,colorlinks=true,linkcolor=black,citecolor=black]{hyperref}
@@
 \def\UrlFont{\em}
-
 % Graphics
@@
 \usetikzlibrary{arrows,positioning,shapes.misc}
-%\let\subcaption\relax
+\let\subcaption\relax
 \usepackage{graphicx}
@@
 %\usepackage{caption}
-\usepackage{subcaption}
-\captionsetup[figure]{labelfont=bf,textfont=bf}
-\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont}
-\definecolor{lightgrey}{RGB}{170, 170, 170}
-\definecolor{darkblue}{RGB}{0, 0, 0}
-\definecolor{darkred}{RGB}{170, 0, 0}
-\definecolor{darkgreen}{RGB}{0, 110, 0}
+%\usepackage{subcaption}
+%\captionsetup[figure]{labelfont=bf,textfont=bf}
+%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont}
+%\definecolor{lightgrey}{RGB}{170, 170, 170}
+%\definecolor{darkblue}{RGB}{0, 0, 0}
+%\definecolor{darkred}{RGB}{170, 0, 0}
+%\definecolor{darkgreen}{RGB}{0, 110, 0}
 
@@
 \definecolor{grey}{RGB}{130,130,130}
+\usepackage{color}
+\newdef{definition}{Definition}
 
-\usepackage{color}
-\newcommand{\todo}[1]{\noindent\textcolor{red}{{\bf \{TODO}: #1{\bf \}}}}
-\newcommand{\rv}[1]{\noindent\textcolor{red}{{\bf \{RV}: #1{\bf \}}}}
-\newdef{definition}{Definition}
+\permission{Copyright is held by the International World Wide Web Conference Committee (IW3C2).
+IW3C2 reserves the right to provide a hyperlink to the author's site if the Material is used in electronic media.}
+\conferenceinfo{WWW'16 Companion,}{April 11--15, 2016, Montr\'eal, Qu\'ebec, Canada.} 
+\copyrightetc{ACM \the\acmcopyr}
+\crdata{978-1-4503-4144-8/16/04. \\
+\url{http://dx.doi.org/10.1145/2872518.2891069}}
+
+\clubpenalty=10000 
+\widowpenalty=10000
 
@@
 \numberofauthors{6}
-\author{
-\alignauthor
-Pieter Colpaert
-\textsuperscript{*}
+\author{\alignauthor
+Pieter Colpaert\textsuperscript{*}
 \and
@@
 }
-
 \maketitle
 \begin{abstract}
-%Public transit schedules are made available on the Web in various ways, such as in a GTFS file, a route planning API, or through Linked Connections.
-%This webservice provides an answer to a route planning question with parameters such as ``departuretime'', ``from'' and ``to''.
-% Context & % Need
 In the field of smart cities, researchers need an indication of how people move in and between cities.
 Yet, getting statistics of travel flows within public transit systems has proven to be troublesome.
-% Task
 In order to get an indication of public transit travel flows in Belgium,
@@
 We were able to study $\sim$100k to 500k requests for each month between October 2012 and November 2015, which is between $0.56\%$ and $1.66\%$ of the amount of monthly passengers.
-% Object & Findings
 Using data visualizations, we illustrate the commuting patterns in Belgium and confirm that Brussels, the capital, acts as a central hub. 
@@
 The findings correspond to the real travel demand, according to experts of the passenger federation Trein Tram Bus.
-% Conclusion
 We conclude that query logs of route planners are of high importance in getting an indication of travel flows.
-% Perspectives
 However, better travel intentions would be acquirable using dedicated HTTP POST requests.
-%%Redundant according to RV:
-%Highly expressive transport data publishing methods such as route planning API exist, as well as data dumps or data fragments.
-%In order to be able to gather meaningful logs in all cases, we suggest using a separate POST request containing the entire query.
-
 \end{abstract}
@@
 Today, the Belgian railway company (SNCB) takes manual samples of people getting on and off trains in stations. 
-To date, this is still the SNCB's only source to determine people flows through their network\footnote{\url{http://www.belgianrail.be/nl/corporate/in-de-kijker/comptages_voyageurs_2014.aspx}}.
+To date, this is still the SNCB's only source to determine people flows through their network.
 Also within other transit systems the data is tedious to get, certainly when passengers do not have to check-in and check-out.
@@
 The project offers developers both a data dump for third party apps (since $\sim$~August~2015) and a route planning API (since $\sim$~October~2010).
-The query logs of the latter has been stored since November 2012 (requests before that period got lost due to server migrations).
+The query logs of the latter have been stored since November 2012 (requests before that period got lost due to server migrations).
 
@@
 
-Over the past decade, large datasets have become increasingly commonplace due to the proliferation of sensor networks and portable devices like smartphones.
+Over the past decade, large datasets have become increasingly commonplace thanks to the proliferation of sensor networks and portable devices like smartphones.
 Termed ``Big Data'' due to the large volume of data records that emerge from real-time sensing~\cite{kitchin}, such datasets typically contain information of activities or processes linked to the space and time where they occur.
@@
 \caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
-In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden raise of iRail API queries.}
+In April 2015 an official app of the Belgian railways was discontinued, which explains the sudden rise of iRail API queries.}
 \label{fig:querylogs}
@@
 There has been an increase in the number of queries made between 2012 and 2015. 
-\cref{fig:querylogs} provides a break down of the numbers on a monthly basis. 
+\cref{fig:querylogs} provides a breakdown of the numbers on a monthly basis. 
 A distinctive increase in the number of queries can be observed in April 2015 when one of the official route planning apps provided by the Belgian rail company (RailTime) was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
@@
   \includegraphics[width=14cm]{method-c.pdf}\\
-  \caption{Each station is color coded to indicate the province it will be aggregated to. Stations in major cities as well as stations outside of Belgium are aggregated to seperate groups.}
+  \caption{Each station is color coded to indicate the province it will be aggregated into. Stations in major cities as well as stations outside of Belgium are aggregated to seperate groups.}
   \label{fig:method-c}
@@
 
-\begin{figure}[h]
-  \centering
-  \begin{subfigure}{.2375\textwidth}
-    \includegraphics[width=4cm]{method-a.pdf}\\
-    \subcaption{} % 4 words max
-    \label{fig:method-a}
-  \end{subfigure}%
-  \begin{subfigure}{.2375\textwidth}
-    \includegraphics[width=4cm]{method-b.pdf}\\
-    \subcaption{} % 4words max
-    \label{fig:method-b}
-  \end{subfigure}%
-  \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-  \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
+%\begin{figure}[h]
+%  \centering
+%  \begin{subfigure}{.2375\textwidth}
+%    \includegraphics[width=4cm]{method-a.pdf}\\
+%    \subcaption{} % 4 words max
+%    \label{fig:method-a}
+%  \end{subfigure}%
+%  \begin{subfigure}{.2375\textwidth}
+%    \includegraphics[width=4cm]{method-b.pdf}\\
+%    \subcaption{} % 4words max
+%    \label{fig:method-b}
+%  \end{subfigure}%
+%  \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+%  \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
 (a) Individual data records are depicted geographically with arrows connecting origins to their respective destinations.
 (b) The result of aggregation is a set of weighted connections between provinces.}
-  \label{fig:method-ab}
-\end{figure}
-
+%  \label{fig:method-ab}
+%\end{figure}
+%
 Next, the data is aggregated so that province level flow patterns are emphasized.
@@
 To compare the volume of flow between rural provincial areas and urbanized zones, train stations in major cities were aggregated into seperate groups. 
-Similarly, stations located outside of Belgium were aggregated to distinct group. 
+Similarly, stations located outside of Belgium were aggregated into distinct group. 
 A diagrammatic breakdown is depicted in \cref{fig:method-c} where color and shape is used to demarcate each group.
@@
 In our implementation of the chord diagram, nodes are colour coded to indicate individual provinces and the major cities located within their administrative boundaries. 
-Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when upon mouse-over.
+Interactive filtering is introduced to simplify the visualization and details are provided in pop-up dialogue when on mouse-over.
 
@@
 With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries.
-Queries from Flanders, on the hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Ghent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
+Queries from Flanders, on the other hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Ghent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
 These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels city as a major centre.
 The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
-
-\begin{figure*}
-  \centering
-  \begin{subfigure}{1\textwidth}
-    \centering
-    \includegraphics[width=13cm]{weekdays_morning.pdf}\\
-    \subcaption{Weekdays} % 4 words max
-    \label{fig:weekdays_morning}
-  \end{subfigure}
-  \begin{subfigure}{1\textwidth}
-    \centering
-    \includegraphics[width=13cm]{weekends_morning.pdf}\\
-    \subcaption{Weekends} % 4words max
-    \label{fig:weekends_morning}
-  \end{subfigure}
-  %\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-  \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
-  \label{fig:mornings}
-\end{figure*}
-
+%
+%\begin{figure*}
+%  \centering
+%  \begin{subfigure}{1\textwidth}
+%    \centering
+%    \includegraphics[width=13cm]{weekdays_morning.pdf}\\
+%    \subcaption{Weekdays} % 4 words max
+%    \label{fig:weekdays_morning}
+%  \end{subfigure}
+%  \begin{subfigure}{1\textwidth}
+%    \centering
+%    \includegraphics[width=13cm]{weekends_morning.pdf}\\
+%    \subcaption{Weekends} % 4words max
+%    \label{fig:weekends_morning}
+%  \end{subfigure}
+%  %\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+%  \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
+%  \label{fig:mornings}
+%\end{figure*}
+%
 \begin{figure*}[t]
@@
 
-\subsection{Counter Intuitive Results}
+\subsection{Counter-Intuitive Results}
 
@@
 Alternatively the large number of trips may have originated from users who repeatedly request for updates on delayed trains or from a concentration of people who utilise route planning applications on a regular basis.
-Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liège city.
+Accordingly, the reverse is identified in Liège where an unusually large connection originating from a suburb towards Liège city exists.
 This prompted speculation, as well as discussions, on how such outliers should be treated or what may have led to their existence.
@@
 %label: fig:LDFAxis2
-\input{ldfaxis2}
+%\input{ldfaxis2}
 

### 2016-03-14 paper.tex
Fixed ACM feedback

@@
 \documentclass{sig-alternate}
+\setlength{\paperheight}{11in}
+\setlength{\paperwidth}{8.5in}
+\usepackage[
+  pass,% keep layout unchanged 
+  % showframe,% show the layout
+]{geometry}
+
 \usepackage[utf8]{inputenc}
@@
 \usepackage{graphicx}
-%\usepackage[caption=false]{subfig}
-%\usepackage{caption}
-%\usepackage{subcaption}
-%\captionsetup[figure]{labelfont=bf,textfont=bf}
-%\captionsetup[subfigure]{labelfont=normalfont,textfont=normalfont}
-%\definecolor{lightgrey}{RGB}{170, 170, 170}
-%\definecolor{darkblue}{RGB}{0, 0, 0}
-%\definecolor{darkred}{RGB}{170, 0, 0}
-%\definecolor{darkgreen}{RGB}{0, 110, 0}
+\usepackage[caption=false]{subfig}
+\usepackage{caption}
+\usepackage{subcaption}
+\captionsetup[figure]{labelfont=bf,textfont=bf}
+\definecolor{lightgrey}{RGB}{170, 170, 170}
+\definecolor{darkblue}{RGB}{0, 0, 0}
+\definecolor{darkred}{RGB}{170, 0, 0}
+\definecolor{darkgreen}{RGB}{0, 110, 0}
 
@@
 \clubpenalty=10000 
-\widowpenalty=10000
-
+\widowpenalty = 10000
 \begin{document}
 
-\title{What public transit API logs tell us about travel flows}
+\title{What Public Transit API Logs Tell Us about Travel Flows}
 \numberofauthors{6}
 \author{\alignauthor
-Pieter Colpaert\textsuperscript{*}
+Pieter Colpaert\wu
 \and
 \alignauthor
-Alvin Chua\textsuperscript{\#}
+Alvin Chua\wg
 \and
 \alignauthor
-Ruben Verborgh\textsuperscript{*}
+Ruben Verborgh\wu
 \and
 \alignauthor
-Erik Mannens\textsuperscript{*}
+Erik Mannens\wu
 \and
 \alignauthor
-Rik Van de Walle\textsuperscript{*}
+Rik Van de Walle\wu
 \and
 \alignauthor
-Andrew Vande Moere\textsuperscript{\#}
+Andrew Vande Moere\wg
 \sharedaffiliation
-\affaddr{\textsuperscript{*} iMinds -- Ghent University -- MMLab\\
-\email{\texttt{firstname.lastname@ugent.be}}}\\
-\affaddr{\textsuperscript{\#} Research[x]Design -- KU Leuven\\
-\email{\texttt{firstname.lastname@kuleuven.be}}}
+\begin{tabular}{ccc}
+\affaddr{{\wu}iMinds -- Ghent University -- Data Science Lab} & & \affaddr{{\wg}Research[x]Design -- KU Leuven{\ }}\\
+\affaddr{\email{\texttt{firstname.lastname@ugent.be}}} & & \affaddr{\email{\texttt{firstname.lastname@kuleuven.be}}}\\
+\end{tabular}
 }
@@
 \end{abstract}
-
-\vspace{1em}
 
@@
 \begin{enumerate}
-\item \emph{Route Planning} (\url{http://api.irail.be/connections/{?from,to,date,time}}) plans a route from one station to another, taking into account a preferred start time.
-\item \emph{Next Departures} (\url{http://api.irail.be/liveboard/{?station,date,time}}) shows the next train departures in a certain station, useful for creating quick overviews.
-\item \emph{Trip Status} (\url{http://api.irail.be/vehicle/{?id,date}}) returns the stations this train passes by on the day of the request.
-\item \emph{List of all stations} (\url{http://api.irail.be/stations/}) gives an overview of all stations Belgian trains can arrive at.
+\item \emph{Route Planning}\footnote{\url{http://api.irail.be/connections/{?from,to,date,time}}} plans a route from one station to another, taking into account a preferred start time.
+\item \emph{Next Departures}\footnote{\url{http://api.irail.be/liveboard/{?station,date,time}}} shows the next train departures in a certain station, useful for creating quick overviews.
+\item \emph{Trip Status}\footnote{\url{http://api.irail.be/vehicle/{?id,date}}} returns the stations this train passes by on the day of the request.
+\item \emph{List of all stations}\footnote{\url{http://api.irail.be/stations/}} gives an overview of all stations Belgian trains can arrive at.
 \end{enumerate}
@@
 \centering
-\includegraphics[width=8.6cm]{querylogs}
+\includegraphics[width=8.5cm]{querylogs}
 \caption{Amount of queries on the route planning part of the iRail API between October 2012 and December 2015.
@@
 A distinctive increase in the number of queries can be observed in April 2015 when one of the official route planning apps provided by the Belgian rail company (RailTime) was discontinued\footnote{\url{https://hello.irail.be/2015/04/22/april-updates/}}, resulting in widespread adoption of alternative software applications built on top of the iRail API.
-Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://www4.vlaanderen.be/sites/svr/Cijfers/Exceltabellen/mobiliteit/vervoersprestaties/personenvervoer/MOBIOPEN006.xls}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
+Calculated with indicators provided by the Flemish regional government of Belgium\footnote{\url{http://ur1.ca/ojeyz}}, we infer a monthly average of 19.3, 19.4, 19,6 and 19.7 million passengers in 2012, 2013, 2014 and 2015 respectively.
 Correspondingly, the average number of iRail queries per month amounts to 0.11, 0.17, 0.15 and 0.33 million respectively.
@@
 
-%\begin{figure}[h]
-%  \centering
-%  \begin{subfigure}{.2375\textwidth}
-%    \includegraphics[width=4cm]{method-a.pdf}\\
-%    \subcaption{} % 4 words max
-%    \label{fig:method-a}
-%  \end{subfigure}%
-%  \begin{subfigure}{.2375\textwidth}
-%    \includegraphics[width=4cm]{method-b.pdf}\\
-%    \subcaption{} % 4words max
-%    \label{fig:method-b}
-%  \end{subfigure}%
-%  \vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-%  \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
+\begin{figure}[h]
+  \centering
+  \begin{tabular}{cc}
+    \includegraphics[width=4cm]{method-a.pdf}\label{fig:method-a} & \includegraphics[width=4cm]{method-b.pdf}\label{fig:method-b}\\
+  \end{tabular}
+  \caption{Spatial aggregation based on provincial administrative boundaries diagrammatically represented.
 (a) Individual data records are depicted geographically with arrows connecting origins to their respective destinations.
 (b) The result of aggregation is a set of weighted connections between provinces.}
-%  \label{fig:method-ab}
-%\end{figure}
-%
+  \label{fig:method-ab}
+\end{figure}
+
 Next, the data is aggregated so that province level flow patterns are emphasized.
@@
 \centering
-\includegraphics[width=8.6cm]{avg-all}
+\includegraphics[width=8.5cm]{avg-all}
 \caption{Distribution of iRail route planning API queries on average per day of the week between October 2012 and December 2015}
@@
 \centering
-\includegraphics[width=8.6cm]{dec24}
+\includegraphics[width=8.5cm]{dec24}
 \caption{Comparison of route planning queries on December 24th (Christmass Eve) 2015 with the average on a Thursday between October 2012 and December 2015.
@@
 
-The chord diagrams shown in \cref{fig:mornings} depict the aggregated number of queries made between any two stations on weekdays (see \cref{fig:weekdays_morning}) and weekends (see \cref{fig:weekends_morning}) respectively.
-An interactive version of this visualization can be viewed at \url{http://rxd.architectuur.kuleuven.be/irail/weekdays.html} and \url{http://rxd.architectuur.kuleuven.be/irail/weekends.html}. 
+An interactive chord diagram can be viewed at \url{http://rxd.architectuur.kuleuven.be/irail/weekdays.html} and \url{http://rxd.architectuur.kuleuven.be/irail/weekends.html}. 
+The chord diagrams depict the aggregated number of queries made between any two stations on weekdays, as well as on weekends. 
 Visualizing our dataset in this manner reveals the complexity of flows on the Belgian rail system and the significance of cities in daily travel.
@@
 %The reverse occurs in the evening when trips are made towards provincial stations.
-Brussels City stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity (see \cref{fig:mornings}a).
+Brussels City stands out as the most distinctive visual element in both figures, indicating its function as the principal centre of rail activity.
 On a regional level, its function as a centre appears to be more distinctive for the Walloon region than Flanders.
-With exception to Liège city (see \cref{fig:mornings}e), queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries.
-Queries from Flanders, on the other hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp (see \cref{fig:mornings}b), Ghent (see \cref{fig:mornings}c) and Leuven (see \cref{fig:mornings}d).
+With exception to Liège city, queries from the Walloon region are generally made from provincial stations towards Brussels instead of major cities within their administrative boundaries. 
+Queries from Flanders, on the other hand, tend to be distributed among four major cities otherwise known as the Flemish Diamond, a network of metropolitan areas in Belgium comprised of Brussels, Antwerp, Ghent and Leuven.
 These insights indicate that the structure of flows in Flanders appear to follow a polycentric pattern while that of the Walloon region is relatively monocentric, with Brussels city as a major centre.
 The difference between both regions appear to correspond with existing measures of population density and degree of urbanisation, providing valuable insight as well as alternative perspectives into the function of cities in rural and urban settings.
-%
+
 %\begin{figure*}
@@
 %    \includegraphics[width=13cm]{weekdays_morning.pdf}\\
-%    \subcaption{Weekdays} % 4 words max
+%    \subcaption{Weekdays}  4 words max
 %    \label{fig:weekdays_morning}
@@
 %    \includegraphics[width=13cm]{weekends_morning.pdf}\\
-%    \subcaption{Weekends} % 4words max
+%    \subcaption{Weekends}  4words max
 %    \label{fig:weekends_morning}
 %  \end{subfigure}
-%  %\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+%  \vspace{5.0pt} To create a bit of vertical space between the 2 figure captions
 %  \caption{Comparison of morning rush hour travel flows on weekdays and weekends.}
@@
 %
-\begin{figure*}[t]
-\centering
-\begin{tabular}{ccc}
-  \includegraphics[width=.33\textwidth]{wd-brussels.pdf}& 
-  \includegraphics[width=.33\textwidth]{wd-antwerp.pdf}& 
-  \includegraphics[width=.33\textwidth]{wd-gent.pdf}\\
-  (a) Brussels City&(b) Antwerp City&(c) Ghent\\[6pt]
-  \includegraphics[width=.33\textwidth]{wd-leuven.pdf}& 
-  \includegraphics[width=.33\textwidth]{wd-liege.pdf}& 
-  \includegraphics[width=.33\textwidth]{wd-mons.pdf}\\ 
-  (d) Leuven&(e) Liège City&(f) Mons\\
-\end{tabular}
-\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
-\caption{Comparison of morning rush hour travel flows on weekdays originating from or destined for six major cities in Belgium.}
-\label{fig:cities}
-\end{figure*}
-
+%\begin{figure*}[t]
+%\centering
+%\begin{tabular}{ccc}
+%  \includegraphics[width=.3\textwidth]{wd-brussels.pdf}& 
+%  \includegraphics[width=.3\textwidth]{wd-antwerp.pdf}& 
+%  \includegraphics[width=.3\textwidth]{wd-gent.pdf}\\
+%  (a) Brussels City&(b) Antwerp City&(c) Ghent\\[6pt]
+%  \includegraphics[width=.3\textwidth]{wd-leuven.pdf}& 
+%  \includegraphics[width=.3\textwidth]{wd-liege.pdf}& 
+%  \includegraphics[width=.3\textwidth]{wd-mons.pdf}\\ 
+%  (d) Leuven&(e) Liège City&(f) Mons\\
+%\end{tabular}
+%\vspace{5.0pt}% To create a bit of vertical space between the 2 figure captions
+%\caption{Comparison of morning rush hour travel flows on weekdays originating from or destined for six major cities in Belgium.}
+%\label{fig:cities}
+%\end{figure*}
+%
 \subsection{Counter-Intuitive Results}
@@
 %label: fig:LDFAxis2
-%\input{ldfaxis2}
+\input{ldfaxis2}
 
+\vfill\eject
 \section{Conclusion and future work}
